### 2019-10-10 R/temporal-sampling.Rmd
Initial commit

---
title: "Effect of temporal sampling"
author: "Rosalie Bruel"
date: "03/10/2019"
output: 
  html_document:
    df_print: paged
    fig_caption: true
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
  df_print: paged
fontsize: 11pt
---

Before running the script, define below in the function _getpath4data_ the local folder where you are storing the dataset(s). <br>
The Lake Zurich countings were carried out as part of a study during my PhD, comparing several lakes trajectory. I still haven't published that study.
I have an agreement with the different persons who gave me access to sequences that they will be co-author on the publication. In the case of Lake Zurich sequence, this person is Nathalie Dubois (ETH Zurich, CH). So until this is figured out, I'd rather not publish any data on my Github.

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE, warning = FALSE, message = FALSE)

getpath4data <- function() {
  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
  if(Sys.getenv("USER")=="put here your USER") return("replace here with path to data")
  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
}

fig_cap <- captioner::captioner()
tab_cap <- captioner::captioner("Table")
```

# Lake Zurich

```{r read zurich data, include=T}
zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt"))
```

## Study site
Lake Zürich (CH) has a surface area of 65 km<sup>2</sup> and a maximum depth of 136 meters. The lake is divided into two basins. The lake is considered to be monomictic or dimictic, but the increase of strength of thermal stratification in the last few decades as a consequence of climate warming impedes complete mixing of the water column (Anneville et al., 2004^[Anneville, O., Souissi, S., Gammeter, S., Straile, D., 2004. Seasonal and inter-annual scales of variability in phytoplankton assemblages: comparison of phytoplankton dynamics in three peri-alpine lakes over a period of 28 years. Freshwater Biology 49, 98–115. https://doi.org/10.1046/j.1365-2426.2003.01167.x]).

## Coring and dating
A short core (ZH-17-21, 103 cm) was taken from the pelagic zone in 2017. Only the upper 50 cm were sampled and analysed within this study. The sediment of Lake Zurich presents a laminated facies on the surface (0-28.3 cm) that allowed varves counting. A linear interpolation allowed getting ages for the rest of the core. There was no change of sedimentation for the top 22.5-cm (1930-2017) which support the realism of such approximation. From 22.5 cm downward, the model fitted with only 3 points indicates a higher sedimentation rate.
The sediment sequence covers the 1945-2017 period. The core was subsampled every centimetre for biological analysis (diatoms and cladocera remains).

## Cladoceran community

Shortly, cladoceran stratigraphy  and result of CONISS clustering.

```{r coniss, include=T, echo=F}
require(rioja)
diss<-dist(sqrt(zh[,-c(1)]/100)^2)
clust<-chclust(diss,method="coniss")
par(mfrow=c(2,1))
par(mar=c(5.1,4.1,2.1,20))
bstick(clust, cex=0.8)
par(mar=c(4,3.1,1.1,1.1))
plot(clust, labels=zh[,1], cex=0.8)
```
`r fig_cap("coniss clustering zurich", "Selection of significant groups and constrained hierarchical clustering of Lake Zürich sequence")`

```{r strat plot zurich, include=T, echo=F}
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(zh[,-c(1)],yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=colnames(zh[-c(1)]), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
```
`r fig_cap("strat plot zurich", "Stratigraphic abundance of cladoceran remains in Lake Zürich sediment core ZH17-21. (I'm aware it's impossible to read it with that many taxa, I have the same code somewhere applied to sub groups which make it way easier to read)")`

```{r strat plot zurich selected sp, include=T, echo=F}
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(cbind(zh[,colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")], rowSums(zh[,-1][,!colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")])),yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=c(colnames(zh[colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")]), "others"), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
```
`r fig_cap("strat plot zurich selected sp", "Stratigraphic abundance of dominant cladoceran remains in Lake Zürich sediment core ZH17-21. Abbreviations: BL= Bosmina longirostris, EC= Eubosmina coregoni, EL= Eubosmina longispina, DL= Daphnia spp., LK= Leptodora kindti, BYL= Bythotrephes longismanus, SC= Sida crystallina")`

CONISS clustering identified four significant different cladoceran assemblages over the past 170 years. From 1845 to 1903, _Eubosmina longispina_ and the predator _Bythotrephes longismanus_ dominate the pelagic assemblage. macrophyte-associated species are present ( _Eurycercus_ sp., _Alona quandrangularis_, _Sida crystallina_) attesting for the presence of macrophytic littoral zone. From 1905 to 1908, there is a transient assemblage dominated by pelagic taxa ( _Daphnia longispina_ and _Bosmina longirostris_). The later is a small taxa, wall the usually large Daphnia longispina lose 25 μm in average at this period . This smaller assemblage is associated with a transient disappearance of predator species ( _Leptodora kindti_). This dynamic alone would require more investigation and may come from a change in the top-down pressure, if fish happened to be introduced at that time. However, I did not find any information in the English-written literature mentioning fisheries management of Lake Zurich at that time. The following period from 1911 to 1957 is characterized by a dominance of _Daphnia_ spp. and _Bosmina longirostris_ species in the pelagic compartment. The predator _B. longismanus_ and _Leptodora kindtii_ are also present. The overall abundance of littoral taxa increases, but the macrophyte-associated taxa are replaced by more ubiquitous taxa ( _Chydorus sphaericus_). The last transition took place from 1961, and characterise the recent assemblage. The overall abundance increases for almost all groups. Larger pelagic grazers returned ( _Eubosmina_ sp., larger size _Daphnia_ spp. Figure not included here, refer to thesis). Large predators share the habitat. Chydoridae taxa remained an important part of the assemblage despite eutrophication, unlike what happened in many other lakes of the study. <br>
Boucherle & Züllig (1983)^[Boucherle, M.M., Züllig, H., 1983. Cladoceran remains as evidence of change in trophic state in three Swiss lakes. Hydrobiologia 103, 141–146. https://doi.org/10.1007/BF00028442] used the shift between assemblages dominated by Daphnia spp. and Bosmina sp. to infer the level eutrophication. Using this ratio, we are in accordance with their results and timing of water quality. The insight brought from diatoms analysis on the trophic level also corroborates a two-stages eutrophication, with an increase in water quality in the early years of World War II. Globally, Daphnia spp. group is once again (Alric et al., 2013)^[Alric, B., Möst, M., Domaizon, I., Pignol, C., Spaak, P., Perga, M.-E., 2016. Local human pressures influence gene flow in a hybridizing Daphnia species complex. J. Evol. Biol. 29, 720–735. https://doi.org/10.1111/jeb.12820] providing an early warning for eutrophication with their abundance increasing as soon as 1883. <br>
Having said that, the identification to the specie level in the _Bosmina_ sp. group differs between their study and the present sequence. We did observe the appearance of the small _Bosmina longirostris_ group from the late 19th century, but found remains of _Eubosmina longispina_ during the 20<sup>th</sup> century as well while Boucherle & Züllig (1983) reported its total disappearance from the sediment. Overall, the results are still very similar, and the difference may come from misidentifications or updated identification keys.

# Changepoint analysis

In most paleo-reconstructions, we're interested in the changepoints in the series: when the assemblage shifted. Changepoint are detected on a single vector, so we would usually run a PCA first on the dataset.

Would random sampling generate the same results as if all the samples were analyzed?

We need to loop the different steps:<br>
  1)  Which samples are going to be sampled <br>
  2)  Run PCA on subset <br>
  3)  Run changepoint analysis on 1<sup>st</sup> (and maybe 2<sup>nd</sup>?) component <br>

```{r eval=FALSE, include=FALSE}
require(vegan)
require(ade4)
require(changepoint)

# Change here the sample size
sample_size <- c(rep(5:40, each=3),50)

output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
colnames(output) <- c("Number_samples","PC1_var","PC2_var","AMOC_PC1","AMOC_PC2")

line=0
for (i in sample_size) {
  line=line+1
  # 1. Sample years
  which_samples <- sample(1:nrow(zh), size = i,replace = F)
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

  # 2. Do ordination 
  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
  PCAout<-dudi.pca(PCAout, scannf = FALSE)
  #summary(PCAout)

  # 3. Run changepoints on PC1
  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA

  # 4. Run changepoints on PC2
  cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
  cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
  if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA

  # Save outputs
  output$Number_samples[line] <- i
  output$PC1_var[line]        <- PCAout$eig[1]/sum(PCAout$eig)*100
  output$PC2_var[line]        <- PCAout$eig[2]/sum(PCAout$eig)*100
  output$AMOC_PC1[line]       <- cptmean.AMOC.PC1
  output$AMOC_PC2[line]       <- cptmean.AMOC.PC2

}
output

# # PELT method with AIC penalty
# cptmean.AIC <- cpt.mean(y, method="PELT",penalty = "AIC")
# (cptmean.AIC.date<-as.numeric(time[cpts(cptmean.AIC)]))
# 
# # PELT method with manual penalty (elbow method)
# plot(0,0,xlim=c(0,20), ylim=c(0,15),col ="white", ylab="Number of changes in mean", xlab="Penalty", font=2,main="Elbow plot - choice of the penalty for changepoints analysis \nPruned Exact Linear Time (PELT) method")
# abline(h = c(seq(from=0, to=20, by=1),15), lty=3, col=adjustcolor("black",alpha.f = 0.5))
# abline(h=length(cptmean.AIC.date), lty=3, col="black", lwd=2)
# for (i in seq(from = 0, to = 20, by = 0.05)) {
#   cptmean=cpt.mean(y, method="PELT", penalty = "Manual", pen.value = i)
#   nb <- length(cpts(cptmean))
#   points(nb~i, pch=20)
# }
# legend("topright",legend = c("AIC","Manual - fluctuating penalty"), title = "Methods to chose the number of changes in mean",bty="n", pch = c(NA,20), lty=c(3,NA), lwd=c(2,NA), cex = 0.8)
# # change the pen.value according to the elbow plot
# cptmean.Manual <- cpt.mean(y, method="PELT",penalty = "Manual", pen.value = 20)
# (cptmean.Manual.date<-as.numeric(time[cpts(cptmean.Manual)]))
# 
# # Select here the final changepoint(s) you want to keep
# my_cpt <- (cptmean.Manual.date<-as.numeric(time[cpts(cptmean.Manual)]))
# if(is.null(my_cpt))
#   cat("\nWarning! No changepoints were detected. One of the condition for alternative stable state\nis an abrupt transition (see Andersen et al 2009, and our discussion in Bruel et al 2O18).\n\n")
# if(!is.null(my_cpt)) {
#   my_cpt <- as.data.frame(matrix(c(my_cpt,rep(NA,length(my_cpt))),ncol=2,byrow=F))
#   for (i in 1:nrow(my_cpt)) my_cpt[i,2]=time[which(time==my_cpt[i,1])+1]
#   # Re-arrange you my_cpt matrix
#   my_cpt <- my_cpt[order(my_cpt[,1],decreasing = T),order(my_cpt[1,],decreasing = T)]
# }

```

```{r}
require(ggplot2)
require(gridExtra)
p1 <- ggplot(output, aes(Number_samples, PC1_var)) + geom_point() + stat_smooth()
p2 <- ggplot(output, aes(Number_samples, PC2_var)) + geom_point() + stat_smooth()
grid.arrange(p1,p2, nrow=1)
```
`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`

```{r}
p1 <- ggplot(output, aes(AMOC_PC1, Number_samples)) + geom_point() + xlim(1850,2017)
p2 <- ggplot(output, aes(AMOC_PC2,Number_samples)) + geom_point() + xlim(1850,2017)
grid.arrange(p1,p2, nrow=2)
```
`r fig_cap("timing of cpts", "Timing of changepoint")`

# Optimizing sampling by adding closest sample to shift

When analysing more or less sample is easy (e.g., counting cladoceran samples vs analyzing DNA samples), can we find an optimal sampling?

Logic is a bit different, we want to know what is the minimum number of random sample we should start with, and then add samples around it to actually find the shifting point

```{r}
# Change here the sample size
sample_size <- c(rep(13:30, each=100))
#sample_size <- 15
output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")

output <- output[,c(1,2,4)]

for (i in 1:length(sample_size)) {
  if(i == 1)  {
    mdata <- zh[order(zh[,1]),]
    # 2. Do ordination 
    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
    PCAout<-dudi.pca(PCAout, scannf = FALSE)
    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100

    # 3. Run changepoints on PC1
    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA

    # # 4. Run changepoints on PC2
    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA

    # Save outputs
    c(nrow(zh), true.eig1, true.eig2,
      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
  } 
  # 1. Sample years
  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
  which_samples <- c(which_samples, 1, nrow(zh))
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

  # 2. Do ordination 
  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
  PCAout<-dudi.pca(PCAout, scannf = FALSE)

  # 3. Run changepoints on PC1
  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA

  # # 4. Run changepoints on PC2
  # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
  # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
  # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
  # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA

  # Save outputs
  output$Initial_number_samples[i]     <- sample_size[i]
  output$Initial_AMOC_PC1[i]           <- cptmean.AMOC.PC1
  #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1) <8){
        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
        numberPC1_ok = TRUE
        } else numberPC1_ok = FALSE
  # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
  #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
  #   numberPC2_ok = TRUE
  # } else numberPC2_ok = FALSE

  # 5. Now, how many more samples are needed to get true cptmean.AMOC.PC1?
  for (j in 1:(nrow(zh)-sample_size[i])) {
    if(!numberPC1_ok) {
      h=1
      while((which_samples[cpts(cptmean.AMOC.PC1_num)]-h) %in% which_samples & which_samples[cpts(cptmean.AMOC.PC1_num)]-h > 0) h=h+1
      which_samples <- c(which_samples, which_samples[cpts(cptmean.AMOC.PC1_num)]-h)

      h=1
      while((which_samples[cpts(cptmean.AMOC.PC1_num)]+h) %in% which_samples & which_samples[cpts(cptmean.AMOC.PC1_num)]+h<nrow(zh)) h=h+1
      which_samples <- c(which_samples, which_samples[cpts(cptmean.AMOC.PC1_num)]+h)

      which_samples <- which_samples[which_samples>0 & which_samples<= nrow(zh)]
      which_samples <- unique(which_samples)

    which_samples <- which_samples[order(which_samples)]
    mdata <- zh[which_samples,]
    # reorder chronologically
    mdata <- mdata[order(mdata[,1]),]

    # 2. Do ordination 
    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
    PCAout<-dudi.pca(PCAout, scannf = FALSE)
    #summary(PCAout)

    # 3. Run changepoints on PC1
    cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
    cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
    cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
    if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA

    # # 4. Run changepoints on PC2
    # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
    # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
    # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
    # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA

    if(!is.na(cptmean.AMOC.PC1) & !is.na(true.cptmean.AMOC.PC1) & cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
        numberPC1_ok = TRUE
        } else numberPC1_ok = FALSE
    # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
    #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
    #   numberPC2_ok = TRUE
    # } else numberPC2_ok = FALSE
  }  
    }

}
output$total_added <- output$Number_needed_for_AMOC_PC1-output$Initial_number_samples-2
output
```

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point(alpha=.3)
library(dplyr)
output2 <- output %>% 
  group_by(Initial_number_samples) %>% 
  summarise(total0 = sum(total_added==0))

plot(output2)
abline(h=80)
```
`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`

# Optimizing sampling by adding intermediate samples between shift and next sample analyzed

```{r}
# Change here the sample size
sample_size <- c(rep(15:30, each=30))
#sample_size <- 15
output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")

output <- output[,c(1,2,4)]

for (i in 1:length(sample_size)) {
  if(i == 1)  {
    mdata <- zh[order(zh[,1]),]
    # 2. Do ordination 
    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
    PCAout<-dudi.pca(PCAout, scannf = FALSE)
    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100

    # 3. Run changepoints on PC1
    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA

    # # 4. Run changepoints on PC2
    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA

    # Save outputs
    c(nrow(zh), true.eig1, true.eig2,
      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
  } 
  # 1. Sample years
  which_samples <- sample(1:nrow(zh), size = sample_size[i],replace = F)
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

  # 2. Do ordination 
  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
  PCAout<-dudi.pca(PCAout, scannf = FALSE)

  # 3. Run changepoints on PC1
  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA

  # # 4. Run changepoints on PC2
  # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
  # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
  # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
  # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA

  # Save outputs
  output$Initial_number_samples[i]     <- sample_size[i]
  output$Initial_AMOC_PC1[i]           <- cptmean.AMOC.PC1
  #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
  if(cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
        numberPC1_ok = TRUE
        } else numberPC1_ok = FALSE
  # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
  #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
  #   numberPC2_ok = TRUE
  # } else numberPC2_ok = FALSE

  # 5. Now, how many more samples are needed to get true cptmean.AMOC.PC1?
  for (j in 1:(nrow(zh)-sample_size[i])) {
    if(!numberPC1_ok) {

      h=0
      max_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)+h]+which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])/2)
      while(max_samp %in% which_samples & (cpts(cptmean.AMOC.PC1_num)+h+1)<=nrow(zh)) {
        h=h+1
        other_limit <- ifelse(length(which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])<=nrow(zh),
                              which_samples[cpts(cptmean.AMOC.PC1_num)+h+1],
                              nrow(zh))
        max_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)+h]+which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])/2)
        }

      h=0
      min_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)-h]+which_samples[cpts(cptmean.AMOC.PC1_num)-h-1])/2)
      while(min_samp %in% which_samples & (cpts(cptmean.AMOC.PC1_num)-h-1)>0) {
        h=h+1
        other_limit <- ifelse(length(which_samples[cpts(cptmean.AMOC.PC1_num)-h-1])>0,
                              which_samples[cpts(cptmean.AMOC.PC1_num)-h-1],
                              1)
        min_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)-h]+other_limit)/2)
      }

      which_samples <- c(which_samples, max_samp, min_samp)
      which_samples <- which_samples[!is.na(which_samples)]
      which_samples <- which_samples[which_samples>0 & which_samples<= nrow(zh)]
      which_samples <- unique(which_samples)

    which_samples <- which_samples[order(which_samples)]
    mdata <- zh[which_samples,]
    # reorder chronologically
    mdata <- mdata[order(mdata[,1]),]

    # 2. Do ordination 
    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
    PCAout<-dudi.pca(PCAout, scannf = FALSE)
    #summary(PCAout)

    # 3. Run changepoints on PC1
    cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
    cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
    cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
    if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA

    # # 4. Run changepoints on PC2
    # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
    # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
    # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
    # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA

    if(!is.na(cptmean.AMOC.PC1) & !is.na(true.cptmean.AMOC.PC1) & cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
        numberPC1_ok = TRUE
        } else numberPC1_ok = FALSE
    # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
    #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
    #   numberPC2_ok = TRUE
    # } else numberPC2_ok = FALSE
  }  
    }

}
output$total_added <- output$Number_needed_for_AMOC_PC1-output$Initial_number_samples
output
```

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point()
```
`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`

# Limits

A few methods looking at critical transition (e.g., Taranu et al 2018^[Taranu, Z.E., Carpenter, S.R., Frossard, V., Jenny, J.-P., Thomas, Z., Vermaire, J.C., Perga, M.-E., 2018. Can we detect ecosystem critical transitions and signals of changing resilience from paleo-ecological records? Ecosphere 9, e02438. https://doi.org/10.1002/ecs2.2438]), or increase variance (e.g., Dakos et al 2012^[Dakos, V., Carpenter, S.R., Brock, W.A., Ellison, A.M., Guttal, V., Ives, A.R., Kéfi, S., Livina, V., Seekell, D.A., van Nes, E.H., Scheffer, M., 2012. Methods for Detecting Early Warnings of Critical Transitions in Time Series Illustrated Using Simulated Ecological Data. PloS One 7, e41010. https://doi.org/10.1371/journal.pone.0041010]), etc., will value having more continuous sampling (although the online DLMs method in Taranu et al 2018 allows for non-continuous sampling).

### 2019-10-17 R/temporal-sampling.Rmd
adding the figure of error against initial number sampled

@@
 tab_cap <- captioner::captioner("Table")
+
+# Libraries
+library(dplyr)
+require(rioja)
+require(vegan)
+require(ade4)
+require(changepoint)
+require(ggplot2)
+require(gridExtra)
+require(plotly)
+library(reshape)
 ```
@@
 ```{r read zurich data, include=T}
-zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt"))
+zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt")); is.zh=T
+zh <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
 ```
@@
 ```{r coniss, include=T, echo=F}
-require(rioja)
 diss<-dist(sqrt(zh[,-c(1)]/100)^2)
@@
 ```
-`r fig_cap("coniss clustering zurich", "Selection of significant groups and constrained hierarchical clustering of Lake Zürich sequence")`
+<br>`r fig_cap("coniss clustering zurich", "Selection of significant groups and constrained hierarchical clustering of Lake Zürich sequence")`
 
@@
 ```
-`r fig_cap("strat plot zurich", "Stratigraphic abundance of cladoceran remains in Lake Zürich sediment core ZH17-21.
+<br>`r fig_cap("strat plot zurich", "Stratigraphic abundance of cladoceran remains in Lake Zürich sediment core ZH17-21.
 (I'm aware it's impossible to read it with that many taxa, I have the same code somewhere applied to sub groups which make it way easier to read)")`
@@
 ```{r strat plot zurich selected sp, include=T, echo=F}
+if(is.zh){
   par(mfrow=c(1,1))
@@
 addClustZone(clado.clust,clust,4,col="black")
+}
 ```
-`r fig_cap("strat plot zurich selected sp", "Stratigraphic abundance of dominant cladoceran remains in Lake Zürich sediment core ZH17-21.
+<br>`r fig_cap("strat plot zurich selected sp", "Stratigraphic abundance of dominant cladoceran remains in Lake Zürich sediment core ZH17-21.
 Abbreviations: BL= Bosmina longirostris, EC= Eubosmina coregoni, EL= Eubosmina longispina, DL= Daphnia spp., LK= Leptodora kindti, BYL= Bythotrephes longismanus, SC= Sida crystallina")`
@@
 
-```{r eval=FALSE, include=FALSE}
-require(vegan)
-require(ade4)
-require(changepoint)
-
+```{r include=FALSE}
 # Change here the sample size
@@
   # 1. Sample years
-  which_samples <- sample(1:nrow(zh), size = i,replace = F)
+  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
+  which_samples <- c(which_samples,1,nrow(zh))
   which_samples <- which_samples[order(which_samples)]
@@
 ```{r}
-require(ggplot2)
-require(gridExtra)
 p1 <- ggplot(output, aes(Number_samples, PC1_var)) + geom_point() + stat_smooth()
@@
 ```
-`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`
+<br>`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`
 
@@
 ```
-`r fig_cap("timing of cpts", "Timing of changepoint")`
+<br>`r fig_cap("timing of cpts", "Timing of changepoint")`
 
@@
 # Change here the sample size
-sample_size <- c(rep(13:30, each=100))
+sample_size <- c(rep(13:30, each=30))
 #sample_size <- 15
@@
     # # 4. Run changepoints on PC2
+    true.cptmean.AMOC.PC2 <- NA
     # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
@@
   #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
-  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1) <8){
+  if(cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
         output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
@@
 ggplot(output, aes(Initial_number_samples,total_added)) + geom_point(alpha=.3)
-library(dplyr)
-output2 <- output %>% 
-  group_by(Initial_number_samples) %>% 
-  summarise(total0 = sum(total_added==0))
 
-plot(output2)
-abline(h=80)
 ```
-`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`
+<br>`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`
 
@@
   # 1. Sample years
-  which_samples <- sample(1:nrow(zh), size = sample_size[i],replace = F)
+  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
+  which_samples <- c(which_samples,1,nrow(zh))
   which_samples <- which_samples[order(which_samples)]
@@
 ```
-`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`
+<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`
+
+```{r}
+output2 <- output %>% 
+  group_by(Initial_number_samples) %>% 
+  summarise(total0 = sum(total_added==0))
+
+plot(output2)
+abline(h=80)
+```
+
+# Chances of getting close to the real break
+
+Following discussion with Easton on 10/10/2019.
+
+```{r}
+# Change here the threshold year: how close you need to get close to the actual year
+threshold_yr = c(2,4,6,8,10,12,14)
+
+# Change here the sample size
+sample_size <- c(rep(15:50, each=100))
+#sample_size <- 15
+output3 <- as.data.frame(matrix(rep(NA, 4*length(sample_size)*length(threshold_yr)), ncol=4))
+colnames(output3) <- c("Initial_number_samples","threshold_yr","Initial_AMOC_PC1","Within_threshold")
+
+for (i in 1:length(sample_size)) {
+  if(i == 1)  {
+    row=0
+    mdata <- zh[order(zh[,1]),]
+    # 2. Do ordination 
+    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
+    PCAout<-dudi.pca(PCAout, scannf = FALSE)
+    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
+    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100
+
+    # 3. Run changepoints on PC1
+    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
+    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
+    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA
+
+    # # 4. Run changepoints on PC2
+    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
+    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
+    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA
+
+    # Save outputs
+    c(nrow(zh), true.eig1, true.eig2,
+      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
+  }
+
+  for (j in 1:length(threshold_yr)) {
+    row=row+1
+      # 1. Sample years
+  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
+  which_samples <- c(which_samples,1,nrow(zh))
+  which_samples <- which_samples[order(which_samples)]
+  mdata <- zh[which_samples,]
+  # reorder chronologically
+  mdata <- mdata[order(mdata[,1]),]
+  #remove rare species
+  # present less than 3 times in the assemblage
+  mdata <- mdata[,colSums(mdata>0)>3]
+
+  # 2. Do ordination 
+  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
+  PCAout<-dudi.pca(PCAout, scannf = FALSE)
+
+  # 3. Run changepoints on PC1
+  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
+  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
+  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
+  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- 0
+
+  # Save outputs
+  output3$Initial_number_samples[row]     <- sample_size[i]
+  output3$threshold_yr[row]               <- threshold_yr[j]
+  output3$Initial_AMOC_PC1[row]           <- cptmean.AMOC.PC1
+  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1)<=threshold_yr[j]){
+        numberPC1_ok = 1
+        } else {numberPC1_ok = 0}
+
+  output3$Within_threshold[row] <- numberPC1_ok
+  }
+
+}
+head(output3)
+tail(output3)
+```
+
+```{r summary }
+names(output3)
+
+outsumm <- as.data.frame(with(output3, tapply(Within_threshold,list("Initial_number_samples"=Initial_number_samples,"threshold_yr"=threshold_yr), sum)))
+outsumm$InitialNumber <- as.numeric(rownames(outsumm))
+outsumm <- melt(outsumm, id.vars = "InitialNumber")
+outsumm$more80 <- ifelse(outsumm$value>=80,1,0)
+ggplot(outsumm, aes(InitialNumber, variable, fill=more80)) + geom_tile() +
+  xlab("Sample analyzed") + ylab("Tolerance for error (year)")
+
+```
+<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`
 
 # Limits
+
+Is changepoint necesseraly what we want? We can be interested in changes in mean and/or variance. Right now, I've just looked at mean.
 

### 2019-10-24 R/uncertainty_tolerance.R
compute change in mean and var on a vector and its sub-samples

@@ -1,0 +1,108 @@
+# compute the tolerance to error against sample size
+
+require(changepoint)
+
+#uncertainty tolerance function
+ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"), threshold_error = 1, each = 10) {
+  if(!is.vector(y)) stop("y is required and must be a vector")
+  if(is.null(x)) x <- 1:length(y)
+  if(length(x) != length(y)) stop("x and y must be the same length")
+  if(is.null(sample_size)) sample_size <- 3:length(y)
+  if(max(sample_size)>length(y))   stop("you cannot sample a greater number than the initial vector y") 
+  
+  sample_size <- rep(sample_size,each=each)
+  
+  out_list <- list()
+  for (i in 1:length(sample_size)) {
+    # Case 1: changepoint in mean
+    if("mean" %in% type) {
+      # Get the "TRUE" changepoint
+      if (i==1) {
+        true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
+        true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
+        if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
+      }
+      # Get the changepoint for the subsample
+      if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.") else
+      {
+        for (j in 1:length(threshold_error)) {
+          # 1. Sample 
+          which_samples <- c(1,length(y),sample(2:(length(y)-1), size = sample_size[i],replace = F))
+          which_samples <- which_samples[order(which_samples)]
+          y2 <- y[which_samples]
+          x2 <- x[which_samples]
+          
+          # 2. Run changepoints 
+          cptmean.AMOC.y2 <- cpt.mean(y2, method="AMOC") 
+          cptmean.AMOC.y2 <- as.numeric(x2[cpts(cptmean.AMOC.y2)])
+          if(length(cptmean.AMOC.y2)==0) cptmean.AMOC.y2 <- 0
+          
+          # 3. Save outputs
+          if (i == 1) 
+            out_list$cptmean_out <- data.frame("sample_size"=sample_size[i],
+                                                         "threshold_error"=threshold_error[j],
+                                                         "cptmean"=ifelse(cptmean.AMOC.y2==0,NA,cptmean.AMOC.y2),
[…]

### 2019-10-24 ar1_process.R
added ar1 code

@@ -1,0 +1,26 @@
+
+# Run a AR-1 time series with a large change forced in the middle
+
+# Set simulation parameters
+mu = 0
+sigma = 0.1
+phi = 1
+shift_time = 50
+max_time = 100
+
+# Set up vectors
+ar1 <- vector(mode = 'numeric',100)
+ar1[1] <- 10
+
+# Build time series
+for (t in 2:max_time){
+  if(t == shift_time){ar1[t-1]=-10}
+  ar1[t] <- ar1[t-1] + phi*rnorm(1,mean=mu,sd = sigma)
+
+}
+
+# standarize the data
+ar1_s <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+
+
+plot(ar1_s,ylim=c(0,1),type='l')

### 2019-10-24 R/uncertainty_tolerance.R
add the 3 options in the function, random sampling, regular sampling, convergent sampling.

@@ -3,106 +3,247 @@
 require(changepoint)
 
 #uncertainty tolerance function
-ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"), threshold_error = 1, each = 10) {
+ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
+               threshold_error = 1, each = 10,
+               random_sampling=T,
+               regular_sampling=T,
+               convergence_sampling=T) {
   if(!is.vector(y)) stop("y is required and must be a vector")
   if(is.null(x)) x <- 1:length(y)
   if(length(x) != length(y)) stop("x and y must be the same length")
   if(is.null(sample_size)) sample_size <- 3:length(y)
   if(max(sample_size)>length(y))   stop("you cannot sample a greater number than the initial vector y") 
   
-  sample_size <- rep(sample_size,each=each)
   
   out_list <- list()
-  for (i in 1:length(sample_size)) {
-    # Case 1: changepoint in mean
-    if("mean" %in% type) {
-      # Get the "TRUE" changepoint
-      if (i==1) {
-        true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
-        true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
-        if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
-      }
-      # Get the changepoint for the subsample
-      if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.") else
-      {
-        for (j in 1:length(threshold_error)) {
-          # 1. Sample 
-          which_samples <- c(1,length(y),sample(2:(length(y)-1), size = sample_size[i],replace = F))
-          which_samples <- which_samples[order(which_samples)]
-          y2 <- y[which_samples]
-          x2 <- x[which_samples]
-          
-          # 2. Run changepoints 
-          cptmean.AMOC.y2 <- cpt.mean(y2, method="AMOC") 
-          cptmean.AMOC.y2 <- as.numeric(x2[cpts(cptmean.AMOC.y2)])
-          if(length(cptmean.AMOC.y2)==0) cptmean.AMOC.y2 <- 0
-          
-          # 3. Save outputs
-          if (i == 1) 
-            out_list$cptmean_out <- data.frame("sample_size"=sample_size[i],
[…]

### 2019-10-24 R/temporal-sampling.Rmd
clean-up the Rmd from all the attempts

@@
 ---
+
+**Note for next time: write function to compute the tolerance to error against sample size**
 
@@
 
-We need to loop the different steps:<br>
-  1)  Which samples are going to be sampled <br>
-  2)  Run PCA on subset <br>
-  3)  Run changepoint analysis on 1<sup>st</sup> (and maybe 2<sup>nd</sup>?) component <br>
-
-```{r include=FALSE}
-# Change here the sample size
-sample_size <- c(rep(5:40, each=3),50)
-
-output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
-colnames(output) <- c("Number_samples","PC1_var","PC2_var","AMOC_PC1","AMOC_PC2")
-
-line=0
-for (i in sample_size) {
-  line=line+1
-  # 1. Sample years
-  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
-  which_samples <- c(which_samples,1,nrow(zh))
-  which_samples <- which_samples[order(which_samples)]
-  mdata <- zh[which_samples,]
-  # reorder chronologically
-  mdata <- mdata[order(mdata[,1]),]
-  #remove rare species
-  # present less than 3 times in the assemblage
-  mdata <- mdata[,colSums(mdata>0)>3]
-
-  # 2. Do ordination 
-  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-  PCAout<-dudi.pca(PCAout, scannf = FALSE)
-  #summary(PCAout)
-
-  # 3. Run changepoints on PC1
-  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA
-
-  # 4. Run changepoints on PC2
-  cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-  cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
-  if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA
-
-  # Save outputs
-  output$Number_samples[line] <- i
-  output$PC1_var[line]        <- PCAout$eig[1]/sum(PCAout$eig)*100
-  output$PC2_var[line]        <- PCAout$eig[2]/sum(PCAout$eig)*100
-  output$AMOC_PC1[line]       <- cptmean.AMOC.PC1
-  output$AMOC_PC2[line]       <- cptmean.AMOC.PC2
-
-}
-output
-
-# # PELT method with AIC penalty
-# cptmean.AIC <- cpt.mean(y, method="PELT",penalty = "AIC")
-# (cptmean.AIC.date<-as.numeric(time[cpts(cptmean.AIC)]))
-# 
-# # PELT method with manual penalty (elbow method)
-# plot(0,0,xlim=c(0,20), ylim=c(0,15),col ="white", ylab="Number of changes in mean", xlab="Penalty", font=2,main="Elbow plot - choice of the penalty for changepoints analysis \nPruned Exact Linear Time (PELT) method")
-# abline(h = c(seq(from=0, to=20, by=1),15), lty=3, col=adjustcolor("black",alpha.f = 0.5))
-# abline(h=length(cptmean.AIC.date), lty=3, col="black", lwd=2)
-# for (i in seq(from = 0, to = 20, by = 0.05)) {
-#   cptmean=cpt.mean(y, method="PELT", penalty = "Manual", pen.value = i)
-#   nb <- length(cpts(cptmean))
-#   points(nb~i, pch=20)
-# }
-# legend("topright",legend = c("AIC","Manual - fluctuating penalty"), title = "Methods to chose the number of changes in mean",bty="n", pch = c(NA,20), lty=c(3,NA), lwd=c(2,NA), cex = 0.8)
-# # change the pen.value according to the elbow plot
-# cptmean.Manual <- cpt.mean(y, method="PELT",penalty = "Manual", pen.value = 20)
-# (cptmean.Manual.date<-as.numeric(time[cpts(cptmean.Manual)]))
-# 
-# # Select here the final changepoint(s) you want to keep
-# my_cpt <- (cptmean.Manual.date<-as.numeric(time[cpts(cptmean.Manual)]))
-# if(is.null(my_cpt))
-#   cat("\nWarning! No changepoints were detected. One of the condition for alternative stable state\nis an abrupt transition (see Andersen et al 2009, and our discussion in Bruel et al 2O18).\n\n")
-# if(!is.null(my_cpt)) {
-#   my_cpt <- as.data.frame(matrix(c(my_cpt,rep(NA,length(my_cpt))),ncol=2,byrow=F))
-#   for (i in 1:nrow(my_cpt)) my_cpt[i,2]=time[which(time==my_cpt[i,1])+1]
-#   # Re-arrange you my_cpt matrix
-#   my_cpt <- my_cpt[order(my_cpt[,1],decreasing = T),order(my_cpt[1,],decreasing = T)]
-# }
-
-```
-
 ```{r}
@@
 
+# Types of sampling
+
+1. Random <br>
+2. Regular <br>
+3. Convergent <br>
+
 ```{r}
-p1 <- ggplot(output, aes(AMOC_PC1, Number_samples)) + geom_point() + xlim(1850,2017)
-p2 <- ggplot(output, aes(AMOC_PC2,Number_samples)) + geom_point() + xlim(1850,2017)
-grid.arrange(p1,p2, nrow=2)
+source("ar1_process.R")
+source("R/uncertainty_tolerance.R")
+args(ut)
 ```
-<br>`r fig_cap("timing of cpts", "Timing of changepoint")`
-
-# Optimizing sampling by adding closest sample to shift
 
@@
 
-```{r}
-# Change here the sample size
-sample_size <- c(rep(13:30, each=30))
-#sample_size <- 15
-output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
-colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")
+```{r, cache = T}
+output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = c(2:10))
+names(output)
+head(output$convergence_cptmean)
+ggplot(output$convergence_cptmean, aes(sample_size,number_sample_added)) + geom_point(alpha=.3)
 
-output <- output[,c(1,2,4)]
-
-for (i in 1:length(sample_size)) {
-  if(i == 1)  {
-    mdata <- zh[order(zh[,1]),]
-    # 2. Do ordination 
-    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-    PCAout<-dudi.pca(PCAout, scannf = FALSE)
-    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
-    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100
-
-    # 3. Run changepoints on PC1
-    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
-    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA
-
-    # # 4. Run changepoints on PC2
-    true.cptmean.AMOC.PC2 <- NA
-    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
-    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA
-
-    # Save outputs
-    c(nrow(zh), true.eig1, true.eig2,
-      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
-  } 
-  # 1. Sample years
-  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
-  which_samples <- c(which_samples, 1, nrow(zh))
-  which_samples <- which_samples[order(which_samples)]
-  mdata <- zh[which_samples,]
-  # reorder chronologically
-  mdata <- mdata[order(mdata[,1]),]
-  #remove rare species
-  # present less than 3 times in the assemblage
-  mdata <- mdata[,colSums(mdata>0)>3]
-
-  # 2. Do ordination 
-  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-  PCAout<-dudi.pca(PCAout, scannf = FALSE)
-
-  # 3. Run changepoints on PC1
-  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
-  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA
-
-  # # 4. Run changepoints on PC2
-  # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-  # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
-  # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
-  # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA
-
-  # Save outputs
-  output$Initial_number_samples[i]     <- sample_size[i]
-  output$Initial_AMOC_PC1[i]           <- cptmean.AMOC.PC1
-  #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
-  if(cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
-        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
-        numberPC1_ok = TRUE
-        } else numberPC1_ok = FALSE
-  # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
-  #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
-  #   numberPC2_ok = TRUE
-  # } else numberPC2_ok = FALSE
-
-  # 5. Now, how many more samples are needed to get true cptmean.AMOC.PC1?
-  for (j in 1:(nrow(zh)-sample_size[i])) {
-    if(!numberPC1_ok) {
-      h=1
-      while((which_samples[cpts(cptmean.AMOC.PC1_num)]-h) %in% which_samples & which_samples[cpts(cptmean.AMOC.PC1_num)]-h > 0) h=h+1
-      which_samples <- c(which_samples, which_samples[cpts(cptmean.AMOC.PC1_num)]-h)
-
-      h=1
-      while((which_samples[cpts(cptmean.AMOC.PC1_num)]+h) %in% which_samples & which_samples[cpts(cptmean.AMOC.PC1_num)]+h<nrow(zh)) h=h+1
-      which_samples <- c(which_samples, which_samples[cpts(cptmean.AMOC.PC1_num)]+h)
-
-      which_samples <- which_samples[which_samples>0 & which_samples<= nrow(zh)]
-      which_samples <- unique(which_samples)
-
-    which_samples <- which_samples[order(which_samples)]
-    mdata <- zh[which_samples,]
-    # reorder chronologically
-    mdata <- mdata[order(mdata[,1]),]
-
-    # 2. Do ordination 
-    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-    PCAout<-dudi.pca(PCAout, scannf = FALSE)
-    #summary(PCAout)
-
-    # 3. Run changepoints on PC1
-    cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-    cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
-    cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-    if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA
-
-    # # 4. Run changepoints on PC2
-    # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-    # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
-    # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
-    # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA
-
-    if(!is.na(cptmean.AMOC.PC1) & !is.na(true.cptmean.AMOC.PC1) & cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
-        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
-        numberPC1_ok = TRUE
-        } else numberPC1_ok = FALSE
-    # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
-    #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
-    #   numberPC2_ok = TRUE
-    # } else numberPC2_ok = FALSE
-  }  
-    }
-
-}
-output$total_added <- output$Number_needed_for_AMOC_PC1-output$Initial_number_samples-2
-output
-```
-
-```{r}
-ggplot(output, aes(Initial_number_samples,total_added)) + geom_point(alpha=.3)
-
-```
-<br>`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`
-
-# Optimizing sampling by adding intermediate samples between shift and next sample analyzed
-
-```{r}
-# Change here the sample size
-sample_size <- c(rep(15:30, each=30))
-#sample_size <- 15
-output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
-colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")
-
-output <- output[,c(1,2,4)]
-
-for (i in 1:length(sample_size)) {
-  if(i == 1)  {
-    mdata <- zh[order(zh[,1]),]
-    # 2. Do ordination 
-    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-    PCAout<-dudi.pca(PCAout, scannf = FALSE)
-    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
-    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100
-
-    # 3. Run changepoints on PC1
-    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
-    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA
-
-    # # 4. Run changepoints on PC2
-    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
-    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA
-
-    # Save outputs
-    c(nrow(zh), true.eig1, true.eig2,
-      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
-  } 
-  # 1. Sample years
-  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
-  which_samples <- c(which_samples,1,nrow(zh))
-  which_samples <- which_samples[order(which_samples)]
-  mdata <- zh[which_samples,]
-  # reorder chronologically
-  mdata <- mdata[order(mdata[,1]),]
-  #remove rare species
-  # present less than 3 times in the assemblage
-  mdata <- mdata[,colSums(mdata>0)>3]
-
-  # 2. Do ordination 
-  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-  PCAout<-dudi.pca(PCAout, scannf = FALSE)
-
-  # 3. Run changepoints on PC1
-  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
-  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA
-
-  # # 4. Run changepoints on PC2
-  # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-  # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
-  # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
-  # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA
-
-  # Save outputs
-  output$Initial_number_samples[i]     <- sample_size[i]
-  output$Initial_AMOC_PC1[i]           <- cptmean.AMOC.PC1
-  #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
-  if(cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
-        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
-        numberPC1_ok = TRUE
-        } else numberPC1_ok = FALSE
-  # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
-  #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
-  #   numberPC2_ok = TRUE
-  # } else numberPC2_ok = FALSE
-
-  # 5. Now, how many more samples are needed to get true cptmean.AMOC.PC1?
-  for (j in 1:(nrow(zh)-sample_size[i])) {
-    if(!numberPC1_ok) {
-
-      h=0
-      max_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)+h]+which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])/2)
-      while(max_samp %in% which_samples & (cpts(cptmean.AMOC.PC1_num)+h+1)<=nrow(zh)) {
-        h=h+1
-        other_limit <- ifelse(length(which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])<=nrow(zh),
-                              which_samples[cpts(cptmean.AMOC.PC1_num)+h+1],
-                              nrow(zh))
-        max_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)+h]+which_samples[cpts(cptmean.AMOC.PC1_num)+h+1])/2)
-        }
-
-      h=0
-      min_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)-h]+which_samples[cpts(cptmean.AMOC.PC1_num)-h-1])/2)
-      while(min_samp %in% which_samples & (cpts(cptmean.AMOC.PC1_num)-h-1)>0) {
-        h=h+1
-        other_limit <- ifelse(length(which_samples[cpts(cptmean.AMOC.PC1_num)-h-1])>0,
-                              which_samples[cpts(cptmean.AMOC.PC1_num)-h-1],
-                              1)
-        min_samp <- round((which_samples[cpts(cptmean.AMOC.PC1_num)-h]+other_limit)/2)
-      }
-
-      which_samples <- c(which_samples, max_samp, min_samp)
-      which_samples <- which_samples[!is.na(which_samples)]
-      which_samples <- which_samples[which_samples>0 & which_samples<= nrow(zh)]
-      which_samples <- unique(which_samples)
-
-    which_samples <- which_samples[order(which_samples)]
-    mdata <- zh[which_samples,]
-    # reorder chronologically
-    mdata <- mdata[order(mdata[,1]),]
-
-    # 2. Do ordination 
-    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-    PCAout<-dudi.pca(PCAout, scannf = FALSE)
-    #summary(PCAout)
-
-    # 3. Run changepoints on PC1
-    cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-    cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
-    cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-    if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- NA
-
-    # # 4. Run changepoints on PC2
-    # cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-    # cptmean.AMOC.PC2_num <- cptmean.AMOC.PC2
-    # cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC2)])
-    # if(length(cptmean.AMOC.PC2)==0) cptmean.AMOC.PC2 <- NA
-
-    if(!is.na(cptmean.AMOC.PC1) & !is.na(true.cptmean.AMOC.PC1) & cptmean.AMOC.PC1==true.cptmean.AMOC.PC1){
-        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
-        numberPC1_ok = TRUE
-        } else numberPC1_ok = FALSE
-    # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
-    #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
-    #   numberPC2_ok = TRUE
-    # } else numberPC2_ok = FALSE
-  }  
-    }
-
-}
-output$total_added <- output$Number_needed_for_AMOC_PC1-output$Initial_number_samples
-output
-```
-
-```{r}
-ggplot(output, aes(Initial_number_samples,total_added)) + geom_point()
 ```
@@
 
-```{r}
-output2 <- output %>% 
-  group_by(Initial_number_samples) %>% 
-  summarise(total0 = sum(total_added==0))
+```{r summary }
+names(output$random_cptmean)
 
-plot(output2)
-abline(h=80)
-```
-
-# Chances of getting close to the real break
-
-Following discussion with Easton on 10/10/2019.
-
-```{r}
-# Change here the threshold year: how close you need to get close to the actual year
-threshold_yr = c(2,4,6,8,10,12,14)
-
-# Change here the sample size
-sample_size <- c(rep(15:50, each=100))
-#sample_size <- 15
-output3 <- as.data.frame(matrix(rep(NA, 4*length(sample_size)*length(threshold_yr)), ncol=4))
-colnames(output3) <- c("Initial_number_samples","threshold_yr","Initial_AMOC_PC1","Within_threshold")
-
-for (i in 1:length(sample_size)) {
-  if(i == 1)  {
-    row=0
-    mdata <- zh[order(zh[,1]),]
-    # 2. Do ordination 
-    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-    PCAout<-dudi.pca(PCAout, scannf = FALSE)
-    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
-    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100
-
-    # 3. Run changepoints on PC1
-    true.cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-    true.cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC1)])
-    if(length(true.cptmean.AMOC.PC1)==0) true.cptmean.AMOC.PC1 <- NA
-
-    # # 4. Run changepoints on PC2
-    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
-    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
-    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA
-
-    # Save outputs
-    c(nrow(zh), true.eig1, true.eig2,
-      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
-  }
-
-  for (j in 1:length(threshold_yr)) {
-    row=row+1
-      # 1. Sample years
-  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
-  which_samples <- c(which_samples,1,nrow(zh))
-  which_samples <- which_samples[order(which_samples)]
-  mdata <- zh[which_samples,]
-  # reorder chronologically
-  mdata <- mdata[order(mdata[,1]),]
-  #remove rare species
-  # present less than 3 times in the assemblage
-  mdata <- mdata[,colSums(mdata>0)>3]
-
-  # 2. Do ordination 
-  PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
-  PCAout<-dudi.pca(PCAout, scannf = FALSE)
-
-  # 3. Run changepoints on PC1
-  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
-  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
-  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
-  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- 0
-
-  # Save outputs
-  output3$Initial_number_samples[row]     <- sample_size[i]
-  output3$threshold_yr[row]               <- threshold_yr[j]
-  output3$Initial_AMOC_PC1[row]           <- cptmean.AMOC.PC1
-  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1)<=threshold_yr[j]){
-        numberPC1_ok = 1
-        } else {numberPC1_ok = 0}
-
-  output3$Within_threshold[row] <- numberPC1_ok
-  }
-
-}
-head(output3)
-tail(output3)
-```
-
-```{r summary }
-names(output3)
-
-outsumm <- as.data.frame(with(output3, tapply(Within_threshold,list("Initial_number_samples"=Initial_number_samples,"threshold_yr"=threshold_yr), sum)))
+outsumm <- as.data.frame(with(output$random_cptmean, tapply(same_cptmean,list("Initial_number_samples"=sample_size,"threshold_yr"=threshold_error), sum)))
 outsumm$InitialNumber <- as.numeric(rownames(outsumm))
 outsumm <- melt(outsumm, id.vars = "InitialNumber")
-outsumm$more80 <- ifelse(outsumm$value>=80,1,0)
+outsumm$more80 <- ifelse(outsumm$value>=7,1,0)
 ggplot(outsumm, aes(InitialNumber, variable, fill=more80)) + geom_tile() +
@@
 https://doi.org/10.1371/journal.pone.0041010]), etc., will value having more continuous sampling (although the online DLMs method in Taranu et al 2018 allows for non-continuous sampling).
+
+If we know a priori the shape of the relationship, a non-random sampling is preferred, e.g., <a href="https://onlinelibrary.wiley.com/doi/full/10.1002/env.2552"> VB growth function </a>.

### 2019-10-25 R/temporal-sampling.Rmd
sincrease repetition size

@@
 ```{r, cache = T}
-output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = c(2:10))
+output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = c(2:10), each = 100, type="mean")
 names(output)
@@
 ```
-<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`
+<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample.")`
 
@@
 outsumm <- melt(outsumm, id.vars = "InitialNumber")
-outsumm$more80 <- ifelse(outsumm$value>=7,1,0)
+outsumm$more80 <- ifelse(outsumm$value>=80,1,0)
+
 ggplot(outsumm, aes(InitialNumber, variable, fill=more80)) + geom_tile() +
@@
 ```
-<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`
+<br>`r fig_cap("how many random samples to get the break on average", "Dark blue: the random number of samples analyzed did not allow detection of the break")`
 

### 2019-10-25 R/uncertainty_tolerance.R

@@ -18,16 +18,19 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
   out_list <- list()
   
   # Get the "TRUE" changepoint in mean
-  true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
-  true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
-  if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
-  if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.")
-  
+  if("mean" %in% type) {
+    true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
+    true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
+    if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
+    if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.")
+  }  
   # Get the "TRUE" changepoint in variance
-  true.cptvar.AMOC.y <- cpt.var(y, method="AMOC") 
-  true.cptvar.AMOC.y <- as.numeric(x[cpts(true.cptvar.AMOC.y)])
-  if(length(true.cptvar.AMOC.y)==0) true.cptvar.AMOC.y <- 0
-  if (true.cptvar.AMOC.y == 0) cat("No changepoint in variance was detected on the full y serie -- cannot compute the impact of lower sampling.")
+  if("sd" %in% type) {
+    true.cptvar.AMOC.y <- cpt.var(y, method="AMOC") 
+    true.cptvar.AMOC.y <- as.numeric(x[cpts(true.cptvar.AMOC.y)])
+    if(length(true.cptvar.AMOC.y)==0) true.cptvar.AMOC.y <- 0
+    if (true.cptvar.AMOC.y == 0) cat("No changepoint in variance was detected on the full y serie -- cannot compute the impact of lower sampling.")
+  }  
   
   # I - Regular sampling ####
   if(regular_sampling) {

### 2019-10-25 R/temporal-sampling.Rmd
knit rmd

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(include = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_knit$set(root.dir = dirname(getwd()))
+knitr::opts_chunk$set(warning = FALSE, message = FALSE)
 
@@
 library(reshape)
+
 ```
@@
 
-```{r strat plot zurich selected sp, include=T, echo=F}
+```{r strat plot zurich selected sp, echo=FALSE}
 if(is.zh){
@@
 
-```{r}
-p1 <- ggplot(output, aes(Number_samples, PC1_var)) + geom_point() + stat_smooth()
-p2 <- ggplot(output, aes(Number_samples, PC2_var)) + geom_point() + stat_smooth()
-grid.arrange(p1,p2, nrow=1)
-```
-<br>`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`
-
 # Types of sampling
@@
 
-```{r, cache = T}
+```{r echo=FALSE, message=FALSE, warning=FALSE, cache=T}
 output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = c(2:10), each = 100, type="mean")
@@
 
-```{r summary }
+```{r summary, echo=FALSE, message=FALSE, warning=FALSE}
 names(output$random_cptmean)
@@
 <br>`r fig_cap("how many random samples to get the break on average", "Dark blue: the random number of samples analyzed did not allow detection of the break")`
+
+# Do the same but change the location of the changepoint
+
+```{r message=FALSE, warning=FALSE, cache=T, include=FALSE}
+shift_vector <- sample(2:99, size = 50, replace = F)
+
+for (i in shift_vector) {
+  # Set simulation parameters
+mu = 0
+sigma = 0.1
+phi = 1
+shift_time = i
+max_time = 100
+
+# Set up vectors
+ar1 <- vector(mode = 'numeric',100)
+ar1[1] <- 10
+
+# Build time series
+for (t in 2:max_time){
+  if(t == shift_time){ar1[t-1]=-10}
+  ar1[t] <- ar1[t-1] + phi*rnorm(1,mean=mu,sd = sigma)
+
+}
+
+temporary <- ut(y=exp(ar1), sample_size = c(10:50), threshold_error = c(1:10), each = 1, type="mean")
+
+if(i==shift_vector[1]) {
+  out_regular_cptmean     <- temporary$regular_cptmean
+  out_random_cptmean      <- temporary$random_cptmean
+  out_convergence_cptmean <-temporary$convergence_cptmean
+} else {
+  out_regular_cptmean     <- rbind(out_regular_cptmean, temporary$regular_cptmean)
+  out_random_cptmean      <- rbind(out_random_cptmean, temporary$random_cptmean)
+  out_convergence_cptmean <- rbind(out_convergence_cptmean, temporary$convergence_cptmean)
+}
+}
+
+dim(out_regular_cptmean)
+dim(out_random_cptmean)
+dim(out_convergence_cptmean)
+
+outsumm <- as.data.frame(with(out_random_cptmean, tapply(same_cptmean,list("Initial_number_samples"=sample_size,"threshold_yr"=threshold_error), sum)))
+outsumm$InitialNumber <- as.numeric(rownames(outsumm))
+outsumm <- melt(outsumm, id.vars = "InitialNumber")
+outsumm$more80 <- ifelse(outsumm$value>=50*.8,1,0)
+
+```
+
+```{r}
+ggplot(outsumm, aes(InitialNumber, variable, fill=more80)) + geom_tile() +
+  xlab("Sample analyzed") + ylab("Tolerance for error (year)")
+
+# Something is wrong here, it is sometimes adding >900 samples.I think it has to do with the while loop in the ut() function.
+ggplot(out_convergence_cptmean, aes(sample_size,number_sample_added)) + geom_point(alpha=.3) + ylim(0,25)
+
+boxplot(out_convergence_cptmean$number_sample_added ~ out_convergence_cptmean$sample_size, ylim=c(0,25))
+```
 

### 2019-10-31 R/temporal-sampling.Rmd
work on 10/31

@@
 ---
-
-**Note for next time: write function to compute the tolerance to error against sample size**
 
@@
   mu = 0
-sigma = 0.1
+  sigma = 0.5
   phi = 1
@@
 
-temporary <- ut(y=exp(ar1), sample_size = c(10:50), threshold_error = c(1:10), each = 1, type="mean")
+  temporary <- ut(y=exp(ar1), sample_size = c(10:50), threshold_error = c(1:4), each = 1, type="mean")
 
@@
 # Something is wrong here, it is sometimes adding >900 samples.I think it has to do with the while loop in the ut() function.
-ggplot(out_convergence_cptmean, aes(sample_size,number_sample_added)) + geom_point(alpha=.3) + ylim(0,25)
+ggplot(out_convergence_cptmean, aes(sample_size,number_sample_added)) + geom_point(alpha=.15)  + facet_wrap(~threshold_error)
 

### 2019-10-31 R/uncertainty_tolerance.R

@@ -191,6 +191,8 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
       y2 <- y[which_samples]
       x2 <- x[which_samples]
       
+      which_samples_reset <- which_samples
+      
       # 2. Case 1: changepoint in mean
       if("mean" %in% type) {
         if (true.cptmean.AMOC.y != 0) {
@@ -199,15 +201,20 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
           num <- cpts(cptmean.AMOC.y2)
           cptmean.AMOC.y2 <- as.numeric(x2[cpts(cptmean.AMOC.y2)])
           if(length(cptmean.AMOC.y2)==0) cptmean.AMOC.y2 <- 0
+          num_reset <- num
           
           for (j in seq_along(threshold_error)) {
+            which_samples <- which_samples_reset
+            num <- num_reset
+            print(paste0(sample_size[i],"-",threshold_error[j]))
+            
             # 3. Decision
             t1 = sample_size[i] # The while function doesn't stop sometimes, so stopping the function once we reached the max length of the dataset
             while(abs(cptmean.AMOC.y2-true.cptmean.AMOC.y)>threshold_error[j] & t1 < length(y)) {
               t1 = t1 + 1
               # 3.1. Add an extra point
               # random point is a random sample between change point and following sample
-              additional_sample <- sample(c(which_samples[num]+1):c(which_samples[num+1]-1), size = 1)
+              if (length(c(which_samples[num] + 1):c(which_samples[num + 1] - 1))>0) additional_sample <- sample(c(which_samples[num]+1):c(which_samples[num+1]-1), size = 1)
               which_samples <- c(which_samples,additional_sample)
               which_samples <- which_samples[order(which_samples)]
               y2 <- y[which_samples]
@@ -218,17 +225,19 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
               num <- cpts(cptmean.AMOC.y2)
               cptmean.AMOC.y2 <- as.numeric(x2[cpts(cptmean.AMOC.y2)])
               if(length(cptmean.AMOC.y2)==0) cptmean.AMOC.y2 <- 0
-              print(additional_sample)
+              #print(additional_sample)
             }
             
             # 4. Save outputs
-            if (i == 1 & j == 1) 
+            if (i == 1 & j == 1) {
[…]

### 2019-11-06 R/temporal-sampling.Rmd
added intro section and sketch of theoretical results section

@@
 ---
-title: "Effect of temporal sampling"
-author: "Rosalie Bruel"
+title: "Sampling requirements to detect ecosystem change"
+author: "Rosalie Bruel and Easton White"
 date: "03/10/2019"
@@
     df_print: paged
-    fig_caption: true
+    fig_caption: yes
+    number_sections: yes
     toc: yes
     toc_depth: 3
-    toc_float: true
-    number_sections: true
+    toc_float: yes
+  pdf_document:
+    toc: yes
+    toc_depth: '3'
+fontsize: 11pt
 editor_options:
@@
   df_print: paged
-fontsize: 11pt
 ---
+
+# Things to do:
+
+- Easton: build theoretical results figure
+- Sketch out figures for case studies
+- Conceptual diagram 
+- Begin writing - probably a shorter piece
+- Possible journals: Methods Ecology and Evolution, Biological Conservation, Journal of Environmental Management, Environmental Monitoring and Assessment, BioScience,
+
+# Introduction and setup
+
+- Set up ideas about monitoring and cost of monitoring
+- Idea of detecting change over time
+- Historical data and ability to choose how much to sample
+- Here, we develop a set of tools to ... We first investigate these tools using a simulated-based apporach. We then apply these tools to the detection of change points for a set of lakes...
+
+# Methods and results
 
@@
   if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
-  if(Sys.getenv("USER")=="put here your USER") return("replace here with path to data")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
   if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, cache=T}
-output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = c(2:10), each = 100, type="mean")
+output <- ut(y=exp(ar1), sample_size = c(20:50), threshold_error = 5, each = 100, type="mean")
 names(output)
@@
 
+# Theoretical results
+
+We began with a theoretical exploration of the determiners
+
+We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
+The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occured ($\tau$):
+
+$$
+Model here
+$$
+
+We then explored how each of these model components affected our ability to detect a change point.
+We simulate an entire time series, but we also examine the effect of not sampling every year.
+We specifically study how the amount of samples included and the type of sampling affects the detection probability.
+In particular, we examine three types of sampling: random, regular, and iterative.
+
 # Limits

### 2019-11-20 R/temporal-sampling.Rmd
Added theoretical results

@@
 
-We began with a theoretical exploration of the determiners
-
+We began with a theoretical exploration of the determiners of sampling required to detect a changepoint.
 We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
@@
 $$
-Model here
+\mbox{Model here - need pair of equations}
 $$
@@
 In particular, we examine three types of sampling: random, regular, and iterative.
+
+```{r theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F}
+# 4(or 6?) panel plot
+## Example time series
+## Changepoint detection vs number of samples - maybe just stick with regular sampling
+## Number of samples required vs shift size, shift time, variance, autocorrelation
+
+# Need to build function to output the time series
+# Use purrr to try with a bunch of different parameters
+
+source('R/uncertainty_tolerance.R')
+
+build_time_series <- function(mu=0,sigma=5,phi=0.5,shift_size=25,shift_time=50,max_time=100){
+
+  ar1 <- vector(mode = 'numeric',max_time)
+  ar1[1] <- 100
+  for (t in 2:max_time){
+    if(t == shift_time){ar1[t-1]=(ar1[1]-shift_size)}
+    ar1[t] <- ar1[t-1] + phi*rnorm(1,mean=mu,sd = sigma)
+  }
+  return(ar1) 
+}
+
+require(tidyverse)
+
+par(mfrow=c(2,2))
+# Sigma plot
+params <- tibble(sigma=rep(seq(0.1,20,0.1),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(sigma) %>%
+  summarize(prob_correct = sum(detection==1)/length(unique(sigma))) %>%
+  plot(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1)
+
+# phi plot
+params <- tibble(phi=rep(seq(-1,1,0.01),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(phi) %>%
+  summarize(prob_correct = sum(detection==1)/length(unique(phi))) %>%
+  plot(ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1)
+
+# shift_size plot
+params <- tibble(shift_size=rep(seq(1,95,0.5),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(shift_size) %>%
+  summarize(prob_correct = sum(detection==1)/length(unique(shift_size))) %>%
+  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1)
+
+# Shift timing plot
+params <- tibble(shift_time=rep(seq(5,90,1),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(shift_time) %>%
+  summarize(prob_correct = sum(detection==1)/length(shift_time/unique(shift_time))) %>%
+  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1)
+
+#params <- tibble(phi=rep(seq(-1,1,0.01),1),shift_size=rep(30,201))
+#output <- params %>%
+#  pmap(build_time_series)
+#plot(output[[1]],type='l')
+#points(output[[15]],type='l')
+```
 

### 2019-11-21 R/temporal-sampling.Rmd
edits on temporal sampling

@@
 - Here, we develop a set of tools to ... We first investigate these tools using a simulated-based apporach. We then apply these tools to the detection of change points for a set of lakes...
+
+- In paleo studies, if we cannot sample continously, we would sometimes chose to sample at regular intervals (e.g., every 5 cm).
+Would an iterative approach be more effective in detecting major changes?
+- Limitations: right now we are detecting changes on a single vector (in ecological dataset, it would be for example a PC axis).
+We talked about how a user could sample, e.g., 5 samples, and then decide where to add some samples, depending on when the change was detected.
+However, the PC ordination will be done depending on the community structure.
+Are we including a bias by doing that?
 
@@
 zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt")); is.zh=T
-zh <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
+zh2 <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
+zh3 <- cbind(zh, matrix(rep(0,nrow(zh)*(ncol(zh)-1)), nrow(zh)))
+zh3 <- rbind(zh3, cbind(matrix(c(1700:1710), rep(0,10*(ncol(zh)-1))),unlist(as.vector(zh2[1:10,-1]))) )
+zh<-zh3
 ```

### 2019-11-21 R/temporal-sampling.Rmd
added theoretical results

--- 2019-11-20 R/temporal-sampling.Rmd
@@
 # Sigma plot
+sample_size=10
 params <- tibble(sigma=rep(seq(0.1,20,0.1),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(sigma) %>%
+  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
+  plot(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
+
+sample_size=10
+params <- tibble(sigma=rep(seq(0.1,20,0.1),10))
 
@@
   group_by(sigma) %>%
-  summarize(prob_correct = sum(detection==1)/length(unique(sigma))) %>%
-  plot(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1)
+  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
+  points(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col='red')
 
@@
   pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
   mutate(params,detection=.)
@@
   group_by(phi) %>%
-  summarize(prob_correct = sum(detection==1)/length(unique(phi))) %>%
-  plot(ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1)
+  summarize(prob_correct = sum(detection==1)/length(phi/unique(phi))) %>%
+  plot(ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
 
@@
   group_by(shift_size) %>%
-  summarize(prob_correct = sum(detection==1)/length(unique(shift_size))) %>%
-  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1)
+  summarize(prob_correct = sum(detection==1)/length(shift_size/unique(shift_size))) %>%
+  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
 
@@
   summarize(prob_correct = sum(detection==1)/length(shift_time/unique(shift_time))) %>%
-  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1)
+  plot(ylab='Probability correct',xlab='Shift timing',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
+
+#params <- tibble(phi=rep(seq(-1,1,0.01),1),shift_size=rep(30,201))
+#output <- params %>%
+#  pmap(build_time_series)
+#plot(output[[1]],type='l')
+#points(output[[15]],type='l')
+```
+
+In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\\sigma$) and autocorrelation ($\\phi$).
+We also found that the probability of correct changepoint detection increases with larger shift sizes.
+
+The probability of correcting identifying a changepoint, of course, depends strongly on the number of samples.
+
+```{r different_methods, fig.cap='Theoretical results here',message=F,warning=F,echo=F}
+# 4(or 6?) panel plot
+
+par(mfrow=c(1,1))
+# Sigma plot
+sample_size=10
+params <- tibble(sigma=rep(seq(0.1,20,0.1),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(sigma) %>%
+  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
+  plot(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
+
+sample_size=10
+params <- tibble(sigma=rep(seq(0.1,20,0.1),10))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(sigma) %>%
+  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
+  points(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col='red')
+
+# phi plot
+params <- tibble(phi=rep(seq(-1,1,0.01),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(phi) %>%
+  summarize(prob_correct = sum(detection==1)/length(phi/unique(phi))) %>%
+  plot(ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
+
+# shift_size plot
+params <- tibble(shift_size=rep(seq(1,95,0.5),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(shift_size) %>%
+  summarize(prob_correct = sum(detection==1)/length(shift_size/unique(shift_size))) %>%
+  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
+
+# Shift timing plot
+params <- tibble(shift_time=rep(seq(5,90,1),100))
+
+output <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
+  mutate(params,detection=.)
+
+output %>%
+  group_by(shift_time) %>%
+  summarize(prob_correct = sum(detection==1)/length(shift_time/unique(shift_time))) %>%
+  plot(ylab='Probability correct',xlab='Shift timing',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
 

### 2019-11-21 R/temporal-sampling.Rmd
comment the weird creation of a fake data table (keeping it for later)

@@
 zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt")); is.zh=T
-zh2 <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
-zh3 <- cbind(zh, matrix(rep(0,nrow(zh)*(ncol(zh)-1)), nrow(zh)))
-zh3 <- rbind(zh3, cbind(matrix(c(1700:1710), rep(0,10*(ncol(zh)-1))),unlist(as.vector(zh2[1:10,-1]))) )
-zh<-zh3
+# zh2 <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
+# zh3 <- cbind(zh, matrix(rep(0,nrow(zh)*(ncol(zh)-1)), nrow(zh)))
+# zh3 <- rbind(zh3, cbind(matrix(c(1700:1710), rep(0,10*(ncol(zh)-1))),unlist(as.vector(zh2[1:10,-1]))) )
+# zh<-zh3
 ```

### 2020-01-02 R/temporal-sampling.Rmd
Add comments on limitations

@@
 Are we including a bias by doing that?
+==> The script Rosalie initially worked on computed the PCA post first change.
+- Note: if the result is really impacted by the ordination, we could have here an important warning, i.e., the sampling really restrict the message.
+Which would not be new at all, but not demonstrated with data to my knowledge.
+And for people working on temporal ecology, this would be an interesting paper to refer to (we can talk more about it during our work session).
+Other point: what happens if the transition is right at the beginning or end of the record?
 

### 2020-01-03 R/temporal-sampling.Rmd
added a few pieces of text

--- 2019-11-21 R/temporal-sampling.Rmd
@@
 
-build_time_series <- function(mu=0,sigma=5,phi=0.5,shift_size=25,shift_time=50,max_time=100){
+build_time_series <- function(mu=0,sigma=2,phi=0.5,shift_size=25,shift_time=50,max_time=100){
 
@@
   ar1[1] <- 100
+  shift_time = sample(5:95,size = 1)
   for (t in 2:max_time){
@@
 
-sample_size=10
+sample_size=20
 params <- tibble(sigma=rep(seq(0.1,20,0.1),10))
@@
 ```{r different_methods, fig.cap='Theoretical results here',message=F,warning=F,echo=F}
-# 4(or 6?) panel plot
 
-par(mfrow=c(1,1))
-# Sigma plot
-sample_size=10
-params <- tibble(sigma=rep(seq(0.1,20,0.1),100))
+# Need to build a single plot comparing the three methods - look at probability of detection versus sample size... We would do this for a set of default parameters
 
-output <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
-  mutate(params,detection=.)
-
-output %>%
-  group_by(sigma) %>%
-  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
-  plot(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
-
-sample_size=10
-params <- tibble(sigma=rep(seq(0.1,20,0.1),10))
-
-output <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = 20, threshold_error = 5, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
-  mutate(params,detection=.)
-
-output %>%
-  group_by(sigma) %>%
-  summarize(prob_correct = sum(detection==1)/length(sigma/unique(sigma))) %>%
-  points(ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col='red')
-
-# phi plot
-params <- tibble(phi=rep(seq(-1,1,0.01),100))
-
-output <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
-  mutate(params,detection=.)
-
-output %>%
-  group_by(phi) %>%
-  summarize(prob_correct = sum(detection==1)/length(phi/unique(phi))) %>%
-  plot(ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
-
-# shift_size plot
-params <- tibble(shift_size=rep(seq(1,95,0.5),100))
-
-output <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
-  mutate(params,detection=.)
-
-output %>%
-  group_by(shift_size) %>%
-  summarize(prob_correct = sum(detection==1)/length(shift_size/unique(shift_size))) %>%
-  plot(ylab='Probability correct',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
-
-# Shift timing plot
-params <- tibble(shift_time=rep(seq(5,90,1),100))
-
-output <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) ut(y=df, sample_size = sample_size, threshold_error = 3, each = 1, type="mean",random_sampling = F,convergence_sampling = F,regular_sampling = T)$regular_cptmean$same_cpt) %>%
-  mutate(params,detection=.)
-
-output %>%
-  group_by(shift_time) %>%
-  summarize(prob_correct = sum(detection==1)/length(shift_time/unique(shift_time))) %>%
-  plot(ylab='Probability correct',xlab='Shift timing',pch=16,cex.lab=1.3,las=1,ylim=c(0,1))
-
-#params <- tibble(phi=rep(seq(-1,1,0.01),1),shift_size=rep(30,201))
-#output <- params %>%
-#  pmap(build_time_series)
-#plot(output[[1]],type='l')
-#points(output[[15]],type='l')
 ```

### 2020-01-03 R/uncertainty_tolerance.R

@@ -3,7 +3,7 @@
 require(changepoint)
 
 #uncertainty tolerance function
-ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
+ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),simulated_data = F,simulated_true_changepoint=NULL,
                threshold_error = 1, each = 10,
                random_sampling=T,
                regular_sampling=T,
@@ -19,10 +19,14 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
   
   # Get the "TRUE" changepoint in mean
   if("mean" %in% type) {
-    true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
-    true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
-    if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
-    if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.")
+    if(simulated_data){
+      true.cptmean.AMOC.y <- simulated_true_changepoint
+    }else{
+      true.cptmean.AMOC.y <- cpt.mean(y, method="AMOC") 
+      true.cptmean.AMOC.y <- as.numeric(x[cpts(true.cptmean.AMOC.y)])
+      if(length(true.cptmean.AMOC.y)==0) true.cptmean.AMOC.y <- 0
+      if (true.cptmean.AMOC.y == 0) cat("No changepoint in mean was detected on the full y serie -- cannot compute the impact of lower sampling.")
+    }
   }  
   # Get the "TRUE" changepoint in variance
   if("sd" %in% type) {
@@ -191,8 +195,6 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
       y2 <- y[which_samples]
       x2 <- x[which_samples]
       
-      which_samples_reset <- which_samples
-      
       # 2. Case 1: changepoint in mean
       if("mean" %in% type) {
         if (true.cptmean.AMOC.y != 0) {
@@ -201,20 +203,15 @@ ut <- function(x = NULL, y, sample_size = NULL, type=c("mean","sd"),
           num <- cpts(cptmean.AMOC.y2)
           cptmean.AMOC.y2 <- as.numeric(x2[cpts(cptmean.AMOC.y2)])
[…]

### 2020-01-03 R/sampling_regular.R
function for regular sampling

@@ -1,0 +1,28 @@
+#' Regular sampling of a 1D vector
+#'
+#' Function to sample regularly a vector
+#'
+#' @export
+#' @param x Initial vector
+#' @param n Number of samples analyzed
+#' @keywords sampling
+#' @examples 
+#'
+
+sample_regular <- function(x, n)
+  .sample_regular
+.sample_regular <- function(x, n) {
+  require(ecp)
+  
+  if(!is.vector(x)) stop("x is required and must be a vector")
+  if(is.null(n)) n <- length(x)
+  
+  # Sample n samples from initial x
+  which_samples <- seq(1,length(x), length.out = n)
+  which_samples <- round(which_samples)
+  which_samples <- which_samples[order(which_samples)]
+  
+  x2 <- matrix(c(which_samples,x[which_samples]), ncol=2)
+  
+  return(e.divisive(x2)$estimates[-1])
+} 

### 2020-01-03 R/sampling_random.R
function for random sampling

@@ -1,0 +1,27 @@
+#' Random sampling of a 1D vector
+#'
+#' Function to sample randomly a vector
+#'
+#' @export
+#' @param x Initial vector
+#' @param n Number of samples analyzed
+#' @keywords sampling
+#' @examples 
+#'
+
+sample_random <- function(x, n)
+  .sample_regular
+.sample_regular <- function(x, n) {
+  require(ecp)
+  
+  if(!is.vector(x)) stop("x is required and must be a vector")
+  if(is.null(n)) n <- length(x)
+  
+  # Sample n samples from initial x
+  which_samples <- sample(x, n)
+  which_samples <- which_samples[order(which_samples)]
+  
+  x2 <- matrix(c(which_samples,x[which_samples]), ncol=2)
+  
+  return(e.divisive(x2)$estimates[-1])
+} 

### 2020-01-03 R/sampling_random.R
tiny edit on random sampling

@@ -18,7 +18,7 @@ sample_random <- function(x, n)
   if(is.null(n)) n <- length(x)
   
   # Sample n samples from initial x
-  which_samples <- sample(x, n)
+  which_samples <- sample(1:length(x), n)
   which_samples <- which_samples[order(which_samples)]
   
   x2 <- matrix(c(which_samples,x[which_samples]), ncol=2)

### 2020-01-03 R/sampling_random.R
edit function random and regular sampling

@@ -10,7 +10,7 @@
 #'
 
 sample_random <- function(x, n)
-  .sample_regular
+  .sample_regular(x, n)
 .sample_regular <- function(x, n) {
   require(ecp)
   
@@ -21,7 +21,18 @@ sample_random <- function(x, n)
   which_samples <- sample(1:length(x), n)
   which_samples <- which_samples[order(which_samples)]
   
-  x2 <- matrix(c(which_samples,x[which_samples]), ncol=2)
+  x2 <- matrix(x[which_samples], ncol=1)
   
-  return(e.divisive(x2)$estimates[-1])
+  changepoints <- e.divisive(x2, min.size = 2)
+  
+  # Print the main changepoint
+  print(which_samples[changepoints$order.found[3]])
+  
+  # More elaborated output
+  out_list <- list()
+  out_list$changepoint <- which_samples[changepoints$order.found[3]]
+  out_list$changepoint_all <- changepoints
+  out_list$matrix <- data.frame("index"=which_samples,"x"=x2)
+  
+  invisible(out_list)
 } 

### 2020-01-03 R/sampling_regular.R

@@ -10,7 +10,7 @@
 #'
 
 sample_regular <- function(x, n)
-  .sample_regular
+  .sample_regular(x, n)
 .sample_regular <- function(x, n) {
   require(ecp)
   
@@ -22,7 +22,21 @@ sample_regular <- function(x, n)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   
-  x2 <- matrix(c(which_samples,x[which_samples]), ncol=2)
+  x2 <- matrix(x[which_samples], ncol=1)
   
-  return(e.divisive(x2)$estimates[-1])
+  changepoints <- e.divisive(x2, min.size = 2)
+  
+  # Print the main changepoint
+  print(which_samples[changepoints$order.found[3]])
+  
+  # More elaborated output
+  out_list <- list()
+      out_list$changepoint <- which_samples[changepoints$order.found[3]]
+  out_list$changepoint_all <- changepoints
+           out_list$matrix <- data.frame("index"=which_samples,"x"=x2)
+  
+  invisible(out_list)
+           
 } 
+
+

### 2020-01-03 R/sampling_random.R
add argument messages to turn on and off the message printing

@@ -9,9 +9,9 @@
 #' @examples 
 #'
 
-sample_random <- function(x, n)
-  .sample_regular(x, n)
-.sample_regular <- function(x, n) {
+sample_random <- function(x, n, messages = T)
+  .sample_regular(x, n, messages)
+.sample_regular <- function(x, n, messages) {
   require(ecp)
   
   if(!is.vector(x)) stop("x is required and must be a vector")
@@ -26,7 +26,7 @@ sample_random <- function(x, n)
   changepoints <- e.divisive(x2, min.size = 2)
   
   # Print the main changepoint
-  print(which_samples[changepoints$order.found[3]])
+  if(messages) print(which_samples[changepoints$order.found[3]])
   
   # More elaborated output
   out_list <- list()

### 2020-01-03 R/sampling_regular.R

@@ -9,9 +9,9 @@
 #' @examples 
 #'
 
-sample_regular <- function(x, n)
-  .sample_regular(x, n)
-.sample_regular <- function(x, n) {
+sample_regular <- function(x, n, messages = T)
+  .sample_regular(x, n, messages)
+.sample_regular <- function(x, n, messages) {
   require(ecp)
   
   if(!is.vector(x)) stop("x is required and must be a vector")
@@ -27,7 +27,7 @@ sample_regular <- function(x, n)
   changepoints <- e.divisive(x2, min.size = 2)
   
   # Print the main changepoint
-  print(which_samples[changepoints$order.found[3]])
+  if(messages) print(which_samples[changepoints$order.found[3]])
   
   # More elaborated output
   out_list <- list()

### 2020-01-03 R/sampling_random.R
use na.locf from zoo package to fill-in NA gaps

@@ -13,6 +13,7 @@ sample_random <- function(x, n, messages = T)
   .sample_regular(x, n, messages)
 .sample_regular <- function(x, n, messages) {
   require(ecp)
+  require(zoo)
   
   if(!is.vector(x)) stop("x is required and must be a vector")
   if(is.null(n)) n <- length(x)
@@ -21,7 +22,11 @@ sample_random <- function(x, n, messages = T)
   which_samples <- sample(1:length(x), n)
   which_samples <- which_samples[order(which_samples)]
   
-  x2 <- matrix(x[which_samples], ncol=1)
+  # transform into matrix to run e.divisive
+  x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
+  x2[which_samples,2] <- x[which_samples]
+  x2[,2] <- na.locf(x2[,2])
+  x2 <- matrix(x2[,2])
   
   changepoints <- e.divisive(x2, min.size = 2)
   

### 2020-01-03 R/sampling_regular.R

@@ -13,6 +13,7 @@ sample_regular <- function(x, n, messages = T)
   .sample_regular(x, n, messages)
 .sample_regular <- function(x, n, messages) {
   require(ecp)
+  require(zoo)
   
   if(!is.vector(x)) stop("x is required and must be a vector")
   if(is.null(n)) n <- length(x)
@@ -22,7 +23,11 @@ sample_regular <- function(x, n, messages = T)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   
-  x2 <- matrix(x[which_samples], ncol=1)
+  # transform into matrix to run e.divisive
+  x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
+  x2[which_samples,2] <- x[which_samples]
+  x2[,2] <- na.locf(x2[,2])
+  x2 <- matrix(x2[,2])
   
   changepoints <- e.divisive(x2, min.size = 2)
   

### 2020-01-03 R/sampling_iterative.R
Changepoint method using na.locf

@@ -1,0 +1,83 @@
+#' Iterative sampling of a 1D vector
+#'
+#' Function to sample a vector iteratively
+#'
+#' @export
+#' @param x Initial vector
+#' @param n Number of samples analyzed
+#' @keywords sampling
+#' @examples 
+#'
+
+sample_iterative <- function(x, n, messages = T)
+  .sample_iterative(x, n, messages)
+.sample_iterative <- function(x, n, messages) {
+  require(ecp)
+  require(zoo)
+  
+  if(!is.vector(x)) stop("x is required and must be a vector")
+  if(is.null(n)) n <- length(x)
+  
+  # Sample initial number of samples
+  n2 = 5
+  which_samples <- seq(1,length(x),length.out = n2)
+  which_samples <- round(which_samples)
+  which_samples <- which_samples[order(which_samples)]
+  
+  
+  # transform into matrix to run e.divisive
+  x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
+  x2[which_samples,2] <- x[which_samples]
+  x2[,2] <- na.locf(x2[,2])
+  x2 <- matrix(x2[,2])
+  
+  temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3]
+  
+  if(is.na(temp_chgpt)) {
+    which_samples <- seq(1,2,length(x),length.out = n2*2-1)
+    which_samples <- round(which_samples)
+    which_samples <- which_samples[order(which_samples)]
+    
+    # transform into matrix to run e.divisive
+    x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
+    x2[which_samples,2] <- x[which_samples]
+    x2[,2] <- na.locf(x2[,2])
+    x2 <- matrix(x2[,2])
+    
+    temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3]
+  }
+  
+  # Now, start adding samples
+  for(i in 1:(n-n2)) {
+    # Here we only want to add sample if we haven't already found the closest changepoint
+    if(!round((which_samples[which(which_samples==temp_chgpt)-1]+which_samples[which(which_samples==temp_chgpt)])/2) %in% which_samples) {
+      which_samples <- c(which_samples, round((which_samples[which(which_samples==temp_chgpt)-1]+which_samples[which(which_samples==temp_chgpt)])/2))
+      which_samples <- which_samples[order(which_samples)]
[…]

### 2020-01-03 R/sampling_random.R

@@ -31,13 +31,14 @@ sample_random <- function(x, n, messages = T)
   changepoints <- e.divisive(x2, min.size = 2)
   
   # Print the main changepoint
-  if(messages) print(which_samples[changepoints$order.found[3]])
+  if(messages) print(changepoints$order.found[3])
   
   # More elaborated output
-  out_list <- list()
-  out_list$changepoint <- which_samples[changepoints$order.found[3]]
+                  out_list <- list()
+      out_list$changepoint <- changepoints$order.found[3]
+          out_list$final_n <- length(which_samples)
   out_list$changepoint_all <- changepoints
-  out_list$matrix <- data.frame("index"=which_samples,"x"=x2)
+           out_list$matrix <- data.frame("index"=which_samples,"x"=x[which_samples])
   
   invisible(out_list)
 } 

### 2020-01-03 R/sampling_regular.R

@@ -32,13 +32,14 @@ sample_regular <- function(x, n, messages = T)
   changepoints <- e.divisive(x2, min.size = 2)
   
   # Print the main changepoint
-  if(messages) print(which_samples[changepoints$order.found[3]])
+  if(messages) print(changepoints$order.found[3])
   
   # More elaborated output
-  out_list <- list()
-      out_list$changepoint <- which_samples[changepoints$order.found[3]]
+                  out_list <- list()
+      out_list$changepoint <- changepoints$order.found[3]
+          out_list$final_n <- length(which_samples)
   out_list$changepoint_all <- changepoints
-           out_list$matrix <- data.frame("index"=which_samples,"x"=x2)
+           out_list$matrix <- data.frame("index"=which_samples,"x"=x[which_samples])
   
   invisible(out_list)
            

### 2020-01-03 R/build_time_series.R
command to run ar-1 process time series

@@ -1,0 +1,15 @@
+
+
+# Build ar-1 time series with various model inputs
+
+build_time_series <- function(mu=0,sigma=2,phi=0.5,shift_size=0.25,shift_time=50,max_time=100){
+  
+  ar1 <- vector(mode = 'numeric',max_time)
+  ar1[1] <- 100
+  #shift_time = sample(5:(max_time-5),size = 1)
+  for (t in 2:max_time){
+    if(t == shift_time){ar1[t-1]=(1-shift_size)*ar1[1]}
+    ar1[t] <- ar1[t-1] + phi*rnorm(1,mean=mu,sd = sigma)
+  }
+  return(ar1) 
+}

### 2020-01-03 R/sampling_random.R
changing the initial sample selection so the first and last sample are selected. at least the first one is needed for na.locf to work.

@@ -19,7 +19,7 @@ sample_random <- function(x, n, messages = T)
   if(is.null(n)) n <- length(x)
   
   # Sample n samples from initial x
-  which_samples <- sample(1:length(x), n)
+  which_samples <- c(1,length(x),sample(1:length(x), n-2))
   which_samples <- which_samples[order(which_samples)]
   
   # transform into matrix to run e.divisive

### 2020-01-03 R/sampling_random.R
problem when calling function

@@ -10,8 +10,8 @@
 #'
 
 sample_random <- function(x, n, messages = T)
-  .sample_regular(x, n, messages)
-.sample_regular <- function(x, n, messages) {
+  .sample_random(x, n, messages)
+.sample_random <- function(x, n, messages) {
   require(ecp)
   require(zoo)
   

### 2020-01-04 R/temporal-sampling.Rmd
added new theoretical results section

@@
 We then explored how each of these model components affected our ability to detect a change point.
-We simulate an entire time series, but we also examine the effect of not sampling every year.
+We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
 We specifically study how the amount of samples included and the type of sampling affects the detection probability.
@@
 
-```{r theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F}
+```{r old_theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F,eval=F}
 # 4(or 6?) panel plot
@@
 
+```{r theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F,eval=F}
+
+source('R/sampling_regular.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+
+require(tidyverse)
+
+par(mfrow=c(2,2))
+# Sigma plot
+sample_size=10
+
+sigma15 <- param_sensitivity('sigma',15)
+
+plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+```
+
 In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\\sigma$) and autocorrelation ($\\phi$).

### 2020-01-04 R/manuscript_draft.Rmd
moved some text to a rough draft of the manuscript

---
title: "Sampling requirements to detect ecosystem change"
author: "Rosalie Bruel and Easton White"
date: "03/10/2019"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
fontsize: 11pt
editor_options:
  chunk_output_type: console
  df_print: paged
---

# Things to do:

- writing intro
- writing methods/results pair of section
- writing dicussion
- setting up bibliography system
- Possible journals: Methods Ecology and Evolution, Biological Conservation, *Journal of Environmental Management*, *Environmental Monitoring and Assessment*, BioScience

# Main messages of manuscript

- We can choose the number of samples and effort we use...
- Number of samples required depends on question asked and the time series parameters
- Case study of paleo data shows the utility of iterative sampling in detecting change points

# Introduction and setup

- Set up ideas about monitoring and cost of monitoring

Environmental monitoring is one of the core components to modern research and management (citations). However, many monitoring programs are expensive and time-consuming (citations). Therefore, an optimal...

- Idea of detecting change over time

One of the primary goals of monitoring programs is to detect long-term ecological change.

- Historical data and ability to choose how much to sample

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occured. For example, paleoecological lake cores can be extracted and ecological communities can be built from this... Similarly, samples for eDNA or videographical approaches can record time points of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleo core be analyzed at every centimeter? Should the video be assessed at every minute? As long as analyzing samples is expensive, these tradeoffs will remain.

- Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added. We first investigate these tools using a simulated-based apporach. We then apply these tools to the detection of change points for a set of lakes...

# Theoretical results

- explain the modeling approach
- show the model
- explain the analysis we will do
- explain the theoretical model results with a figure
- explain effect of different sampling strategies in this context

# Lake case study

# Discussion

# References

### 2020-01-04 R/temporal-sampling.Rmd
moved some text to a rough draft of the manuscript

@@
 ---
-
-# Things to do:
-
-- Easton: build theoretical results figure
-- Sketch out figures for case studies
-- Conceptual diagram 
-- Begin writing - probably a shorter piece
-- Possible journals: Methods Ecology and Evolution, Biological Conservation, Journal of Environmental Management, Environmental Monitoring and Assessment, BioScience,
-
-# Introduction and setup
-
-- Set up ideas about monitoring and cost of monitoring
-- Idea of detecting change over time
-- Historical data and ability to choose how much to sample
-- Here, we develop a set of tools to ... We first investigate these tools using a simulated-based apporach. We then apply these tools to the detection of change points for a set of lakes...
 
@@
 
-```{r theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F,eval=F}
+```{r theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F,eval=T}
 
@@
 # Sigma plot
-sample_size=10
+sigma15 <- param_sensitivity('sigma',15)
+plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
 
-sigma15 <- param_sensitivity('sigma',15)
+sigma20 <- param_sensitivity('sigma',20)
+points(sigma20,ylab='Probability correct',xlab='sigma',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
 
-plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+# phi plot
+phi15 <- param_sensitivity('phi',15)
+plot(phi15,ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+phi20 <- param_sensitivity('phi',20)
+points(phi20,ylab='Probability correct',xlab='phi',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
 

### 2020-01-04 R/calc_min_time.R
draft code to calculate minimum time required

@@ -1,0 +1,35 @@
+
+
+# Find min_time required
+# Increase sample size until 0.8 power is achieved...
+
+sample_size = 9
+power = 0
+result=NULL
+
+for (sigma_value in seq(0.5,6,0.3)){
+power=0
+  
+while(power < 0.8 & sample_size<50){
+  
+  sample_size = sample_size + 1
+
+  params <- tibble(sigma=rep(sigma_value,100))
+  
+  output <- params %>%
+    pmap(build_time_series) %>%
+    map_dbl(function(df) sample_regular(df,sample_size)$changepoint) %>%
+    mutate(params,detection=.)
+
+output$detection = ifelse(abs(output$detection-50)<5,1,0)
+output$detection[is.na(output$detection)]=FALSE
+
+power = sum(output$detection)/nrow(output)
+
+
+}
+
+result = rbind(result,c(sigma_value,sample_size))  
+#print(paste(sigma_value,':',sample_size))
+
+}

### 2020-01-04 R/manuscript_draft.Rmd
minor text changes

@@
 
-One of the primary goals of monitoring programs is to detect long-term ecological change.
+One of the primary goals of monitoring programs is to detect long-term ecological change...
 
@@
 
-- Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
+Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
 We first investigate these tools using a simulated-based apporach.

### 2020-01-04 R/manuscript_draft.Rmd
moved theoretical results writeup

@@
 
+We began with a theoretical exploration of the determiners of sampling required to detect a changepoint.
+We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
+The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occured ($\tau$):
+
+$$
+\mbox{Model here - need pair of equations}
+$$
+
+We then explored how each of these model components affected our ability to detect a change point.
+We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
+We specifically study how the amount of samples included and the type of sampling affects the detection probability.
+
+In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\\sigma$) and autocorrelation ($\\phi$).
+We also found that the probability of correct changepoint detection increases with larger shift sizes.
+
+The probability of correcting identifying a changepoint, of course, depends strongly on the number of samples.
+
 # Lake case study

### 2020-01-04 R/temporal-sampling.Rmd
moved theoretical results writeup

@@
 
-# Theoretical results
-
-We began with a theoretical exploration of the determiners of sampling required to detect a changepoint.
-We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
-The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occured ($\tau$):
-
-$$
-\mbox{Model here - need pair of equations}
-$$
-
-We then explored how each of these model components affected our ability to detect a change point.
-We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
-We specifically study how the amount of samples included and the type of sampling affects the detection probability.
-In particular, we examine three types of sampling: random, regular, and iterative.
-
 ```{r old_theoretical_results, fig.cap='Theoretical results here',message=F,warning=F,echo=F,eval=F}
@@
 
-In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\\sigma$) and autocorrelation ($\\phi$).
-We also found that the probability of correct changepoint detection increases with larger shift sizes.
-
-The probability of correcting identifying a changepoint, of course, depends strongly on the number of samples.
-
 ```{r different_methods, fig.cap='Theoretical results here',message=F,warning=F,echo=F}

### 2020-01-04 R/sampling_iterative.R
edited the 3 functions to allow input of a dataframe

@@ -5,42 +5,70 @@
 #' @export
 #' @param x Initial vector
 #' @param n Number of samples analyzed
+#' @param n2 Initial number of sample, regularly spaced. Default to 5. Will be changed to (n2*2-1) if no changepoints were detected for initial value of n2.
+#' @param messages Print changepoint or not
+#' @param input_vector Default input vector = T. If turned to F, allow a data frame
+#' @param xcol if input_vector = F, index of columns not to use for ordination.
+#' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_iterative <- function(x, n, messages = T)
-  .sample_iterative(x, n, messages)
-.sample_iterative <- function(x, n, messages) {
+sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
+  .sample_iterative(x, n, messages, input_vector, xcol, DCA_axis)
+.sample_iterative <- function(x, n, messages, input_vector, xcol, DCA_axis) {
   require(ecp)
   require(zoo)
+  require(vegan)
+  do_dca <- function(dataset, xcol= NA) {
+    data_temp <- dataset[,-c(xcol)]
+    data_temp <- data_temp[, colSums(data_temp != 0) > 0]
+    data_temp <- decorana(data_temp)
+    return(as.data.frame(data_temp$rproj))
+  }
   
-  if(!is.vector(x)) stop("x is required and must be a vector")
-  if(is.null(n)) n <- length(x)
+  x_stable <- x
+  
+  if(!is.vector(x)&input_vector) stop("x is required and must be a vector, unless input_vector is turned to FALSE. If so, x must be a dataframe.")
+  if(input_vector) lastpoint = length(x) else lastpoint = nrow(x)
+  if(is.null(n)) n <- lastpoint
   
   # Sample initial number of samples
-  n2 = 5
-  which_samples <- seq(1,length(x),length.out = n2)
+  which_samples <- seq(1,lastpoint,length.out = n2)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   
+  # Do DCA to get only one vector if input is a data frame
+  if(!input_vector) {
+    x <- do_dca(x[which_samples,], xcol)
+    x <- x[,DCA_axis]
+  }
   
[…]

### 2020-01-04 R/sampling_random.R

@@ -5,26 +5,44 @@
 #' @export
 #' @param x Initial vector
 #' @param n Number of samples analyzed
+#' @param messages Print changepoint or not
+#' @param input_vector Default input vector = T. If turned to F, allow a data frame
+#' @param xcol if input_vector = F, index of columns not to use for ordination.
+#' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_random <- function(x, n, messages = T)
-  .sample_random(x, n, messages)
-.sample_random <- function(x, n, messages) {
+sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
+  .sample_random(x, n, messages, input_vector, xcol, DCA_axis)
+.sample_random <- function(x, n, messages, input_vector, xcol, DCA_axis) {
   require(ecp)
   require(zoo)
+  require(vegan)
+  do_dca <- function(dataset, xcol= NA) {
+    data_temp <- dataset[,-c(xcol)]
+    data_temp <- data_temp[, colSums(data_temp != 0) > 0]
+    data_temp <- decorana(data_temp)
+    return(as.data.frame(data_temp$rproj))
+  }
   
-  if(!is.vector(x)) stop("x is required and must be a vector")
-  if(is.null(n)) n <- length(x)
+  if(!is.vector(x)&input_vector) stop("x is required and must be a vector, unless input_vector is turned to FALSE. If so, x must be a dataframe.")
+  if(input_vector) lastpoint = length(x) else lastpoint = nrow(x)
+  if(is.null(n)) n <- lastpoint
   
   # Sample n samples from initial x
-  which_samples <- c(1,length(x),sample(1:length(x), n-2))
+  which_samples <- c(1,lastpoint,sample(1:lastpoint, n-2))
   which_samples <- which_samples[order(which_samples)]
   
+  # Do DCA to get only one vector if input is a data frame
+  if(!input_vector) {
+    x <- do_dca(x[which_samples,], xcol)
+    x <- x[,DCA_axis]
+  }
+  
   # transform into matrix to run e.divisive
-  x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
-  x2[which_samples,2] <- x[which_samples]
+  x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
[…]

### 2020-01-04 R/sampling_regular.R

@@ -5,27 +5,46 @@
 #' @export
 #' @param x Initial vector
 #' @param n Number of samples analyzed
+#' @param messages Print changepoint or not
+#' @param input_vector Default input vector = T. If turned to F, allow a data frame
+#' @param xcol if input_vector = F, index of columns not to use for ordination.
+#' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_regular <- function(x, n, messages = T)
-  .sample_regular(x, n, messages)
-.sample_regular <- function(x, n, messages) {
+sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
+  .sample_regular(x, n, messages, input_vector, xcol, DCA_axis)
+.sample_regular <- function(x, n, messages, input_vector, xcol, DCA_axis) {
   require(ecp)
   require(zoo)
+  require(vegan)
+  do_dca <- function(dataset, xcol= NA) {
+    data_temp <- dataset[,-c(xcol)]
+    data_temp <- data_temp[, colSums(data_temp != 0) > 0]
+    data_temp <- decorana(data_temp)
+    return(as.data.frame(data_temp$rproj))
+    #scores.species<-as.data.frame(data_temp$cproj)
+  }
   
-  if(!is.vector(x)) stop("x is required and must be a vector")
-  if(is.null(n)) n <- length(x)
+  if(!is.vector(x)&input_vector) stop("x is required and must be a vector, unless input_vector is turned to FALSE. If so, x must be a dataframe.")
+  if(input_vector) lastpoint = length(x) else lastpoint = nrow(x)
+  if(is.null(n)) n <- lastpoint
   
   # Sample n samples from initial x
-  which_samples <- seq(1,length(x), length.out = n)
+  which_samples <- seq(1,lastpoint, length.out = n)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   
+  # Do DCA to get only one vector if input is a data frame
+  if(!input_vector) {
+    x <- do_dca(x[which_samples,], xcol)
+    x <- x[,DCA_axis]
+  }
+  
   # transform into matrix to run e.divisive
-  x2 <- matrix(c(1:length(x),rep(NA,length(x))), ncol=2)
-  x2[which_samples,2] <- x[which_samples]
+  x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
[…]

### 2020-01-04 R/Fig_conceptual.R
edit figure conceptual

@@ -1,0 +1,126 @@
+library(wesanderson)
+mypal <- wes_palette("Chevalier1")
+
+ar1
+ar1 <- build_time_series()
+mydata <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+
+cex_titles <- .8
+
+# Require an output from any of the 3 functions sampling_regular, sampling_random, sampling_iterative
+pol_break <- function(cpt, col=1, y1=-.2, y2=1.2) {
+  if(col==1) mycol=mypal[1]
+  if(col==2) mycol=mypal[2]
+  x1   <- cpt$changepoint
+  x2   <- cpt$matrix$index[which(cpt$matrix$index==cpt$changepoint)-1]
+  pol_x <- c(x1, x2, x2, x1)
+  pol_y <- c(y1, y1, y2, y2)
+  polygon(pol_x, pol_y, col= adjustcolor(mycol, alpha.f = .3), density = NULL, border = NA)
+}
+
+pdf(paste0(getwd(),"/Output/Fig_conceptual.pdf"), fonts = "Helvetica")
+
+layout(matrix(c(1,2,0,
+         3,4,5,
+         0,0,6,
+         0,0,7), ncol=3, byrow = T))
+par(mar=c(1,2,6,2))
+
+# Plot 1 -- all time series ####
+plot(c(1,100),c(0,1), pch=NA, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+for (i in 1:50) {
+  ar1 <- build_time_series(shift_time = sample(30:45, size = 1))
+  ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+  lines(ar1, lwd=.8, col=adjustcolor("grey", alpha.f = .4))
+}
+mydata <- ar1
+lines(mydata)
+mtext("a. 50 simulations", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+
+# Plot 2 -- example time series ####
+plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp <- sample_regular(mydata, n=length(mydata), messages = F)
+pol_break(cpt = cp)
+lines(mydata, col="black", lwd=1.5)
+text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
[…]

### 2020-01-04 R/sampling_iterative.R
Add the type of method as an output

@@ -108,6 +108,7 @@ sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol 
   
   # More elaborated output
                            out_list <- list()
+                    out_list$method <- "Iterative"
                out_list$changepoint <- changepoints$order.found[3]
                    out_list$final_n <- length(which_samples)
            out_list$changepoint_all <- changepoints

### 2020-01-04 R/sampling_random.R

@@ -53,6 +53,7 @@ sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_a
   
   # More elaborated output
                            out_list <- list()
+                    out_list$method <- "Random"
                out_list$changepoint <- changepoints$order.found[3]
                    out_list$final_n <- length(which_samples)
            out_list$changepoint_all <- changepoints

### 2020-01-04 R/sampling_regular.R

@@ -55,6 +55,7 @@ sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_
   
   # More elaborated output
                            out_list <- list()
+                    out_list$method <- "Regular"
                out_list$changepoint <- changepoints$order.found[3]
                    out_list$final_n <- length(which_samples)
            out_list$changepoint_all <- changepoints

### 2020-01-04 R/manuscript_draft.Rmd
small edits to draft manuscript
add figure for case study

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_knit$set(root.dir = dirname(getwd()))
+```
 
@@
 
+# Which sampling strategies are available?
+
+- Random
+- Regular
+- Iterative
+
+<img src="Fig_conceptual.pdf" alt="conceptual figure"  width="4200" height="6200">
+
 # Theoretical results
@@
 
+```{r case study read data output, include=FALSE}
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
+myoutput$method_f <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+
+```
+
+With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+library(ggplot2)
+library(gridExtra)
+p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method_f) + theme_bw() +
+  labs(title="a",x="", y="Distance to real changepoint")
+p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
+  facet_wrap(~method_f) + 
+  geom_violin() +
+  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()) + theme_bw() +
+  labs(title="b",x="", y="Distance to real changepoint")
+grid.arrange(p1, p2, nrow = 2)
+```
+
 # Discussion

### 2020-01-04 R/manuscript_draft.Rmd
add info on case study

@@
 
-# Lake case study
+# Case study
+
+We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
+Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26 m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
+It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
+Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, Bruel et al, 2018).
+Air temperature is now driving changes in plankton communities (Bruel et al, 2018).
+
+Data are reconstructed cladoceran assemblage from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period.
+Half of VAR 10.10 was sliced at a 1-cm interval and was used to count cladoceran remains (Bruel et al, 2018).
+We defined the "true" changepoint as the one identified from the complete record, in 1983/1988.
+We used the same ordination method as in Bruel et al (2018), but note that DCA is best used to conserve multiple gradients.
+As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988).
+We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).
+
+We used the full record (74 observation) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
+
+Random sampling perform the worst, as changepoint analysis was left to chance.
+Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error).
+Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necesserary to get the true changepoint
 
@@
 ```
-
-With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
 
@@
 
+Case study: With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
+Same results for the major change, with only 12% of the samples analyzed.
+However, need more samples to detect further changes.
+__Note that it would be possible to edit the function to go ahead and detect next changes once the main one has been detected.
+Is it worth coding?
+Shouldn't be too hard to do__
+
 # References

### 2020-01-04 R/case_study_Lake_Varese.R
figure for case study

@@ -1,0 +1,66 @@
+getpath4data <- function() {
+  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
+  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
+}
+# Function to return DCA
+library(vegan)
+do_dca <- function(dataset, xcol= NA) {
+  data_temp <- dataset[,-c(xcol)]
+  data_temp <- decorana(data_temp)
+  return(as.data.frame(data_temp$rproj))
+  #scores.species<-as.data.frame(data_temp$cproj)
+}
+
+# Read data
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+
+# DCA and changepoint on all data.
+# Note that only the most important changepoint is reported
+dca_all <- do_dca(vr,1)
+cpvar1 <- sample_regular(dca_all$DCA1, nrow(vr), input_vector = T)
+cpvar2 <- sample_regular(dca_all$DCA2, nrow(vr))
+
+# Do the subsampling
+vrindex <- 1:nrow(vr)
+
+myoutput <- NULL
+for (i in 6:30) {
+  vrsub <- sample_regular(vr, i, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+  myoutput <- c(myoutput, vrsub$method, vrsub$final_n, vrsub$changepoint)
+  vrsub <- sample_random(vr, i, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+  myoutput <- c(myoutput, vrsub$method, vrsub$final_n, vrsub$changepoint)
+  vrsub <- sample_iterative(vr, i, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+  myoutput <- c(myoutput, vrsub$method, vrsub$final_n, vrsub$changepoint)
+}
+myoutput <- data.frame(matrix(myoutput, ncol=3, byrow=T))
+myoutput[,2] <- as.numeric(paste(myoutput[,2]))
+myoutput[,3] <- as.numeric(paste(myoutput[,3]))
+colnames(myoutput) <- c("method", "final_n", "changepoint")
[…]

### 2020-01-04 R/Fig_conceptual.R
lwd=2 for points on figure f

@@ -119,7 +119,7 @@ pol_break(cpt = cp_iterative, col=2)
 lines(mydata, col="black", lwd=1.5)
 axis(1, lwd=2, at = c(-10,120))
 axis(2, lwd=2, at = c(-1,2))
-points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4)
+points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
 text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
 mtext("g. Iterative sampling (final)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
 

### 2020-01-06 R/sampling_iterative.R
remove notes for example test

@@ -118,6 +118,4 @@ sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol 
   invisible(out_list)
 } 
 
-test <- sample_iterative(ar1, n = 10)
-e.divisive(matrix(test$matrix$x), min.size = 2)$order.found[3]
 

### 2020-01-06 R/sampling_iterative.R
adding conditions any(is.na(x2)) in case no NA and no need to run na.locf

@@ -47,7 +47,7 @@ sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol 
   # transform into matrix to run e.divisive
   x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
   if(input_vector) x2[which_samples,2] <- x[which_samples] else x2[which_samples,2] <- x
-  x2[,2] <- na.locf(x2[,2])
+  if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
   temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3]
@@ -69,7 +69,7 @@ sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol 
     # transform into matrix to run e.divisive
     x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
     if(input_vector) x2[which_samples,2] <- x[which_samples] else x2[which_samples,2] <- x
-    x2[,2] <- na.locf(x2[,2])
+    if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
     x2 <- matrix(x2[,2])
     
     temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3]
@@ -94,7 +94,7 @@ sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol 
       # transform into matrix to run e.divisive
       x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
       if(input_vector) x2[which_samples,2] <- x[which_samples] else x2[which_samples,2] <- x
-      x2[,2] <- na.locf(x2[,2])
+      if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
       x2 <- matrix(x2[,2])
       
       temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3]

### 2020-01-06 R/sampling_random.R

@@ -43,7 +43,7 @@ sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_a
   # transform into matrix to run e.divisive
   x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
   if(input_vector) x2[which_samples,2] <- x[which_samples] else x2[which_samples,2] <- x
-  x2[,2] <- na.locf(x2[,2])
+  if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
   changepoints <- e.divisive(x2, min.size = 2)

### 2020-01-06 R/sampling_regular.R

@@ -45,7 +45,7 @@ sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_
   # transform into matrix to run e.divisive
   x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
   if(input_vector)  x[which_samples] else x2[which_samples,2] <- x
-  x2[,2] <- na.locf(x2[,2])
+  if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
   changepoints <- e.divisive(x2, min.size = 2)
@@ -65,5 +65,6 @@ sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_
   invisible(out_list)
            
 } 
+
 
 

### 2020-01-06 R/sampling_regular.R
found tiny issue

@@ -44,7 +44,7 @@ sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_
   
   # transform into matrix to run e.divisive
   x2 <- matrix(c(1:lastpoint,rep(NA,lastpoint)), ncol=2)
-  if(input_vector)  x[which_samples] else x2[which_samples,2] <- x
+  if(input_vector)  x2[which_samples,2] <- x[which_samples] else x2[which_samples,2] <- x
   if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   

### 2020-01-06 R/sampling_iterative.R
n2 was not specified in .sample_iterative()

@@ -15,8 +15,8 @@
 #'
 
 sample_iterative <- function(x, n, n2 = 5, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
-  .sample_iterative(x, n, messages, input_vector, xcol, DCA_axis)
-.sample_iterative <- function(x, n, messages, input_vector, xcol, DCA_axis) {
+  .sample_iterative(x, n, n2, messages, input_vector, xcol, DCA_axis)
+.sample_iterative <- function(x, n, n2, messages, input_vector, xcol, DCA_axis) {
   require(ecp)
   require(zoo)
   require(vegan)

### 2020-01-06 R/Fig_conceptual.R
add conceptual figure to draft and edit initial script to make it work (source function code mainly)

@@ -1,7 +1,13 @@
+R_U_KNITTING=T
+
+source(paste0(getwd(),"/R/build_time_series.R"))
+source(paste0(getwd(),"/R/sampling_regular.R"))
+source(paste0(getwd(),"/R/sampling_random.R"))
+source(paste0(getwd(),"/R/sampling_iterative.R"))
+
 library(wesanderson)
 mypal <- wes_palette("Chevalier1")
 
-ar1
 ar1 <- build_time_series()
 mydata <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
 
@@ -18,7 +24,7 @@ pol_break <- function(cpt, col=1, y1=-.2, y2=1.2) {
   polygon(pol_x, pol_y, col= adjustcolor(mycol, alpha.f = .3), density = NULL, border = NA)
 }
 
-pdf(paste0(getwd(),"/Output/Fig_conceptual.pdf"), fonts = "Helvetica")
+if(!R_U_KNITTING) pdf(paste0(getwd(),"/Output/Fig_conceptual.pdf"), fonts = "Helvetica")
 
 layout(matrix(c(1,2,0,
          3,4,5,
@@ -123,4 +129,4 @@ points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
 text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
 mtext("g. Iterative sampling (final)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
 
-dev.off()
+if(!R_U_KNITTING) dev.off()

### 2020-01-15 R/manuscript_draft.Rmd
add conceptual figure to draft and edit initial script to make it work (source function code mainly)
updates to introduction and formatting of article

@@
 ---
-title: "Sampling requirements to detect ecosystem change"
-author: "Rosalie Bruel and Easton White"
-date: "03/10/2019"
-output:
-  html_document:
-    df_print: paged
-    fig_caption: yes
-    number_sections: yes
-    toc: yes
-    toc_depth: 3
-    toc_float: yes
-  pdf_document:
-    toc: yes
-    toc_depth: '3'
-fontsize: 11pt
+bibliography: Whitebib.bib
+csl: ecology-letters.csl
 editor_options:
   chunk_output_type: console
-  df_print: paged
+fontsize: 12pt
+geometry: margin=1in
+header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
+  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
+  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite}
+output:
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: yes
 ---
@@
 
+Title: Sampling requirements to detect ecosystem change
+
+\vspace{7 mm}
+
+Authors: Rosalie Bruel and Easton R. White
+
+\vspace{5 mm}
+
+Addresses: \normalsize{\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}
+
+\vspace{5 mm}
+Keywords: {time-series} \vspace{10 mm}
+
+\vspace{5 mm}
+
+Number of words: XXXX \vspace{5 mm}
+
+Number of references:
+
+Number of figures and panels: X
+
+\vspace{5 mm}
+
+Submission as:
+
+\vspace{5 mm}
+
+Preprint available at:
+
+\clearpage
+
+\begin{center}
+\textbf{\Large Title here}
+\vspace{7 mm}
+
+Authors: \textsc{Rosalie Bruel and Easton R. White}
+\vspace{5 mm}
+
+Addresses: {\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}
+
+\end{center}
+
+\textbf{Abstract}
+
+Abstract here
+
+\vspace{3 mm}
+
+Keywords: {time series}
+
 # Things to do:
@@
 - writing methods/results pair of section
-- writing dicussion
+- writing discussion
 - setting up bibliography system
@@
 
-- Set up ideas about monitoring and cost of monitoring
+Environmental monitoring is one of the core components to modern research and management (citations).
+Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (citations).
+Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010).
+However, environmental monitoring is still often expensive and time-consuming (citations).
+Therefore, monitoring programs need to be designed in such a way to address the question of interest.
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation).
+Monitoring programs also need to be designed to be cost-effective given limited bugets (citations).
 
-Environmental monitoring is one of the core components to modern research and management (citations).
-However, many monitoring programs are expensive and time-consuming (citations).
-Therefore, an optimal...
+Accounting for these considerations in the design of a monitoring program is necessary to detect long-term ecological change.
+The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
+For example, @White2019 found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
+Other work has focused on the frequency of monitoring (citations), the allocation of resources spatially versus temporally (citations), and the ability to use citizen science to increase monitoring (citations).
+Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection (citation).
 
-- Idea of detecting change over time
-
-One of the primary goals of monitoring programs is to detect long-term ecological change...
-
-- Historical data and ability to choose how much to sample
-
-Because of new technological advances, there are many data sources that can be derived long after the actual  processes occured.
-For example, paleoecological lake cores can be extracted and ecological communities can be built from this...
-Similarly, samples for eDNA or videographical approaches can record time points of a system and be analyzed later (citations).
+Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
+For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie).
+Similarly, environmental samples (e.g.
+water, soil) can be saved with eDNA processed later (citation).
+Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
-Should the paleo core be analyzed at every centimeter?
+Should the paleoecological core be analyzed at every centimeter?
 Should the video be assessed at every minute?
-As long as analyzing samples is expensive, these tradeoffs will remain.
+As long as processsing samples is expensive, these trade-offs will remain.
 
 Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We first investigate these tools using a simulated-based apporach.
-We then apply these tools to the detection of change points for a set of lakes...
+We focus on paleoecological core samples as one example of this type of data.
+We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint.
+However, our approaches are widely applicable to other questions and data types.
+We first investigate these tools using a simulated-based approach to As a case study, we apply these tools to the detection of change points for a set of lakes.
+(more from Rosalie here).
 
@@
 
-<img src="Fig_conceptual.pdf" alt="conceptual figure"  width="4200" height="6200">
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual figure for... \\label{fig:conceptual}'}
+source(paste0(getwd(),"/R/Fig_conceptual.R"))
+
+```
 
@@
 We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
-The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occured ($\tau$):
+The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occurred ($\tau$):
 
@@
 We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
-Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26 m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
+Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26m), monomictic lake, in the subalpine region of north-western Italy (238m asl).
 It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
@@
 Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error).
-Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necesserary to get the true changepoint
+Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint
 
-```{r case study read data output, include=FALSE}
+```{r case study read data output, include=FALSE,eval=F}
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
+```{r echo=FALSE, message=FALSE, warning=FALSE,eval=F}
 library(ggplot2)
@@
 
+\clearpage
+
 # References

### 2020-01-15 R/manuscript_draft.Rmd
edits draft

--- 2020-01-04 R/manuscript_draft.Rmd
@@
 
+Methods to dectect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
+
 - Historical data and ability to choose how much to sample
@@
 
-<img src="Fig_conceptual.pdf" alt="conceptual figure"  width="4200" height="6200">
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5}
+source(paste0(getwd(),"/R/Fig_conceptual.R"))
+
+```
+
+We hypothesize  that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
 
@@
 We defined the "true" changepoint as the one identified from the complete record, in 1983/1988.
+Changepoints were detected on first component computed from the data sub-sampled.
 We used the same ordination method as in Bruel et al (2018), but note that DCA is best used to conserve multiple gradients.
@@
 We used the full record (74 observation) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
+Iterative sampling final number could be lower than maximum number if both samples framing the shift were analyzed.
 
@@
 
+__Main point__: Changepoint was detected even though we used a transformed vector (from community to first component of DCA)
+
 # Discussion
@@
 
+Limits: Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
+The lower number of input data prior to ordination did not impact the finding of the main transition.
+However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.
+
 # References

### 2020-01-15 R/manuscript_draft.Rmd

--- 2020-01-15 R/manuscript_draft.Rmd
@@
 
+Methods to dectect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
+
+- Historical data and ability to choose how much to sample
+
 Accounting for these considerations in the design of a monitoring program is necessary to detect long-term ecological change.
@@
 ```
+
+We hypothesize  that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
 

### 2020-01-15 R/manuscript_draft.Rmd
eval=T for case studies figures

@@
 
-```{r case study read data output, include=FALSE,eval=F}
+```{r case study read data output, include=FALSE}
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE,eval=F}
+```{r echo=FALSE, message=FALSE, warning=FALSE}
 library(ggplot2)

### 2020-01-22 R/manuscript_draft.Rmd
modified theoretical model structure and updated equations in text

@@
 
-# Theoretical results
+# Theoretical approach
 
@@
 
-We began with a theoretical exploration of the determiners of sampling required to detect a changepoint.
-We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition.
-The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occurred ($\tau$):
+Theoretical plots
+
+- One combined plot to see the effect (on prob correct or number of samples) of each parameter - should do this with one of the three types of sampling (probably regular)
+- Theoretical plot comparing the three different methods - we could choose one set of parameter values and show how the various methods perform with different sampling amounts (could use same plot structure as the case study)
+- SM plot that shows how the three different methods are affected by changes in the underlying parameter values
+
+## Theoretical model
+
+We began with a theoretical exploration of the sampling requirements to detect a changepoint.
+We modeled a simple first order autoregressive (AR-1) process with a response variable that represents either population size, biodiversity, or some metric of community composition.
+The model includes temporal autocorrelation ($\phi$), the mean of the process ($\mu_X$), and a white noise term ($w_t$).
+The white noise term is a normal distribution with mean ($\mu_w$) and variance ($\sigma$).
 
 $$
-\mbox{Model here - need pair of equations}
+\begin{array}{rcl} 
+X_t & = & \mu_X + \phi (X_{t-1} - \mu_X) + w_t \\ 
+w_t & \sim & Norm(\mu_w,\sigma)
+\end{array}
 $$
+We included a changepoint by shifting $\mu_X$ at time $\tau$ given a specific shift size ($\delta$).
 
-We then explored how each of these model components affected our ability to detect a change point.
+We explored how each of these model parameters affected our ability to detect a change point.
 We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
@@
 
-In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\\sigma$) and autocorrelation ($\\phi$).
+```{r theoretical_results, fig.cap='Theoretical results here detailing parameter sensitivity - we should make this about sample size instead of probability correct',message=F,warning=F,echo=F,eval=F}
+
+source('R/sampling_regular.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+
+require(tidyverse)
+
+par(mfrow=c(2,2))
+# Sigma plot
+sigma15 <- param_sensitivity('sigma',15)
+plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+sigma20 <- param_sensitivity('sigma',20)
+points(sigma20,ylab='Probability correct',xlab='sigma',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
+
+# phi plot
+phi15 <- param_sensitivity('phi',15)
+plot(phi15,ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+phi20 <- param_sensitivity('phi',20)
+points(phi20,ylab='Probability correct',xlab='phi',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
+
+```
+
+## Theoretical results
+
+In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\sigma$) and autocorrelation ($\phi$).
 We also found that the probability of correct changepoint detection increases with larger shift sizes.

### 2020-01-22 R/build_time_series.R

@@ -2,14 +2,16 @@
 
 # Build ar-1 time series with various model inputs
 
-build_time_series <- function(mu=0,sigma=2,phi=0.5,shift_size=0.25,shift_time=50,max_time=100){
+build_time_series <- function(mu=0,sigma=3,phi=0.5,shift_size=0.75,shift_time=50,max_time=100){
   
   ar1 <- vector(mode = 'numeric',max_time)
-  ar1[1] <- 100
+  meanXt = 100
+  ar1[1] <- meanXt
   #shift_time = sample(5:(max_time-5),size = 1)
   for (t in 2:max_time){
-    if(t == shift_time){ar1[t-1]=(1-shift_size)*ar1[1]}
-    ar1[t] <- ar1[t-1] + phi*rnorm(1,mean=mu,sd = sigma)
+    #if(t == shift_time){ar1[t-1]=(1-shift_size)*ar1[1]}
+    if(t == shift_time){meanXt = (1-shift_size)*ar1[1]}
+    ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)
   }
   return(ar1) 
 }

### 2020-01-23 R/manuscript_draft.Rmd
edit the case study figure to visualize behavior with several starting number of sample, and for up to 3rd changepoint

@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 9}
 library(ggplot2)
 library(gridExtra)
-p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
+p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
   geom_point(alpha = .3) +
   facet_wrap(~method_f) + theme_bw() +
-  labs(title="a",x="", y="Distance to real changepoint")
-p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
-  facet_wrap(~method_f) + 
-  geom_violin() +
-  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()) + theme_bw() +
-  labs(title="b",x="", y="Distance to real changepoint")
-grid.arrange(p1, p2, nrow = 2)
+  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.95, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method_f) + theme_bw() +
+  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.95, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method_f) + theme_bw() +
+  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.95, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+p3
+grid.arrange(p1,p2,p3,nrow=3)
 ```

### 2020-01-23 R/manuscript_draft.Rmd
change some figure output options

--- 2020-01-22 R/manuscript_draft.Rmd
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8}
 library(ggplot2)
 library(gridExtra)
-p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
+p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
   geom_point(alpha = .3) +
   facet_wrap(~method_f) + theme_bw() +
-  labs(title="a",x="", y="Distance to real changepoint")
-p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
-  facet_wrap(~method_f) + 
-  geom_violin() +
-  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
-  theme(axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()) + theme_bw() +
-  labs(title="b",x="", y="Distance to real changepoint")
-grid.arrange(p1, p2, nrow = 2)
+  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method_f) + theme_bw() +
+  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method_f) + theme_bw() +
+  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.6),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+grid.arrange(p1,p2,p3,nrow=3)
 ```

### 2020-01-28 R/manuscript_draft.Rmd
contribute to manuscript

@@
 
-Authors: Rosalie Bruel and Easton R. White
+Authors: Rosalie Bruel$^{1}$ and Easton R. White$^{2}$
 
@@
 
-Addresses: \normalsize{\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}
+Addresses: \normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
 
@@
 
-Authors: \textsc{Rosalie Bruel and Easton R. White}
+Authors: \textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
 \vspace{5 mm}
 
-Addresses: {\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}
-
+Addresses: \normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
 \end{center}
@@
 Environmental monitoring is one of the core components to modern research and management (citations).
-Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (citations).
+Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (Lovett et al, 2007).
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010).
@@
 
+An application of our study we foresee applies to paleosequences.
+Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
+Samples can be taken before and after perturbation to compare community compostion (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
+
+Maybe extent to paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
+
 We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
-Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26m), monomictic lake, in the subalpine region of north-western Italy (238m asl).
+Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
 It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
-Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, Bruel et al, 2018).
-Air temperature is now driving changes in plankton communities (Bruel et al, 2018).
+Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, @bruel_seeking_2018).
+Air temperature is now driving changes in plankton communities (@bruel_seeking_2018).
 
-Data are reconstructed cladoceran assemblage from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period.
-Half of VAR 10.10 was sliced at a 1-cm interval and was used to count cladoceran remains (Bruel et al, 2018).
+Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period (@bruel_seeking_2018).
 We defined the "true" changepoint as the one identified from the complete record, in 1983/1988.
@@
 
+About recommendation to reduce the frequency of data collected as part of monitoring program:  
+While long-term monitoring program are often expensive, these cost are nothing compared to the value of the resources monitored and the policies affecting these resources (@lovett_who_2007).
+Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may indeed not be necessary to manage ecosystems in the future.
+
 # References

### 2020-02-05 R/manuscript_draft.Rmd
added pieces to theoretical results section

--- 2020-01-23 R/manuscript_draft.Rmd
@@
 We began with a theoretical exploration of the sampling requirements to detect a changepoint.
-We modeled a simple first order autoregressive (AR-1) process with a response variable that represents either population size, biodiversity, or some metric of community composition.
+We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable that represents either population size, biodiversity, or some metric of community composition.
 The model includes temporal autocorrelation ($\phi$), the mean of the process ($\mu_X$), and a white noise term ($w_t$).
@@
 
+```{r theoretical_results_min_time, fig.cap='Theoretical results here detailing parameter sensitivity - we should make this about sample size instead of probability correct',message=F,warning=F,echo=F,eval=T}
+
+source('R/sampling_regular.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+source('R/calc_min_time2.R')
+
+require(tidyverse)
+
+par(mfrow=c(1,3))
+# Sigma plot
+sigma_vec = seq(0.1,20,1)
+min_time=NULL
+
+for (param_num in 1:length(sigma_vec)){
+  min_time = c(min_time,calc_min_time('sigma',sigma_vec[param_num]))
+  print(min_time)
+}
+
+plot(sigma_vec,min_time,ylab='Number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,25),col=1)
+
+phi_vec = seq(-1,1,0.2)
+min_time=NULL
+
+for (param_num in 1:length(phi_vec)){
+  min_time = c(min_time,calc_min_time('phi',phi_vec[param_num]))
+}
+
+plot(phi_vec,min_time,ylab='Number of samples required',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+# Shift size
+
+shift_size_vec = seq(0.05,0.95,0.1)
+min_time=NULL
+
+for (param_num in 1:length(shift_size_vec)){
+  min_time = c(min_time,calc_min_time('shift_size',shift_size_vec[param_num]))
+}
+
+plot(shift_size_vec,min_time,ylab='Number of samples required',xlab='shift size (percent)',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+```
+
 ## Theoretical results
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8}
+```{r echo=FALSE, message=FALSE, warning=FALSE}
 library(ggplot2)
 library(gridExtra)
-p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
   geom_point(alpha = .3) +
   facet_wrap(~method_f) + theme_bw() +
-  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method_f) + theme_bw() +
-  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method_f) + theme_bw() +
-  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-grid.arrange(p1,p2,p3,nrow=3)
+  labs(title="a",x="", y="Distance to real changepoint")
+p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
+  facet_wrap(~method_f) + 
+  geom_violin() +
+  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()) + theme_bw() +
+  labs(title="b",x="", y="Distance to real changepoint")
+grid.arrange(p1, p2, nrow = 2)
 ```

### 2020-02-05 R/manuscript_draft.Rmd

--- 2020-01-28 R/manuscript_draft.Rmd
@@
 points(phi20,ylab='Probability correct',xlab='phi',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
+
+```
+
+```{r theoretical_results_min_time, fig.cap='Theoretical results here detailing parameter sensitivity - we should make this about sample size instead of probability correct',message=F,warning=F,echo=F,eval=T}
+
+source('R/sampling_regular.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+source('R/calc_min_time2.R')
+
+require(tidyverse)
+
+par(mfrow=c(1,3))
+# Sigma plot
+sigma_vec = seq(0.1,20,1)
+min_time=NULL
+
+for (param_num in 1:length(sigma_vec)){
+  min_time = c(min_time,calc_min_time('sigma',sigma_vec[param_num]))
+  print(min_time)
+}
+
+plot(sigma_vec,min_time,ylab='Number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,25),col=1)
+
+phi_vec = seq(-1,1,0.2)
+min_time=NULL
+
+for (param_num in 1:length(phi_vec)){
+  min_time = c(min_time,calc_min_time('phi',phi_vec[param_num]))
+}
+
+plot(phi_vec,min_time,ylab='Number of samples required',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+
+# Shift size
+
+shift_size_vec = seq(0.05,0.95,0.1)
+min_time=NULL
+
+for (param_num in 1:length(shift_size_vec)){
+  min_time = c(min_time,calc_min_time('shift_size',shift_size_vec[param_num]))
+}
+
+plot(shift_size_vec,min_time,ylab='Number of samples required',xlab='shift size (percent)',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
 

### 2020-02-05 R/manuscript_draft.Rmd
modified theory figs and added text

--- 2020-01-28 R/manuscript_draft.Rmd
@@
 
-Authors: \textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
+\textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
 \vspace{5 mm}
 
-Addresses: \normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
+\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
 \end{center}
@@
 We specifically study how the amount of samples included and the type of sampling affects the detection probability.
+For simulations, statistical power is the fraction of simulations that were able to detect a changepoint.
+We define a detection as an estimate within five time points (over 100 time points) of the true changepoint.
+We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.
 
-```{r theoretical_results, fig.cap='Theoretical results here detailing parameter sensitivity - we should make this about sample size instead of probability correct',message=F,warning=F,echo=F,eval=F}
+```{r theoretical_results,message=F,warning=F,echo=F,eval=F}
 
 source('R/sampling_regular.R')
+source('R/sampling_random.R')
 source('R/build_time_series.R')
@@
 
+```{r theoretical_results_min_time,message=F,warning=F,echo=F,eval=F}
+
+source('R/sampling_regular.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+source('R/calc_min_time2.R')
+require(tidyverse)
+
+# Sigma results
+sigma_vec = seq(0.1,30,4)
+min_time  = rep(0,times=length(sigma_vec))
+sigma_output <- data.frame(sigma_vec,min_time)
+
+for (param_num in 1:nrow(sigma_output)){
+  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
+  #print(sigma_output)
+}
+
+write.csv(x = sigma_output,file = 'Output/theoretical_results_sigma.csv',quote=FALSE)
+
+# Phi results
+
+phi_vec = seq(-0.6,0.6,0.2)
+min_time  = rep(0,times=length(phi_vec))
+phi_output <- data.frame(phi_vec,min_time)
+
+for (param_num in 1:nrow(phi_output)){
+  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
+  #print(phi_output)
+}
+
+write.csv(x = phi_output,file = 'Output/theoretical_results_phi.csv',quote=FALSE)
+
+# Shift size
+
+shift_size_vec = rev(seq(0.1,0.9,0.05))
+min_time  = rep(0,times=length(shift_size_vec))
+shift_size_output <- data.frame(shift_size_vec,min_time)
+
+for (param_num in 1:nrow(shift_size_output)){
+  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
+  #print(shift_size_output)
+}
+
+write.csv(x = shift_size_output,file = 'Output/theoretical_results_shift_size.csv',quote=FALSE)
+
+```
+
+```{r theoretical_results_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
+The default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
+\\label{fig:theoretical_min_time}',message=F,warning=F,echo=F,eval=T}
+
+par(mfrow=c(1,3))
+
+sigma <- read.csv('Output/theoretical_results_sigma.csv',header=TRUE)
+plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+
+# Phi results
+
+phi <- read.csv('Output/theoretical_results_phi.csv',header=TRUE)
+plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+
+# Shift size results
+
+shift_size <- read.csv('Output/theoretical_results_shift_size.csv',header=TRUE)
+plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+
+```
+
 ## Theoretical results
 
-In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\sigma$) and autocorrelation ($\phi$).
-We also found that the probability of correct changepoint detection increases with larger shift sizes.
-
-The probability of correcting identifying a changepoint, of course, depends strongly on the number of samples.
+In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$).
+We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size.
+Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
+\ref{fig:theoretical_min_time}).
 
@@
 
-# Discussion
+# General Discussion
+
+Outline:
+
+- theoretical results
+- example with the case study
+- how this fits in with other work
+- limitations and future work
 
@@
 
-Limits: Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
+Limitations:
+
+- It is not limited to paleo data
+- Depends on the question being asked
+- Specific number of samples depends on the system dynamics
+
+Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
 The lower number of input data prior to ordination did not impact the finding of the main transition.
 However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.
+
+Concluding paragraph:
 

### 2020-02-18 R/sampling_iterative.R
n2 for iterative code should be n2-2 (because first and last point are already sampled

@@ -35,7 +35,7 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
   if(is.null(n)) n <- lastpoint
   
   # Sample initial number of samples
-  which_samples <- seq(1,lastpoint,length.out = n2)
+  which_samples <- seq(1,lastpoint,length.out = n2-2)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   

### 2020-02-18 R/sampling_iterative.R
actually n2 was correct, so going back to previous version

@@ -35,7 +35,7 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
   if(is.null(n)) n <- lastpoint
   
   # Sample initial number of samples
-  which_samples <- seq(1,lastpoint,length.out = n2-2)
+  which_samples <- seq(1,lastpoint,length.out = n2)
   which_samples <- round(which_samples)
   which_samples <- which_samples[order(which_samples)]
   

### 2020-02-18 R/manuscript_draft.Rmd
added supp mat document and edited theoretical results figure

@@
     keep_tex: yes
-    number_sections: yes
+    number_sections: no
 ---
@@
 
-Addresses: \normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
+Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA $^{2}$Department of Biology, University of Vermont, VT 05405, USA
 
@@
 
+<!--
 # Things to do:
@@
 - Case study of paleo data shows the utility of iterative sampling in detecting change points
+-->
 
-# Introduction and setup
+# Introduction
 
@@
 
-# Which sampling strategies are available?
+# Sampling approaches and changepoint detection
 
-- Random
-- Regular
-- Iterative
+For both our simulations and case study, we investigate the effect of different sampling strategies on our ability to detect a changepoint.
+We begin by either creating simulated time series or using an actual paleoecological time series (Fig.
+\ref{fig:conceptual}).
+We then subsampled each time series (citation) to test the effect of three different sampling approaches along with varying the sample size (Fig.
+\ref{fig:conceptual}c-f).
+We compared the subsampled time series to the full time series as a measure of the effectiveness of that sampling approach.
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual figure for... \\label{fig:conceptual}'}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual diagram illustrating the process of taking (a) simulations of a time series and (b) selecting a single simulation to analyze with three different sampling approaches: (c) random, (d) regular, and (e) iterative.
+The iterative sampling approach requires (f) adding samples around a detected changepoint until (g) a certain level of accuracy is achieved.
+\\label{fig:conceptual}'}
 source(paste0(getwd(),"/R/Fig_conceptual.R"))
-
 ```
 
+The random sampling approach involves taking a set number of random points throughout the time series (Fig.
+\ref{fig:conceptual}c).
+In the context of sediment cores, this would mean analzying community composition as random spots along the core.
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
+Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig.
+\ref{fig:conceptual}d).
+Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
+\ref{fig:conceptual}e-g).
 
-# Theoretical approach
+Rosalie: Can you write a paragraph here detailing the changepoint detection process?
 
+# Simulation approach
+
+<!--
 - explain the modeling approach
@@
 - SM plot that shows how the three different methods are affected by changes in the underlying parameter values
+
+-->
 
@@
 source('R/sampling_regular.R')
+source('R/sampling_random.R')
 source('R/build_time_series.R')
@@
   sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
-  #print(sigma_output)
+  print(sigma_output)
 }
@@
 
-phi_vec = seq(-0.6,0.6,0.2)
+phi_vec = seq(-0.8,0.8,0.2)
 min_time  = rep(0,times=length(phi_vec))
@@
   phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
-  #print(phi_output)
+  print(phi_output)
 }
@@
 
-shift_size_vec = rev(seq(0.1,0.9,0.05))
+shift_size_vec = rev(seq(0.1,0.9,0.025))
 min_time  = rep(0,times=length(shift_size_vec))
@@
   shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
-  #print(shift_size_output)
+  print(shift_size_output)
 }
@@
 sigma <- read.csv('Output/theoretical_results_sigma.csv',header=TRUE)
-plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
 
@@
 phi <- read.csv('Output/theoretical_results_phi.csv',header=TRUE)
-plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
 
@@
 shift_size <- read.csv('Output/theoretical_results_shift_size.csv',header=TRUE)
-plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)
+plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
 
@@
 \ref{fig:theoretical_min_time}).
+
+```{r theoretical_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
+
+source('R/sampling_regular.R')
+source('R/sampling_random.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+source('R/calc_min_time2.R')
+require(tidyverse)
+
+# Need to run each method on a set of simulated time series. We would then have 100 runs for each method and each sampling amount
+
+approach <- c('random')#,'regular','iterative')
+number_samples <- 5:30
+simulation_number <- 1:50
+output_theory_approach <- data.frame(expand.grid(approach,number_samples,simulation_number))
+names(output_theory_approach) <- c('approach', 'number_samples', 'simulation_number')
+output_theory_approach$distance_to_real = 0
+
+# Build function that takes in approach, simulated time series, and number of samples. The function should then spit out the distance to the real changepoint
+
+# Set of time series (need other params)
+params <- tibble(shift_time = sample(30:70,max(simulation_number),replace = T))
+
+time_series <- params %>%
+  pmap(build_time_series)
+
+# Run through params
+
+for (i in 1:nrow(output_theory_approach)){
+
+  output_theory_approach$distance_to_real[i] <- abs(params$shift_time[output_theory_approach$simulation_number[i]] - calc_distance_to_real(output_theory_approach$approach[i],output_theory_approach$number_samples[i],time_series[[output_theory_approach$simulation_number[i]]]))
+  #print(i)
+}
+
+write.csv(x = output_theory_approach,file = 'Output/theoretical_comparing_approaches.csv',quote = FALSE )
+```
 

### 2020-02-18 R/supp_material.Rmd
added supp mat document and edited theoretical results figure

--- 2020-02-18 R/manuscript_draft.Rmd
@@
   \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite}
+  \renewcommand{\thepage}{S\arabic{page}}
+  \renewcommand{\thesection}{Appendix S\arabic{section}} \renewcommand{\thetable}{S\arabic{table}}
+  \renewcommand{\thefigure}{S\arabic{figure}}
+  \usepackage{amsmath}
+  \renewcommand{\theequation}{S\arabic{equation}}
 output:
@@
 
-Title: Sampling requirements to detect ecosystem change
-
-\vspace{7 mm}
-
-Authors: Rosalie Bruel$^{1}$ and Easton R. White$^{2}$
-
-\vspace{5 mm}
-
-Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA $^{2}$Department of Biology, University of Vermont, VT 05405, USA
-
-\vspace{5 mm}
-Keywords: {time-series} \vspace{10 mm}
-
-\vspace{5 mm}
-
-Number of words: XXXX \vspace{5 mm}
-
-Number of references:
-
-Number of figures and panels: X
-
-\vspace{5 mm}
-
-Submission as:
-
-\vspace{5 mm}
-
-Preprint available at:
-
-\clearpage
-
 \begin{center}
-\textbf{\Large Title here}
+\textbf{\Large Supplementary Material for: Title here}
 \vspace{7 mm}
@@
 
-\textbf{Abstract}
+\vspace{2 mm}
 
-Abstract here
+\tableofcontents
 
-\vspace{3 mm}
+\vspace{1cm}
 
-Keywords: {time series}
+Data and code for all the figures can be found at (https://github.com/rosalieb/temporal-sampling.git). All the analyses were run in R.
 
-<!--
-# Things to do:
+\clearpage
 
-- writing intro
-- writing methods/results pair of section
-- writing discussion
-- setting up bibliography system
-- Possible journals: Methods Ecology and Evolution, Biological Conservation, *Journal of Environmental Management*, *Environmental Monitoring and Assessment*, BioScience
+# Additional figures
 
-# Main messages of manuscript
-
-- We can choose the number of samples and effort we use...
-- Number of samples required depends on question asked and the time series parameters
-- Case study of paleo data shows the utility of iterative sampling in detecting change points
--->
-
-# Introduction
-
-Environmental monitoring is one of the core components to modern research and management (citations).
-Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (Lovett et al, 2007).
-Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010).
-However, environmental monitoring is still often expensive and time-consuming (citations).
-Therefore, monitoring programs need to be designed in such a way to address the question of interest.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation).
-Monitoring programs also need to be designed to be cost-effective given limited bugets (citations).
-
-Methods to dectect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
-
-- Historical data and ability to choose how much to sample
-
-Accounting for these considerations in the design of a monitoring program is necessary to detect long-term ecological change.
-The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
-For example, @White2019 found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
-Other work has focused on the frequency of monitoring (citations), the allocation of resources spatially versus temporally (citations), and the ability to use citizen science to increase monitoring (citations).
-Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection (citation).
-
-Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie).
-Similarly, environmental samples (e.g.
-water, soil) can be saved with eDNA processed later (citation).
-Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
-In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
-Should the paleoecological core be analyzed at every centimeter?
-Should the video be assessed at every minute?
-As long as processsing samples is expensive, these trade-offs will remain.
-
-Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We focus on paleoecological core samples as one example of this type of data.
-We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint.
-However, our approaches are widely applicable to other questions and data types.
-We first investigate these tools using a simulated-based approach to As a case study, we apply these tools to the detection of change points for a set of lakes.
-(more from Rosalie here).
-
-# Sampling approaches and changepoint detection
-
-For both our simulations and case study, we investigate the effect of different sampling strategies on our ability to detect a changepoint.
-We begin by either creating simulated time series or using an actual paleoecological time series (Fig.
-\ref{fig:conceptual}).
-We then subsampled each time series (citation) to test the effect of three different sampling approaches along with varying the sample size (Fig.
-\ref{fig:conceptual}c-f).
-We compared the subsampled time series to the full time series as a measure of the effectiveness of that sampling approach.
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual diagram illustrating the process of taking (a) simulations of a time series and (b) selecting a single simulation to analyze with three different sampling approaches: (c) random, (d) regular, and (e) iterative.
-The iterative sampling approach requires (f) adding samples around a detected changepoint until (g) a certain level of accuracy is achieved.
-\\label{fig:conceptual}'}
-source(paste0(getwd(),"/R/Fig_conceptual.R"))
-```
-
-The random sampling approach involves taking a set number of random points throughout the time series (Fig.
-\ref{fig:conceptual}c).
-In the context of sediment cores, this would mean analzying community composition as random spots along the core.
-We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
-Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig.
-\ref{fig:conceptual}d).
-Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
-\ref{fig:conceptual}e-g).
-
-Rosalie: Can you write a paragraph here detailing the changepoint detection process?
-
-# Simulation approach
-
-<!--
-- explain the modeling approach
-- show the model
-- explain the analysis we will do
-- explain the theoretical model results with a figure
-- explain effect of different sampling strategies in this context
-
-Theoretical plots
-
-- One combined plot to see the effect (on prob correct or number of samples) of each parameter - should do this with one of the three types of sampling (probably regular)
-- Theoretical plot comparing the three different methods - we could choose one set of parameter values and show how the various methods perform with different sampling amounts (could use same plot structure as the case study)
-- SM plot that shows how the three different methods are affected by changes in the underlying parameter values
-
--->
-
-## Theoretical model
-
-We began with a theoretical exploration of the sampling requirements to detect a changepoint.
-We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable that represents either population size, biodiversity, or some metric of community composition.
-The model includes temporal autocorrelation ($\phi$), the mean of the process ($\mu_X$), and a white noise term ($w_t$).
-The white noise term is a normal distribution with mean ($\mu_w$) and variance ($\sigma$).
-
-$$
-\begin{array}{rcl} 
-X_t & = & \mu_X + \phi (X_{t-1} - \mu_X) + w_t \\ 
-w_t & \sim & Norm(\mu_w,\sigma)
-\end{array}
-$$
-We included a changepoint by shifting $\mu_X$ at time $\tau$ given a specific shift size ($\delta$).
-
-We explored how each of these model parameters affected our ability to detect a change point.
-We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
-We specifically study how the amount of samples included and the type of sampling affects the detection probability.
-For simulations, statistical power is the fraction of simulations that were able to detect a changepoint.
-We define a detection as an estimate within five time points (over 100 time points) of the true changepoint.
-We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.
+## Simulation approach
 
@@
 
-## Theoretical results
-
-In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$).
-We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size.
-Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
-\ref{fig:theoretical_min_time}).
-
 ```{r theoretical_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
@@
 
-approach <- c('random')#,'regular','iterative')
-number_samples <- 5:30
-simulation_number <- 1:50
+approach <- c('random','regular','iterative')
+number_samples <- 1:30
+simulation_number <- 1:100
 output_theory_approach <- data.frame(expand.grid(approach,number_samples,simulation_number))
@@
 # Set of time series (need other params)
-params <- tibble(shift_time = sample(30:70,max(simulation_number),replace = T))
+params <- tibble(shift_time = sample(30:70,nrow(params),replace = T))
 
@@
 
-  output_theory_approach$distance_to_real[i] <- abs(params$shift_time[output_theory_approach$simulation_number[i]] - calc_distance_to_real(output_theory_approach$approach[i],output_theory_approach$number_samples[i],time_series[[output_theory_approach$simulation_number[i]]]))
-  #print(i)
+  output_theory_approach$distance_to_real[i] <- calc_distance_to_real(output_theory_approach$approach,output_theory_approach$number_samples,time_series[output_theory_approach$simulation_number])
 }
 
-write.csv(x = output_theory_approach,file = 'Output/theoretical_comparing_approaches.csv',quote = FALSE )
 ```
 
-# Case study
-
-An application of our study we foresee applies to paleosequences.
-Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
-Samples can be taken before and after perturbation to compare community compostion (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
-
-Maybe extent to paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
-
-We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
-Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
-It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
-Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, @bruel_seeking_2018).
-Air temperature is now driving changes in plankton communities (@bruel_seeking_2018).
-
-Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period (@bruel_seeking_2018).
-We defined the "true" changepoint as the one identified from the complete record, in 1983/1988.
-Changepoints were detected on first component computed from the data sub-sampled.
-We used the same ordination method as in Bruel et al (2018), but note that DCA is best used to conserve multiple gradients.
-As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988).
-We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).
-
-We used the full record (74 observation) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
-Iterative sampling final number could be lower than maximum number if both samples framing the shift were analyzed.
-
-Random sampling perform the worst, as changepoint analysis was left to chance.
-Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error).
-Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint
+## Case study
 
@@
 
-# General Discussion
-
-Outline:
-
-- theoretical results
-- example with the case study
-- how this fits in with other work
-- limitations and future work
-
-Case study: With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
-Same results for the major change, with only 12% of the samples analyzed.
-However, need more samples to detect further changes.
-__Note that it would be possible to edit the function to go ahead and detect next changes once the main one has been detected.
-Is it worth coding?
-Shouldn't be too hard to do__
-
-\clearpage
-
-Limitations:
-
-- It is not limited to paleo data
-- Depends on the question being asked
-- Specific number of samples depends on the system dynamics
-
-Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
-The lower number of input data prior to ordination did not impact the finding of the main transition.
-However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.
-
-Concluding paragraph:
-
-About recommendation to reduce the frequency of data collected as part of monitoring program:  
-While long-term monitoring program are often expensive, these cost are nothing compared to the value of the resources monitored and the policies affecting these resources (@lovett_who_2007).
-Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may indeed not be necessary to manage ecosystems in the future.
-
-# References
+## References

### 2020-02-18 R/sampling_iterative.R
cap the code so it stops at n if n2=n

@@ -53,7 +53,7 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
   
   temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3:(2+c)]
   
-  if(all(is.na(temp_chgpt))) {
+  if(all(is.na(temp_chgpt)) && length(which_samples) < n ) {
     which_samples <- seq(1,2,lastpoint,length.out = n2*2-1)
     which_samples <- round(which_samples)
     which_samples <- which_samples[order(which_samples)]
@@ -79,32 +79,34 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
   # Now, start adding samples
   c_order = 1
   for(i in 1:(n-n2)) {
-    if (c_order<=c && !is.na(temp_chgpt[c_order])) {
-      # Here we only want to add sample if we haven't already found the closest changepoint
-      if(!round((which_samples[which(which_samples==temp_chgpt[c_order])-1]+which_samples[which(which_samples==temp_chgpt[c_order])])/2) %in% which_samples) {
-        which_samples <- c(which_samples, round((which_samples[which(which_samples==temp_chgpt[c_order])-1]+which_samples[which(which_samples==temp_chgpt[c_order])])/2))
-        which_samples <- which_samples[order(which_samples)]
-        
-        # Reset x
-        x <- x_stable
-        
-        # Do DCA to get only one vector if input is a data frame
-        if(!input_vector) {
-          x <- do_dca(x[which_samples,], xcol)
-          x <- x[,DCA_axis]
+    if (length(which_samples) < n) {
+      if (c_order<=c && !is.na(temp_chgpt[c_order])) {
+        # Here we only want to add sample if we haven't already found the closest changepoint
+        if(!round((which_samples[which(which_samples==temp_chgpt[c_order])-1]+which_samples[which(which_samples==temp_chgpt[c_order])])/2) %in% which_samples) {
+          which_samples <- c(which_samples, round((which_samples[which(which_samples==temp_chgpt[c_order])-1]+which_samples[which(which_samples==temp_chgpt[c_order])])/2))
+          which_samples <- which_samples[order(which_samples)]
+          
+          # Reset x
+          x <- x_stable
+          
[…]

### 2020-02-18 R/manuscript_draft.Rmd
edit figure from case study, and write up results section

--- 2020-02-05 R/manuscript_draft.Rmd
@@
 We explored how each of these model parameters affected our ability to detect a change point.
-We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year.
+We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every interval.
 We specifically study how the amount of samples included and the type of sampling affects the detection probability.
@@
 
+## Case study background
+
 An application of our study we foresee applies to paleosequences.
@@
 
-Maybe extent to paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
-
 We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
@@
 It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
-Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, @bruel_seeking_2018).
+Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Crosta 1999, @bruel_seeking_2018).
 Air temperature is now driving changes in plankton communities (@bruel_seeking_2018).
@@
 
+## Case study results
+
 Random sampling perform the worst, as changepoint analysis was left to chance.
-Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error).
-Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint
+Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint was left  to chance.
+Iterative sampling (Fig.
+\ref{fig:case_study_number_sample} (c)), where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint.
+Note that when the initial number of samples was set to 7 Fig.
+\ref{fig:case_study_number_sample} (c), square markers), the distance to the real changepoint was initially very high.
+Without any prior, the initial number sampled fell far away from the true changepoint.
+
+We tested how the three methods performed at detecting other changepoints (as three changepoints were detected in the initial study, @bruel_seeking_2018).
+Iterative sampling performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials \textcolor{red}{2}).
+The initial number of samples should be chosen depending on the objective of the study.
+If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoritical results?).
+If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time serie to have a good initial ranking.
 
@@
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
-myoutput$method_f <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
 
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample}'}
 library(ggplot2)
-library(gridExtra)
+#devtools::install_github("rensa/stickylabeller")
+library(stickylabeller)
 p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
@@
   geom_point(alpha = .3) +
-  facet_wrap(~method_f) + theme_bw() +
-  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
+  facet_wrap(~method,
+             labeller = label_glue('({.l}) {method} sampling ')) + 
+  theme_bw() +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.75),
         legend.background = element_rect(fill = "#ffffffaa", colour = NA),
@@
   guides(pch=guide_legend(title="Initial number\nof sample"))
-p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method_f) + theme_bw() +
-  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method_f) + theme_bw() +
-  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.6),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-grid.arrange(p1,p2,p3,nrow=3)
+
+p1
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Iterative sampling and impact on ordination.  \\label{fig:case_study_iterative_sampling}'}
+
+getpath4data <- function() {
+  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
+  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
+}
+# Read data
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+
+vrsub <- sample_iterative(vr, n=5, n2=5, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+
+sample_iterative()
+
 ```
@@
 - limitations and future work
+
+Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
 

### 2020-02-18 R/manuscript_draft.Rmd
Adding second figure for case study. Show full time series and the iterative process

@@
 \ref{fig:case_study_number_sample} (c)), where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint.
+Fig.
+\ref{fig:case_study_iterative_sampling} further illustrates the iterative sampling.
 Note that when the initial number of samples was set to 7 Fig.
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Iterative sampling and impact on ordination.  \\label{fig:case_study_iterative_sampling}'}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 6.5, fig.cap='Iterative sampling and impact on ordination.
+(a) Full time series, on which "true" changepoint was detected.
+(b) First step of iterative sampling, n = 5.
+(c) Final step of iterative samples: by adding 4 samples, the true changepoint was detected.
+\\label{fig:case_study_iterative_sampling}'}
 
@@
 
-vrsub <- sample_iterative(vr, n=5, n2=5, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+# Detect 'true' changepoint
+vrall <- sample_regular(vr, n=nrow(vr), input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+ vrall$matrix$age <- vr$AGE[vrall$matrix$index]
 
-sample_iterative()
+# Creating a dataset with all scores 
+vr_out_matrix <- read.table(paste0(getwd(),"/Output/output_iterative_sampling_varese.txt"))
+
+index_temp <- vr_out_matrix$index[vr_out_matrix$n==max(vr_out_matrix$n)]
+vr_out_order <-
+  data.frame("index" = index_temp,
+             "age"   = vrall$matrix$age[index_temp],
+             "x"     = vrall$matrix$x[index_temp],
+             "count" = summary(as.factor(vr_out_matrix$index)))
+vr_out_order$count <- abs(vr_out_order$count-max(vr_out_order$count)-1)
+
+library(wesanderson)
+mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
+
+# Plot with all data
+ p_init <- ggplot(vrall$matrix, aes(age, x)) + geom_point(alpha = .3) + geom_line(alpha = .3) +
+  theme_bw()  +
+      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
+           geom_rect(mapping=aes(xmin=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)-1], xmax=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)], ymin=0, ymax=2), alpha=0.008, fill="grey") +
+   labs(subtitle = paste0("a. Full time-series, n = ", nrow(vr)), x="Year", y="DCA first component",
+        color='samples added\nat iteration t=') +
+   geom_point(vr_out_order, mapping=aes(age, x, color = as.factor(count)), cex=2) +
+   geom_text(vr_out_order, mapping=aes(age, x, label =count), vjust = 1.5, nudge_x = 1.5) +
+   scale_color_manual(values=mypal)
+
+# Initial plot, then Adding samples
+for (i in 5:12) {
+  # Start a new loop only if the final changepoint hasn't been found at the previous iteration.  
+  if (i==5 || vrsub$matrix$index[vrsub$matrix$index==vrsub$changepoint] - vrsub$matrix$index[which(vrsub$matrix$index==vrsub$changepoint)-1]>1) {
+
+  vrsub <- sample_iterative(vr, n=i, n2=5, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+  vrsub$matrix$age <- vr$AGE[vrsub$matrix$index]
+  vrsub$matrix$count <- vr_out_order$count[vr_out_order$age %in% vrsub$matrix$age]
+
+    if (i==5) {
+      vrsub1 <- vrsub
+    p1 <- ggplot(vrsub$matrix, aes(age, x, color=as.factor(count))) +  geom_line(alpha = .4, col="grey") + geom_point() + guides(col=FALSE) +
+   scale_color_manual(values=mypal) +
+  theme_bw() +
+      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+   labs(x="Year", y="DCA first component",
+        color='samples added\nat iteration t=')
+    p2 <- p1
+  } else {
+    if (i == 6) {
+      p2 <- p1 + 
+        geom_line(vrsub$matrix, mapping=aes(age, x), alpha = .4, col="grey") + 
+        geom_point(vrsub$matrix, mapping=aes(age, x,color=as.factor(count)))   + guides(col=FALSE) 
+    } else {
+        p2 <- p2 + geom_line(vrsub$matrix, mapping=aes(age, x), alpha = .4, col="grey") + 
+          geom_point(vrsub$matrix, mapping=aes(age, x,color=as.factor(count))) + 
+          guides(col=FALSE)
+      }
+  }
+      # Save plot
+    ggsave(p2 + 
+             geom_rect(stat = "identity", mapping=aes(xmin=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)-1], xmax=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
+             annotate("text", label = paste0("Number of samples = ",i), x = min(vrsub$matrix$age), y = 2.8, colour = "black",  hjust=0,vjust=0), 
+           filename = paste0("Output/GIF_adding_sample_case_study/Iterative_",i,".pdf"), width = 6, height = 4)
+
+  }
+}
+
+p1 <- p1 + geom_rect(stat = "identity", mapping=aes(xmin=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)-1], xmax=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
+  labs(subtitle="b. Initial number of samples, n = 5")
+
+p2 <- p2 + geom_rect(stat = "identity", mapping=aes(xmin=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)-1], xmax=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
+  labs(subtitle=paste0("c. n =", nrow(vrsub$matrix)))
+
+library(patchwork)
+
+p_init / p1 / p2
 

### 2020-02-18 R/GIF_figure_iterative_sampling.R
script for figure with samples added iteratively

@@ -1,0 +1,47 @@
+# GIF of adding changepoint
+
+# load Libraries
+library(ggplot2)
+#devtools::install_github('thomasp85/gganimate', force = T)
+library(gganimate)
+
+getpath4data <- function() {
+  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
+  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
+}
+
+# Read data
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+
+# Creating a dataset with all scores 
+vr_out_matrix <-  NULL
+for (i in 5:12) {
+  vrsub <- sample_iterative(vr, n=i, n2=5, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+  if (is.null(vr_out_matrix)) vr_out_matrix <- cbind(vrsub$matrix,"n"=i) else vr_out_matrix <- rbind(vr_out_matrix,cbind(vrsub$matrix,"n"=i))
+}
+
+vr_out_matrix$age <- vr$AGE[vr_out_matrix$index]
+#write.table(vr_out_matrix,"Output/output_iterative_sampling_varese.txt", sep = "\t")
+
+
+# Plot
+p <- ggplot(vr_out_matrix, aes(index, x)) +
+  geom_point() +
+  theme_bw() +
+  # Animation part
+  transition_manual(n)
+
+# We then display it as an animation with the gg_animate function:
+gg_animate(p)
+# not working, I need to look at the error.
+
+# Error: package or namespace load failed for ‘gganimate’ in get(method, envir = home):
+#   lazy-load database '/Library/Frameworks/R.framework/Versions/3.5/Resources/library/gganimate/R/gganimate.rdb' is corrupt
+# In addition: Warning messages:
+#   1: In .registerS3method(fin[i, 1], fin[i, 2], fin[i, 3], fin[i, 4],  :
+#                             restarting interrupted promise evaluation
+#                           2: In get(method, envir = home) :
+#                             restarting interrupted promise evaluation
+#                           3: In get(method, envir = home) : internal error -3 in R_decompress1
+

### 2020-02-18 R/manuscript_draft.Rmd
edit on fig size

@@
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.75),
+  theme(legend.position = c(0.93, 0.72),
         legend.background = element_rect(fill = "#ffffffaa", colour = NA),
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 6.5, fig.cap='Iterative sampling and impact on ordination.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 5, fig.height= 7, fig.cap='Iterative sampling and impact on ordination.
 (a) Full time series, on which "true" changepoint was detected.
@@
       theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
-           geom_rect(mapping=aes(xmin=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)-1], xmax=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)], ymin=0, ymax=2), alpha=0.008, fill="grey") +
+           geom_rect(mapping=aes(xmin=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)-1], xmax=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)], ymin=-1, ymax=2), alpha=0.05, fill="grey") +
    labs(subtitle = paste0("a. Full time-series, n = ", nrow(vr)), x="Year", y="DCA first component",

### 2020-02-18 R/manuscript_draft.Rmd
edit to the second figure

@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 5, fig.height= 7, fig.cap='Iterative sampling and impact on ordination.
-(a) Full time series, on which "true" changepoint was detected.
-(b) First step of iterative sampling, n = 5.
-(c) Final step of iterative samples: by adding 4 samples, the true changepoint was detected.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 8, fig.height= 7.5, fig.cap='Iterative sampling and impact on ordination.
+(a) Initial community dataset.
+Samples are initially sampled at regular interval.
+The multivariate data is converted to univariate vector by Detrended Component Analysis.
+Samples were iteratively added, after computation of changepoint at each step.
+(b) Full time series, on which "true" changepoint was detected.
+(c) First step of iterative sampling, n = 5.
+(d) Final step of iterative samples: by adding 4 samples, the true changepoint was detected.
 \\label{fig:case_study_iterative_sampling}'}
@@
            geom_rect(mapping=aes(xmin=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)-1], xmax=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)], ymin=-1, ymax=2), alpha=0.05, fill="grey") +
-   labs(subtitle = paste0("a. Full time-series, n = ", nrow(vr)), x="Year", y="DCA first component",
+   labs(subtitle = paste0("b. Full time-series, n = ", nrow(vr)), x="Year", y="DCA first component",
         color='samples added\nat iteration t=') +
@@
 
-p1 <- p1 + geom_rect(stat = "identity", mapping=aes(xmin=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)-1], xmax=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
-  labs(subtitle="b. Initial number of samples, n = 5")
+p1 <- p1 + geom_line(vrsub1$matrix, mapping=aes(age, x), alpha = .8, col="black") + 
+          geom_point(vrsub1$matrix, mapping=aes(age, x,color=as.factor(count))) + geom_rect(stat = "identity", mapping=aes(xmin=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)-1], xmax=vrsub1$matrix$age[which(vrsub1$matrix$index==vrsub1$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
+  labs(subtitle="c. Initial number of samples, n = 5")
 
-p2 <- p2 + geom_rect(stat = "identity", mapping=aes(xmin=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)-1], xmax=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
-  labs(subtitle=paste0("c. n =", nrow(vrsub$matrix)))
+p2 <- p2 + geom_line(vrsub$matrix, mapping=aes(age, x), alpha = .8, col="black") + 
+          geom_point(vrsub$matrix, mapping=aes(age, x,color=as.factor(count))) + 
+  geom_rect(stat = "identity", mapping=aes(xmin=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)-1], xmax=vrsub$matrix$age[which(vrsub$matrix$index==vrsub$changepoint)], ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.05) +
+  labs(subtitle=paste0("d. n =", nrow(vrsub$matrix)))
 
+# Fake dataset to explain sample selection
+fakedata <- data.frame(matrix(rep(NA,74*10), nrow=74))
+colnames(fakedata) <- 1:ncol(fakedata)
+fakedata[is.na(fakedata)] <- 0
+fakedata[vr_out_order$index[vr_out_order$count==1],] <- 1 
+fakedata[vr_out_order$index[vr_out_order$count==2],] <- 2 
+fakedata[vr_out_order$index[vr_out_order$count==3],] <- 3 
+fakedata[vr_out_order$index[vr_out_order$count==4],] <- 4 
+fakedata[vr_out_order$index[vr_out_order$count==5],] <- 5 
+fakedata[,1] <- 6
+fakedata <- rbind(rep(7,ncol(fakedata)),fakedata)
+library(reshape)
+fakedata <- cbind(Year=rep(75:1,ncol(fakedata)), melt(fakedata))
+head(fakedata)
+
+p3 <- ggplot(fakedata, aes(variable,Year)) +
+  geom_tile(aes(fill = as.factor(value)), colour="black") +
+  scale_fill_manual(values=c("white",mypal,grey(.8), grey(.8))) + guides(fill=FALSE)  + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
+panel.grid.minor = element_blank(), axis.line = element_blank(), panel.background = element_blank(), axis.title=element_blank(),
+        axis.text=element_blank(),
+        axis.ticks=element_blank()) +
+  geom_text(fakedata[fakedata$Year==75,], mapping = aes(variable,Year, label=c("Year", paste("Sp.", 1:9))),  size=2.5,hjust=0,angle = 90) +
+  labs(subtitle="a. Continuous community dataset")
+
+# Assemble plot
 library(patchwork)
 
-p_init / p1 / p2
+ p3 | (p_init / p1 / p2)
 

### 2020-02-18 R/manuscript_draft.Rmd
added chunk to install packages needed and made suggestions to case study text

--- 2020-02-18 R/manuscript_draft.Rmd
@@
   \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
-  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite}
+  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
 output:
@@
 Keywords: {time series}
+
+```{r load_packages,echo=F,warning=F,message=F}
+if (!require("pacman")){install.packages("pacman");library(pacman)}
+
+pacman::p_load(patchwork,stickylabeller)
+```
 
@@
 
-An application of our study we foresee applies to paleosequences.
-Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
+\easton{As one application of our approach, we examined paleosequence data.} Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
 Samples can be taken before and after perturbation to compare community compostion (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
 
-We tested the three sampling methods on a real community time-series, from Lake Varese (IT).
+\easton{In line with the simulation results, we tested different sampling methods} on a real community time-series, from Lake Varese (IT).
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
@@
 
-Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period (@bruel_seeking_2018).
-We defined the "true" changepoint as the one identified from the complete record, in 1983/1988.
-Changepoints were detected on first component computed from the data sub-sampled.
-We used the same ordination method as in Bruel et al (2018), but note that DCA is best used to conserve multiple gradients.
+Previously, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period [@bruel_seeking_2018].
+We detected changepoints on the first component computed from the assemblage data.
+We used the same ordination method as in Bruel et al (2018), but note that Detrended Correspondence Analysis (DCA) is best used to conserve multiple gradients (\easton{Rosalie: can you explain this a bit more, I'm not sure what it means}).
 As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988).
-We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).
+We defined these as the "true" changepoints given they came from an analysis of the complete data.
+The 1983-1988 changepoint was the largest in magnitude.
+(\easton{Rosalie: does this mean that the changepoint occured sometime between 1983 and 1988}).
+<!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->
 
-We used the full record (74 observation) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
+We used the full record (74 observations) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
 Iterative sampling final number could be lower than maximum number if both samples framing the shift were analyzed.

### 2020-02-21 R/manuscript_draft.Rmd
added draft of abstract and checked for spelling mistakes

@@
 
-Title: Sampling requirements to detect ecosystem change
+Title: Sampling requirements to detect ecosystem shifts
 
@@
 
-Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA $^{2}$Department of Biology, University of Vermont, VT 05405, USA
+Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA $^{2}$Department of Biology, University of Vermont, VT 05405, USA https://orcid.org/0000-0002-0768-9555
 
@@
 
-Abstract here
+(150-250 words)
+
+Environmental monitoring is a key piece of managing ecosystems.
+Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective.
+In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed.
+For example, soil or ice cores are taken in the field and later processed in a laboratory setting.
+Standard approaches either involve either equally-spaced or random sampling throughout each core.
+We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change.
+We test these ideas with a case study of paleoecological plankton data extracted from lake sediment cores.
+We demonstrate that standard approaches to sample processing are inefficient and costly.
+We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
+We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools.
 
@@
 Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation).
-Monitoring programs also need to be designed to be cost-effective given limited bugets (citations).
+Monitoring programs also need to be designed to be cost-effective given limited budgets (citations).
 
-Methods to dectect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
+Methods to detect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
 
@@
 Should the video be assessed at every minute?
-As long as processsing samples is expensive, these trade-offs will remain.
+As long as processing samples is expensive, these trade-offs will remain.
 
@@
 \ref{fig:conceptual}c).
-In the context of sediment cores, this would mean analzying community composition as random spots along the core.
+In the context of sediment cores, this would mean analyzing community composition as random spots along the core.
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
@@
 \ref{fig:conceptual}d).
-Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
+Lastly, iterative sampling involves first taking a set number of samples and then alliteratively adding samples until a pre-determined level of precision is achieved (Fig.
 \ref{fig:conceptual}e-g).
@@
 \easton{As one application of our approach, we examined paleosequence data.} Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
-Samples can be taken before and after perturbation to compare community compostion (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
+Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
 
@@
 The 1983-1988 changepoint was the largest in magnitude.
-(\easton{Rosalie: does this mean that the changepoint occured sometime between 1983 and 1988}).
+(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988}).
 <!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->
@@
 The initial number of samples should be chosen depending on the objective of the study.
-If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoritical results?).
-If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time serie to have a good initial ranking.
+If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoretical results?).
+If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.
 

### 2020-02-25 R/manuscript_draft.Rmd
added new figure and paragraph in discussino

@@
 Similarly, environmental samples (e.g.
-water, soil) can be saved with eDNA processed later (citation).
+water, soil) can be saved and processed later for composition, including eDNA (citation).
 Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
@@
 However, our approaches are widely applicable to other questions and data types.
-We first investigate these tools using a simulated-based approach to As a case study, we apply these tools to the detection of change points for a set of lakes.
+We first investigate these tools using a simulated-based approach.
+As a case study, we apply these tools to the detection of change points for a set of lakes.
 (more from Rosalie here).
@@
 
-# Simulation approach
+# Theoretical approach
 
@@
 
-## Theoretical model
+## Simulation model
 
@@
 
-```{r theoretical_results_min_time,message=F,warning=F,echo=F,eval=F}
+```{r simulation_min_time,message=F,warning=F,echo=F,eval=F}
 
@@
 
-```{r theoretical_results_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
+```{r simulation_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
 The default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
-\\label{fig:theoretical_min_time}',message=F,warning=F,echo=F,eval=T}
+\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
 
@@
 
-```{r theoretical_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
+```{r simulation_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
 
@@
 source('R/sampling_random.R')
+source('R/sampling_iterative.R')
 source('R/build_time_series.R')
@@
 
-approach <- c('random')#,'regular','iterative')
-number_samples <- 5:30
-simulation_number <- 1:50
+approach <- c('Random','Regular','Iterative')
+number_samples <- c(seq(5,20,2),25,30,35)
+simulation_number <- 1:20
 output_theory_approach <- data.frame(expand.grid(approach,number_samples,simulation_number))
@@
   output_theory_approach$distance_to_real[i] <- abs(params$shift_time[output_theory_approach$simulation_number[i]] - calc_distance_to_real(output_theory_approach$approach[i],output_theory_approach$number_samples[i],time_series[[output_theory_approach$simulation_number[i]]]))
-  #print(i)
+  print(i)
 }
@@
 write.csv(x = output_theory_approach,file = 'Output/theoretical_comparing_approaches.csv',quote = FALSE )
+```
+
+```{r simulation_approaches_graph, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_number_sample}'}
+
+myoutput <- read.csv(file='Output/theoretical_comparing_approaches.csv',header=T)
+myoutput$approach <- factor(myoutput$approach,levels = c('Random', 'Regular', 'Iterative'))
+require(dplyr)
+myoutput_summary <- myoutput %>%
+  group_by(approach,number_samples) %>%
+  summarize(mean_distance = mean(distance_to_real),sd_distance=sd(distance_to_real))
+
+library(ggplot2)
+#devtools::install_github("rensa/stickylabeller")
+library(stickylabeller)
+p1  <- ggplot(myoutput_summary, aes(x=number_samples,y=mean_distance)) + geom_line()  + geom_ribbon(aes(ymin = mean_distance - 1.96*sd_distance, ymax = mean_distance + 1.96*sd_distance), fill = "grey80") +geom_line(size=1) + coord_cartesian(ylim = c(2.2, 50))  +
+  facet_wrap(~approach,
+             labeller = label_glue('({.l}) {approach} sampling ')) + 
+  theme_bw() +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") #+
+ # theme(legend.position = c(0.93, 0.72),
+#       legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+#        legend.title=element_text(size=7), 
+ #       legend.text=element_text(size=8)) +
+#  guides(pch=guide_legend(title="Initial number\nof sample"))
+
+p1
+
 ```
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='WHY ARE THERE NEGATIVE VALUES HERE?
+Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
@@
 
-\clearpage
+Our approach goes beyond just paleoecological analyses.
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive (citations).
+The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
+Our specific approach in this paper was on data where more samples can be added long after the dynamics occurred.
+Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
+Suppose instead that the goal was to detect a change in observed individuals with video-based approaches.
+It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address our question of interest.
+Using an interval sampling approach, one could take a fixed number of samples to start.
+The trend over time from simple linear regression could be taken.
+Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates.
+If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample.
+Sampling would continue until the trend estimate began to settle down.
+Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
+These approaches are also applicable to other questions related to the optimal sample size and approach in determining long-term trends, biodiversity estimates, and more (citations).
 
-Limitations:
+<!--Limitations paragraph-->
+
+Although our approach is flexible to be applied to various data sources and questions, this is also one of its limitations.
+The specific number of samples and optimal sampling approach will largely be system specific.
+A similar system may be used to get some sense for the best sampling scheme, but it will depend on the VARIOUS FACTORS HERE.
+Similarly, selecting a set number of samples or specific approach may also limit what future questions can be asked.
+However, more samples could be added to address these additional questions.
 

### 2020-02-26 R/manuscript_draft.Rmd
changed the function to check the presence of package so it can be used for any future package. The function is called pkgTest()

@@
 ```{r load_packages,echo=F,warning=F,message=F}
-if (!require("pacman")){install.packages("pacman");library(pacman)}
+pkgTest <- function(x)
+{
+  if (!require(x,character.only = TRUE))
+  {
+    install.packages(x,dep=TRUE)
+    if(!require(x,character.only = TRUE)) stop("Package not found")
+  }
+  library(x, character.only = TRUE)
+}
+
+pkgTest("pacman")
 

### 2020-02-26 R/manuscript_draft.Rmd
apologies, just realized that checking the status and availability of packages is what the package pacman was doing

--- 2020-02-25 R/manuscript_draft.Rmd
@@
   \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
-  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
+  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{darkseagreen4}{#1}}
 output:
@@
 ```{r load_packages,echo=F,warning=F,message=F}
-if (!require("pacman")){install.packages("pacman");library(pacman)}
+if (!require("pacman",character.only = TRUE))
+  {
+    install.packages("pacman",dep=TRUE)
+    if(!require("pacman",character.only = TRUE)) stop("Package not found")
+  }
+library("pacman")
 
@@
 The 1983-1988 changepoint was the largest in magnitude.
-(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988}).
+(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988} \rosalie{The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988.
+That is just a way to see it, but let me know if it is confusing.
+I reviewed a paper and asked the authors to change the changepoint to this bracket, and they replied telling me they didn't understand my comment, so maybe it is not worth the confusion!
+Put another way, if we were recording the temperature every hour, and it was 6°C at 1, 2, 3, 4 am, then 8°C at 5, 6, 7, 8am, the method would detect the changepoint at 5am.
+But in fact, the temperature could have very well changed anytime between 4.01am and 5am.
+So I would say: "temperature changed between 4 and 5 am.".
+Am I maling any sense?}).
 <!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->

### 2020-02-26 R/manuscript_draft.Rmd
remove library commands throughout the manuscript and get everything under the command pacman

@@
   \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
-  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{darkseagreen4}{#1}}
+  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
 output:
@@
 
-pacman::p_load(patchwork,stickylabeller)
+# Keeping below source for github package. Ask Easton whether pacman works for github packages or not.
+#devtools::install_github("rensa/stickylabeller")
+
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape)
 ```
@@
 myoutput$approach <- factor(myoutput$approach,levels = c('Random', 'Regular', 'Iterative'))
-require(dplyr)
 myoutput_summary <- myoutput %>%
@@
 
-library(ggplot2)
-#devtools::install_github("rensa/stickylabeller")
-library(stickylabeller)
 p1  <- ggplot(myoutput_summary, aes(x=number_samples,y=mean_distance)) + geom_line()  + geom_ribbon(aes(ymin = mean_distance - 1.96*sd_distance, ymax = mean_distance + 1.96*sd_distance), fill = "grey80") +geom_line(size=1) + coord_cartesian(ylim = c(2.2, 50))  +
@@
 \\label{fig:case_study_number_sample}'}
-library(ggplot2)
-#devtools::install_github("rensa/stickylabeller")
-library(stickylabeller)
 p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
@@
 
-library(wesanderson)
 mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
@@
 fakedata <- rbind(rep(7,ncol(fakedata)),fakedata)
-library(reshape)
 fakedata <- cbind(Year=rep(75:1,ncol(fakedata)), melt(fakedata))
@@
 # Assemble plot
-library(patchwork)
 

### 2020-02-26 R/manuscript_draft.Rmd
add some text

@@
 \vspace{5 mm}
-Keywords: {time-series} \vspace{10 mm}
+Keywords: {time-series} {changepoint} \vspace{10 mm}
 
@@
 water, soil) can be saved and processed later for composition, including eDNA (citation).
-Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
+Likewise, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
@@
 Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We focus on paleoecological core samples as one example of this type of data.
+We tailor our analysis to the detection of changepoint.
+Changepoint are an important characteristic of a time serie We focus on paleoecological core samples as one example of this type of data.
 We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint.
@@
 We first investigate these tools using a simulated-based approach.
-As a case study, we apply these tools to the detection of change points for a set of lakes.
-(more from Rosalie here).
-
+We then test the tools on a case study from a paleosequence.
 # Sampling approaches and changepoint detection
@@
 The 1983-1988 changepoint was the largest in magnitude.
-(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988} \rosalie{The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988.
+(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988.} \rosalie{Answer: The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988.
 That is just a way to see it, but let me know if it is confusing.
@@
 
+Many ecological processes are non linear (e.g., D'Amario et al 2019, Scientific reports).
+Detecting a changepoint on one aspect of the ecosystem does not guarantee that the sampling resolution is appropriate to detect the driver for the changepoint.
+
 Concluding paragraph:

### 2020-02-27 R/manuscript_draft.Rmd
contribution to writing

@@
 Title: Sampling requirements to detect ecosystem shifts
+(lame title) Getting closer to the point: sampling requirements to detect ecosystem shifts.
+Recommendation for temporal sampling
 
@@
 In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed.
-For example, soil or ice cores are taken in the field and later processed in a laboratory setting.
-Standard approaches either involve either equally-spaced or random sampling throughout each core.
+For example, sediment or ice cores can be taken in the field in less than an hour, but will be processed in a laboratory setting for anytime between a week and several months, depending on how much information is extracted.
+Sub-sampling allows to reduce the processing time while still giving precious insight.
+Standard sub-sampling approaches often involve equally-spaced sampling.
 We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change.
-We test these ideas with a case study of paleoecological plankton data extracted from lake sediment cores.
-We demonstrate that standard approaches to sample processing are inefficient and costly.
+We test these ideas with a case study of Cladocera community assemblage, which trajectory was reconstructed from a sediment core.
+We demonstrate that standard approaches to sample processing are inefficient, hence costly.
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
-We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools.
+We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.
 
@@
   }
-library("pacman")
 
@@
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010).
-However, environmental monitoring is still often expensive and time-consuming (citations).
-Therefore, monitoring programs need to be designed in such a way to address the question of interest.
+However, environmental monitoring is still often expensive and time-consuming, especially when further analyses are needed following sample collection (citations).
+In Ecology and Genetics especially, there has always been a disparity between the amount of data that can be aquired and stored, and the processing methods that could allow to extract the most information from that data.
+\rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
+With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at.
+If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}
+
+Therefore, monitoring programs need to be designed in such a way to address the question of interest, without wasting resources.
 Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation).
@@
 
+<!-- 
 Methods to detect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
@@
 - Historical data and ability to choose how much to sample
+-->
 
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie).
+For example, the field of paleolimnology paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie).
 Similarly, environmental samples (e.g.
@@
 \ref{fig:conceptual}c).
-In the context of sediment cores, this would mean analyzing community composition as random spots along the core.
+In the context of sediment cores, this would mean analyzing community composition as random location along the core.
+Random sampling is recommanded in designs aiming at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018).
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.

### 2020-02-27 R/manuscript_draft.Rmd
added comments to figure and discussion

--- 2020-02-26 R/manuscript_draft.Rmd
@@
 We began with a theoretical exploration of the sampling requirements to detect a changepoint.
-We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable that represents either population size, biodiversity, or some metric of community composition.
+We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable ($X_t$) that represents either population size, biodiversity, or some metric of community composition at time $t$.
 The model includes temporal autocorrelation ($\phi$), the mean of the process ($\mu_X$), and a white noise term ($w_t$).
-The white noise term is a normal distribution with mean ($\mu_w$) and variance ($\sigma$).
+The white noise term is a normal distribution with mean ($\mu_w$) and variance ($\sigma^2$):
 
@@
 X_t & = & \mu_X + \phi (X_{t-1} - \mu_X) + w_t \\ 
-w_t & \sim & Norm(\mu_w,\sigma)
+w_t & \sim & Norm(\mu_w,\sigma^2).
 \end{array}
@@
 
-```{r theoretical_results,message=F,warning=F,echo=F,eval=F}
+```{r theoretical_results_old,message=F,warning=F,echo=F,eval=F}
 
@@
 ```{r simulation_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
-The default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+Random sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.

### 2020-03-04 R/manuscript_draft.Rmd
added citations and discussion text

@@
 
-Title: Sampling requirements to detect ecosystem shifts
+Title: Sampling requirements and approaches to detect ecosystem shifts
 (lame title) Getting closer to the point: sampling requirements to detect ecosystem shifts.
@@
 \vspace{5 mm}
-Keywords: {time-series} {changepoint} \vspace{10 mm}
+Keywords: time-series, changepoint \vspace{10 mm}
 
@@
 
-Number of references:
+Number of references: \total{citenum}
 
@@
 
-Submission as:
+Submission as: Article at \emph{Ecological Applications}
 
@@
 \begin{center}
-\textbf{\Large Title here}
+\textbf{\Large Sampling requirements and approaches to detect ecosystem shifts}
 \vspace{7 mm}
@@
 We test these ideas with a case study of Cladocera community assemblage, which trajectory was reconstructed from a sediment core.
-We demonstrate that standard approaches to sample processing are inefficient, hence costly.
+We demonstrate that standard approaches to sample processing are inefficient.
+For our case study, using an optimal sampling approach would have resulted in savings of XX human-hours or thousands of dollars in labor costs.
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
@@
 
-Environmental monitoring is one of the core components to modern research and management (citations).
-Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (Lovett et al, 2007).
-Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010).
+Environmental monitoring is one of the core components to modern research and management [@White2019d].
+Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
+Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a].
 However, environmental monitoring is still often expensive and time-consuming, especially when further analyses are needed following sample collection (citations).
@@
 
-Therefore, monitoring programs need to be designed in such a way to address the question of interest, without wasting resources.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation).
-Monitoring programs also need to be designed to be cost-effective given limited budgets (citations).
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@McDonald-Madden2010a,@White2019d,@White2020,@Lindenmayer2020a].
+Monitoring programs also need to be designed to be cost-effective given limited budgets [@Caughlan2001a,@Grantham2008b,@Bennett2016a].
 
@@
 The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
-For example, @White2019 found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
-Other work has focused on the frequency of monitoring (citations), the allocation of resources spatially versus temporally (citations), and the ability to use citizen science to increase monitoring (citations).
-Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection (citation).
+For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
+The specific number of years required depended on the species biology and the detection method used [@White2019d].
+Other work has focused on the frequency of monitoring [@Wauchope2019a], the allocation of resources spatially versus temporally [@Rhodes2011a,@Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
+Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
 
@@
 Similarly, environmental samples (e.g.
-water, soil) can be saved and processed later for composition, including eDNA (citation).
-Likewise, video-graphical approaches can record snapshots of a system and be analyzed later (citations).
+water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a].
+Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later (citations).
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
 Should the paleoecological core be analyzed at every centimeter?
-Should the video be assessed at every minute?
+Should the video be assessed once per minute?
 As long as processing samples is expensive, these trade-offs will remain.
@@
 Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We tailor our analysis to the detection of changepoint.
-Changepoint are an important characteristic of a time serie We focus on paleoecological core samples as one example of this type of data.
+We tailor our analysis to the detection of changepoint in a time series.
+Changepoints are an important characteristic of a time series.
+We focus on paleoecological core samples as one example of this type of data.
 We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint.
@@
 We then test the tools on a case study from a paleosequence.
+
 # Sampling approaches and changepoint detection
@@
 Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
-\ref{fig:theoretical_min_time}).
+\ref{fig:simulation_min_time}).
 
@@
 
-\easton{As one application of our approach, we examined paleosequence data.} Sediment cores are taken from aquatic environment to reconstruct past ecosystems, in order to extend our perspective on ecosystem trajectories.
-Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @perga_high-resolution_2015).
+As one application of our approach, we examined paleosequence data.
+To obtain this data sediment cores are taken from aquatic environments to reconstruct past ecosystems.
+Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @Perga2015).
 
-\easton{In line with the simulation results, we tested different sampling methods} on a real community time-series, from Lake Varese (IT).
+We tested different sampling methods on a real community time-series, from Lake Varese (IT).
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
 It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
-Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Crosta 1999, @bruel_seeking_2018).
-Air temperature is now driving changes in plankton communities (@bruel_seeking_2018).
+Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999,@Bruel2018].
+Air temperature is now driving changes in plankton communities (@Bruel2018).
 
-Previously, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core (VAR10-10) covering the 1816(±26)—2010 period [@bruel_seeking_2018].
-We detected changepoints on the first component computed from the assemblage data.
-We used the same ordination method as in Bruel et al (2018), but note that Detrended Correspondence Analysis (DCA) is best used to conserve multiple gradients (\easton{Rosalie: can you explain this a bit more, I'm not sure what it means}).
-As a result, several changepoints were detected on the first component (in 1926/1928, 1946/1948 and 1983/1988).
+In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
+We used the same ordination method as Bruel et al (2018) to reduce the dimensionality of the assemblage-level data.
+As a result, we detected several changepoints on the first component (in 1926/1928, 1946/1948 and 1983/1988).
 We defined these as the "true" changepoints given they came from an analysis of the complete data.
 The 1983-1988 changepoint was the largest in magnitude.
-(\easton{Rosalie: does this mean that the changepoint occurred sometime between 1983 and 1988.} \rosalie{Answer: The changepoint was detected in 1988, but in fact, it could have taken place anytime between 1983 and 1988.
-That is just a way to see it, but let me know if it is confusing.
-I reviewed a paper and asked the authors to change the changepoint to this bracket, and they replied telling me they didn't understand my comment, so maybe it is not worth the confusion!
-Put another way, if we were recording the temperature every hour, and it was 6°C at 1, 2, 3, 4 am, then 8°C at 5, 6, 7, 8am, the method would detect the changepoint at 5am.
-But in fact, the temperature could have very well changed anytime between 4.01am and 5am.
-So I would say: "temperature changed between 4 and 5 am.".
-Am I maling any sense?}).
-<!--We conducted the analysis on the second component as well, where only one changepoint was detected (in 1944/1946).-->
 
-We used the full record (74 observations) and subsampled following the three strategies described earlier (random, regular, iterative), varrying the maximum number analyzed from 5 to 15.
-Iterative sampling final number could be lower than maximum number if both samples framing the shift were analyzed.
+In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative).
+For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.
 
@@
 
-Random sampling perform the worst, as changepoint analysis was left to chance.
-Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint was left  to chance.
-Iterative sampling (Fig.
-\ref{fig:case_study_number_sample} (c)), where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint.
-Fig.
-\ref{fig:case_study_iterative_sampling} further illustrates the iterative sampling.
-Note that when the initial number of samples was set to 7 Fig.
+We found that random sampling performed the worst, as changepoint analysis was left to chance.
+Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint.
+Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
+\ref{fig:case_study_number_sample} (c)).
+We show how iterative sampling changes the ordination, and the resulting changepoint, as more samples are added (Fig.
+\ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig.
 \ref{fig:case_study_number_sample} (c), square markers), the distance to the real changepoint was initially very high.
-Without any prior, the initial number sampled fell far away from the true changepoint.
+Without any prior, the initial number sampled fell far away from the true changepoint.-->
 
-We tested how the three methods performed at detecting other changepoints (as three changepoints were detected in the initial study, @bruel_seeking_2018).
-Iterative sampling performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials \textcolor{red}{2}).
-The initial number of samples should be chosen depending on the objective of the study.
+We tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
+Iterative sampling still performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials \textcolor{red}{2}).
+<!--This could probably be moved to the discussion: The initial number of samples should be chosen depending on the objective of the study.
 If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoretical results?).
-If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.
+If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.-->
 
@@
 
-__Main point__: Changepoint was detected even though we used a transformed vector (from community to first component of DCA)
-
 # General Discussion
@@
 Shouldn't be too hard to do__
+
+@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
+Each sample took an average of XX hours to process.
+We found that using an interval approach would have eliminated XX hours of sample processing or thousands of US dollars depending on labor costs.
 
@@
 About recommendation to reduce the frequency of data collected as part of monitoring program:  
-While long-term monitoring program are often expensive, these cost are nothing compared to the value of the resources monitored and the policies affecting these resources (@lovett_who_2007).
+While long-term monitoring program are often expensive, these cost are nothing compared to the value of the resources monitored and the policies affecting these resources (@Lovett2007).
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may indeed not be necessary to manage ecosystems in the future.

### 2020-03-04 R/manuscript_draft.Rmd
Work on intro and method

@@
 Title: Sampling requirements and approaches to detect ecosystem shifts
-(lame title) Getting closer to the point: sampling requirements to detect ecosystem shifts.
-Recommendation for temporal sampling
 
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, the field of paleolimnology paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie).
+For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions (Cohen, 2003).
 Similarly, environmental samples (e.g.
@@
 Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We tailor our analysis to the detection of changepoint in a time series.
+We tailor our analysis to the detection of changepoint in a time series, were each point had a certain degree of autocorrelation with the previous, until a changepoint come.
 Changepoints are an important characteristic of a time series.
@@
 \ref{fig:conceptual}e-g).
+In each scenarios, we sample the first and last sample, in case the changepoint takes place early or late in the period covered by the sampling.
 
-Rosalie: Can you write a paragraph here detailing the changepoint detection process?
+We detect changepoints with the function _e.divisive()_ in the R package _ecp_ (James et al 2019).
+There are several methods available to detect changepoint (reviewed in James and Matteson, 2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
+We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of change points.
+Herein, we focus on detecting the most important changepoint, although we tested the method on a time-serie that would have multiple changepoints (Supplementary Materials 1).
+
+We wrote three functions:  
+  * sample_random(): sample first and last sample, in addition of n-2 random samples in between (n = maximum number of sample, chosen by the user),
+  * sample_regular(): sample first and last sample, in addition of n-2 evenly spaced samples in between (n = maximum number of samples, chosen by the user),
+  * sample_regular(): sample first and last sample as well as 3 evenly spaced samples in between (3 is the default but can be modified), in order to initiate the changepoint detection.
+Upon detection of the changepoint on this 5 samples time series, a new sample is added between the changepoint and the previous sample, to narrow down the real changepoint.
+Detection of changepoint / addition of sample is repeated, until it was narrowed down to two consecutive samples, or n (maximum number of samples, chosen by the user) was reached, whichever comes first.
+
+In order to test the performance, we detected the "real" changepoint on the whole time-serie, and compare the changepoint found on the sub-sample with the "real" one.
+The distance to real changepoint served as performance diagnostic.
 

### 2020-03-05 R/manuscript_draft.Rmd
edit figures and remove cases with several n2 init for the case study

@@
   theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") #+
@@
 We found that random sampling performed the worst, as changepoint analysis was left to chance.
-Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint.
+Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
 Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
@@
 
-```{r case study read data output, include=FALSE}
-myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
-myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
-
-```
-
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='WHY ARE THERE NEGATIVE VALUES HERE?
@@
 \\label{fig:case_study_number_sample}'}
-p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+# Read data
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Change to TRUE if we decide it is worth it to show several initial number of n2. It looks more confusing -- but it would bring more information
+several_n2_init = F
+
+# Plot
+p1 <- if(several_n2_init) {
+  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
   geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
@@
   theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
@@
   guides(pch=guide_legend(title="Initial number\nof sample"))
+} else {
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('({.l}) {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") 
+}
 

### 2020-03-05 R/manuscript_draft.Rmd
Add some discussion text Mostly modified figure 2 by adding the case study time series sigma, phi, and shift size parameters.

@@
 
-```{r simulation_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
+```{r read case study data and calculate some parameters to plot them on theoritical plot, message=FALSE, warning=FALSE, include=FALSE}
+getpath4data <- function() {
+  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
+  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
+}
+# Read data
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+
+# Detect 'true' changepoint
+vrall <- sample_regular(vr, n=nrow(vr), input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+vrall$matrix$age <- vr$AGE[vrall$matrix$index]
+
+# Calculate the 3 time-series parameter needed for comparison to the theoritical results
+
+# 1 - Calculate sigma
+sd_vr <- sd(vrall$matrix$x)
+
+# 2 - Calculate autocorrelation of real time serie
+ar_vr <- acf(vrall$matrix$x, lag.max = 5, plot = FALSE)
+
+# 3 - Calculate shift size
+shift_vr <- vrall$matrix$x
+# Normalize
+shift_vr <- (shift_vr-min(shift_vr))/(max(shift_vr)-min(shift_vr))
+# Difference between the two values before/after shift
+shift_vr <- abs(shift_vr[vrall$changepoint]-shift_vr[vrall$changepoint-1])
+
+```
+
+\rosalie{I wonder whether we should show the results with random sampling for Fig.
+\ref{fig:simulation_min_time}: we show later that it is the worst case scenario.
+But maybe that's a discussion we already had and I am missing on something.
+
+I calculated the autocorrelation at lag-1 for the case_study time series and add it to the plot here.}
+
+```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
 Random sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
+Vertical line indicate respective parameter calculated from the case study time series.
 \\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
@@
 sigma <- read.csv('Output/theoretical_results_sigma.csv',header=TRUE)
-plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+#plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+
+p1 <- ggplot(sigma, aes(sigma_vec, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept = ar_vr$acf[2], lty=2) +
+  ylim(20,50) +
+  labs(x="sigma", y= "Minimum number of samples required") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 
@@
 phi <- read.csv('Output/theoretical_results_phi.csv',header=TRUE)
-plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+#plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+
+p2 <- ggplot(phi, aes(phi_vec, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept = sd_vr, lty=2) +
+  ylim(20,50) +
+  labs(x="phi", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 
@@
 shift_size <- read.csv('Output/theoretical_results_shift_size.csv',header=TRUE)
-plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+#plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+
+p3 <- ggplot(shift_size, aes(shift_size_vec, min_time)) +
+  geom_point() +
+  geom_line() + 
+  geom_vline(xintercept = shift_vr, lty=2) +
+  ylim(20,50) +
+  labs(x="phi", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+(p1 | p2 | p3)
 
@@
 
-getpath4data <- function() {
-  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
-  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
-  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
-}
-# Read data
-vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
-
-# Detect 'true' changepoint
-vrall <- sample_regular(vr, n=nrow(vr), input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
- vrall$matrix$age <- vr$AGE[vrall$matrix$index]
-
 # Creating a dataset with all scores 
@@
 
+Analyzing a subsample of the whole time-series inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
+However, in some cases, analyzing the whole sample collected is not an option.
+We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
+
 Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
@@
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
-Each sample took an average of XX hours to process.
-We found that using an interval approach would have eliminated XX hours of sample processing or thousands of US dollars depending on labor costs.
+Each sample took an average of 3 hours to process.
+We found that using an interval approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
+This correspond to several thousands of US dollars depending on labor costs.
+
+Note that the number of sample needed varies depending on the characteristics of the time series.
+For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example is `r round(ar_vr$acf[2],3)`
 

### 2020-03-05 R/manuscript_draft.Rmd
more edits to figure 2

@@
 
-\rosalie{I wonder whether we should show the results with random sampling for Fig.
-\ref{fig:simulation_min_time}: we show later that it is the worst case scenario.
+\rosalie{I wonder whether we should show the results with random sampling for the figure showing the minimum number sampled required for different sigma, phi, and shift size: we show later that it is the worst case scenario.
 But maybe that's a discussion we already had and I am missing on something.
-
-I calculated the autocorrelation at lag-1 for the case_study time series and add it to the plot here.}
+<br> I calculated the autocorrelation at lag-1 for the case study time series and added it to the plot here.}
 
@@
 
-p1 <- ggplot(sigma, aes(sigma_vec, min_time)) +
-  geom_point() +
-  geom_line() +
-  geom_vline(xintercept = ar_vr$acf[2], lty=2) +
-  ylim(20,50) +
-  labs(x="sigma", y= "Minimum number of samples required") + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-
 # Phi results
@@
 #plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
-
-p2 <- ggplot(phi, aes(phi_vec, min_time)) +
-  geom_point() +
-  geom_line() +
-  geom_vline(xintercept = sd_vr, lty=2) +
-  ylim(20,50) +
-  labs(x="phi", y= "") + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 
@@
 
-p3 <- ggplot(shift_size, aes(shift_size_vec, min_time)) +
+# Melt the 3 dataset to be able to use facet_wrap()
+colnames(sigma) <- colnames(phi) <- colnames(shift_size) <- c("X", "Param", "min_time")
+simulation_all <- 
+  rbind(cbind("Parameter" = rep("Standard deviation", nrow(sigma)), sigma),
+      cbind("Parameter" = rep("Autocorrelation at lag-1", nrow(phi)), phi),
+      cbind("Parameter" = rep("Normalized amplitude of shift", nrow(shift_size)), shift_size))
+# param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
+#                        "value" = c(ar_vr$acf[2], sd_vr, shift_vr))
+
+p1 <- 
+ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
   geom_point() +
   geom_line() +
-  geom_vline(xintercept = shift_vr, lty=2) +
+  geom_vline(xintercept=ar_vr$acf[2], lty=2) + 
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('(a) {Parameter}')) +
+  ylim(20,50) +
+  labs(x=c("phi"), y= "Minimum number of samples required") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept = sd_vr, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('(b) {Parameter}')) +
   ylim(20,50) +
@@
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
+  geom_point() +
+  geom_line() + 
+  geom_vline(xintercept = shift_vr, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('(c) {Parameter}')) +
+  ylim(20,50) +
+  labs(x="shift size", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 
-(p1 | p2 | p3)
+# Remove the y-axis for the 2nd plot - p2 then merge 2 plots
+cowplot::plot_grid(p1 + theme(plot.margin = unit(c(0.15, 0.1, 0.15, 0.15), "cm")), 
+                   p2 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   p3 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   nrow = 1,
+                   rel_widths = c(1.2, 1,1),
+                   align = 'h', axis = 'tb')
 

### 2020-03-05 R/manuscript_draft.Rmd
add package cowplot to the pacman loading package line. Correct typo on fig. 2

@@
 
-pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape)
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot)
 ```
@@
   ylim(20,50) +
-  labs(x=c("phi"), y= "Minimum number of samples required") + 
+  labs(x=c("sigma"), y= "Minimum number of samples required") + 
   theme_bw() +

### 2020-03-05 R/manuscript_draft.Rmd
edit the style of facet_wrap title so they match base r and figures look more homogeneous

@@
   facet_wrap(~Parameter, scales="free_x", 
-             labeller = label_glue('(a) {Parameter}')) +
+             labeller = label_glue('a. {Parameter}')) +
   ylim(20,50) +
-  labs(x=c("sigma"), y= "Minimum number of samples required") + 
+  labs(x=c("phi"), y= "Minimum number of samples required") + 
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
 p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
@@
   facet_wrap(~Parameter, scales="free_x", 
-             labeller = label_glue('(b) {Parameter}')) +
+             labeller = label_glue('b. {Parameter}')) +
   ylim(20,50) +
@@
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
 p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
@@
   facet_wrap(~Parameter, scales="free_x", 
-             labeller = label_glue('(c) {Parameter}')) +
+             labeller = label_glue('c. {Parameter}')) +
   ylim(20,50) +
@@
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
 
@@
   facet_wrap(~approach,
-             labeller = label_glue('({.l}) {approach} sampling ')) + 
+             labeller = label_glue('{.l}. {approach} sampling ')) + 
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") #+
@@
   facet_wrap(~method,
-             labeller = label_glue('({.l}) {method} sampling ')) + 
+             labeller = label_glue('{.l}. {method} sampling ')) + 
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
@@
   facet_wrap(~method,
-             labeller = label_glue('({.l}) {method} sampling ')) + 
+             labeller = label_glue('{.l}. {method} sampling ')) + 
   theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
   labs(x="Total number of samples analyzed", y="Distance to real changepoint") 

### 2020-03-05 R/manuscript_draft.Rmd
add some discussion.

@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='WHY ARE THERE NEGATIVE VALUES HERE?
-Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
@@
 fakedata <- cbind(Year=rep(75:1,ncol(fakedata)), melt(fakedata))
-head(fakedata)
+#head(fakedata)
 
@@
         axis.text=element_blank(),
-        axis.ticks=element_blank()) +
-  geom_text(fakedata[fakedata$Year==75,], mapping = aes(variable,Year, label=c("Year", paste("Sp.", 1:9))),  size=2.5,hjust=0,angle = 90) +
+        axis.ticks=element_blank(),
+plot.subtitle=element_text(size=11)) +
+  geom_text(fakedata[fakedata$Year==75,], mapping = aes(variable,Year, label=c("Year", paste("Sp.", c(1:8,"n")))),  size=2.5,hjust=0,angle = 90) +
   labs(subtitle="a. Continuous community dataset")
@@
 
-Analyzing a subsample of the whole time-series inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
-However, in some cases, analyzing the whole sample collected is not an option.
+Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
+For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
+However, in some cases, analyzing the whole samples collected is not an option, because there is not enough processing time or money available.
 We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
 
-Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
-
-Case study: With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
+With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
 Same results for the major change, with only 12% of the samples analyzed.
-However, need more samples to detect further changes.
-__Note that it would be possible to edit the function to go ahead and detect next changes once the main one has been detected.
-Is it worth coding?
-Shouldn't be too hard to do__
-
+The method also work to detect further shifts (Supplementary Materials 1).
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
@@
 
+Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
+
 Note that the number of sample needed varies depending on the characteristics of the time series.
-For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example is `r round(ar_vr$acf[2],3)`
+For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example is `r round(ar_vr$acf[2],3)`, the variance is `r round(sd_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
 

### 2020-03-06 R/manuscript_draft.Rmd
Add a discussion sentence

@@
 
-Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA $^{2}$Department of Biology, University of Vermont, VT 05405, USA https://orcid.org/0000-0002-0768-9555
+Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
 
@@
 
-With the iterative method, we found the main changepoint with maximum 9 samples analyzed, with a starting point of 5 randomly distributed samples.
+Multiple subsampling strategies can be chosen(Fig.
+\ref{fig:conceptual}), but only iterative sampling allowed to approach the real changepoint in a limited number of samples (Fig.
+\ref{fig:simulation_min_time}c).
+Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
+In the context of a real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (\ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
 Same results for the major change, with only 12% of the samples analyzed.

### 2020-03-16 R/manuscript_draft.Rmd
added a few notes on next steps

@@
 
-- writing intro
+- clean up introduction
 - writing methods/results pair of section
-- writing discussion
-- setting up bibliography system
-- Possible journals: Methods Ecology and Evolution, Biological Conservation, *Journal of Environmental Management*, *Environmental Monitoring and Assessment*, BioScience
+- build out the supplementary material
+- run theoretical results for more simulations
+- concluding paragraph
+- add citations
+- Possible journals: Journal of Environmental Management, Environmental Monitoring and Assessment, Ecological Applications
 
-# Main messages of manuscript
-
-- We can choose the number of samples and effort we use...
-- Number of samples required depends on question asked and the time series parameters
-- Case study of paleo data shows the utility of iterative sampling in detecting change points
 -->
@@
 
-- It is not limited to paleo data
-- Depends on the question being asked
-- Specific number of samples depends on the system dynamics
-
 Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).

### 2020-03-16 R/manuscript_draft.Rmd
comments post Skype meeting

--- 2020-03-06 R/manuscript_draft.Rmd
@@
 
-Authors: Rosalie Bruel$^{1}$ and Easton R. White$^{2}$
+Authors: Rosalie Bruel$^{1*}$ and Easton R. White$^{2*}$
 
@@
 Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
+* Both authors contributed contributed equally and order was chosen alphabetically / by latitudinal gradient of birth place
 
@@
 
-- theoretical results
-- example with the case study
-- how this fits in with other work
-- limitations and future work
+- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2))
+- money implications (before or now)  
+- how this fits in with other work: examples of situation it would be applied (Rosalie)  
+- how this doesn't fit with other work (limitations) (Rosalie)
+- future work  (?)
 

### 2020-03-16 R/manuscript_draft.Rmd
add year in table

@@
   geom_text(fakedata[fakedata$Year==75,], mapping = aes(variable,Year, label=c("Year", paste("Sp.", c(1:8,"n")))),  size=2.5,hjust=0,angle = 90) +
+  geom_text(fakedata[fakedata$variable==1&fakedata$Year!=75,], mapping = aes(variable,Year, label=c(vr$AGE)),  size=2.1) +
   labs(subtitle="a. Continuous community dataset")

### 2020-03-18 R/manuscript_draft.Rmd
made changes to intro and simulation sections

@@
   \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
-  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
+  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \renewcommand{\thefootnote}{\fnsymbol{footnote}} \addtocounter{footnote}{-1} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
 output:
@@
 \textbf{\Large Sampling requirements and approaches to detect ecosystem shifts}
-\vspace{7 mm}
+\vspace{5 mm}
 
-\textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
-\vspace{5 mm}
+\textsc{Rosalie Bruel$^{1}$\footnote{The authors contributed equally to this work.} and Easton R. White$^{2}$}
+\vspace{3 mm}
 
@@
 
-(150-250 words)
-
-Environmental monitoring is a key piece of managing ecosystems.
+Environmental monitoring is a key component of managing ecosystems.
 Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective.
 In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed.
-For example, sediment or ice cores can be taken in the field in less than an hour, but will be processed in a laboratory setting for anytime between a week and several months, depending on how much information is extracted.
-Sub-sampling allows to reduce the processing time while still giving precious insight.
+For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting.
 Standard sub-sampling approaches often involve equally-spaced sampling.
 We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change.
-We test these ideas with a case study of Cladocera community assemblage, which trajectory was reconstructed from a sediment core.
+We test these ideas with a case study of Cladocera community assemblage, which was reconstructed from a sediment core.
 We demonstrate that standard approaches to sample processing are inefficient.
-For our case study, using an optimal sampling approach would have resulted in savings of XX human-hours or thousands of dollars in labor costs.
+For our case study, using an optimal sampling approach would have resulted in savings of XX person-hours---thousands of dollars in labor costs.
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
@@
 
-Environmental monitoring is one of the core components to modern research and management [@White2019d].
+Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
 Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a].
-However, environmental monitoring is still often expensive and time-consuming, especially when further analyses are needed following sample collection (citations).
-In Ecology and Genetics especially, there has always been a disparity between the amount of data that can be aquired and stored, and the processing methods that could allow to extract the most information from that data.
+However, environmental monitoring is still often expensive and time-consuming, especially when further procesing is needed following sample collection (citations).
+Thus, for many fields there is a disparity between the amount of data that can be aquired and stored, and the ultimate number of samples that can be processed.
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.<!-- processing methods that could allow to extract the most information from that data.
 \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
@@
 Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@McDonald-Madden2010a,@White2019d,@White2020,@Lindenmayer2020a].
-Monitoring programs also need to be designed to be cost-effective given limited budgets [@Caughlan2001a,@Grantham2008b,@Bennett2016a].
-
-<!-- 
-Methods to detect change overtime: see summary https://lindeloev.github.io/mcp/articles/packages.html
-
-- Historical data and ability to choose how much to sample
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
 -->
 
-Accounting for these considerations in the design of a monitoring program is necessary to detect long-term ecological change.
+Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change.
 The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
@@
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
+Thus, monitoring programs also need to be designed to be cost-effective given limited budgets [@Caughlan2001a,@Grantham2008b,@Bennett2016a].
 
@@
 Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We tailor our analysis to the detection of changepoint in a time series, were each point had a certain degree of autocorrelation with the previous, until a changepoint come.
-Changepoints are an important characteristic of a time series.
+We tailor our analysis to the detection of a changepoint in a time series.
+Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes (citations).
 We focus on paleoecological core samples as one example of this type of data.
-We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint.
+We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, i.e.
+a changepoint.
 However, our approaches are widely applicable to other questions and data types.
-We first investigate these tools using a simulated-based approach.
-We then test the tools on a case study from a paleosequence.
+We first investigate these tools using a simulation-based approach.
+We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy (citation).
 
@@
 \ref{fig:conceptual}).
-We then subsampled each time series (citation) to test the effect of three different sampling approaches along with varying the sample size (Fig.
+We then subsampled each time series [@White2020] to test the effect of three different sampling approaches along with varying the sample size (Fig.
 \ref{fig:conceptual}c-f).
-We compared the subsampled time series to the full time series as a measure of the effectiveness of that sampling approach.
+We compared the estimated changepoint from the subsampled time series to that of the full time series as a measure of the effectiveness.
 
@@
 \ref{fig:conceptual}c).
-In the context of sediment cores, this would mean analyzing community composition as random location along the core.
-Random sampling is recommanded in designs aiming at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018).
+In the context of sediment cores, this would mean analyzing community composition at random locations along the core.
+Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018).
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
@@
 \ref{fig:conceptual}d).
-Lastly, iterative sampling involves first taking a set number of samples and then alliteratively adding samples until a pre-determined level of precision is achieved (Fig.
+Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
 \ref{fig:conceptual}e-g).
-In each scenarios, we sample the first and last sample, in case the changepoint takes place early or late in the period covered by the sampling.
+In each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period.
+We describe each approach in more detail in the supplementary material and provide code.
 
-We detect changepoints with the function _e.divisive()_ in the R package _ecp_ (James et al 2019).
+We detect changepoints with the function _e.divisive_ in the R package _ecp_ (James et al 2019).
 There are several methods available to detect changepoint (reviewed in James and Matteson, 2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
-We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of change points.
-Herein, we focus on detecting the most important changepoint, although we tested the method on a time-serie that would have multiple changepoints (Supplementary Materials 1).
-
-We wrote three functions:  
-  * sample_random(): sample first and last sample, in addition of n-2 random samples in between (n = maximum number of sample, chosen by the user),
-  * sample_regular(): sample first and last sample, in addition of n-2 evenly spaced samples in between (n = maximum number of samples, chosen by the user),
-  * sample_regular(): sample first and last sample as well as 3 evenly spaced samples in between (3 is the default but can be modified), in order to initiate the changepoint detection.
-Upon detection of the changepoint on this 5 samples time series, a new sample is added between the changepoint and the previous sample, to narrow down the real changepoint.
-Detection of changepoint / addition of sample is repeated, until it was narrowed down to two consecutive samples, or n (maximum number of samples, chosen by the user) was reached, whichever comes first.
-
-In order to test the performance, we detected the "real" changepoint on the whole time-serie, and compare the changepoint found on the sub-sample with the "real" one.
+We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of changepoints.
+Herein, we focus on detecting the most important changepoint (i.e.
+the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Supplementary Materials 1).
+In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one.
 The distance to real changepoint served as performance diagnostic.
 
-# Theoretical approach
-
-<!--
-- explain the modeling approach
-- show the model
-- explain the analysis we will do
-- explain the theoretical model results with a figure
-- explain effect of different sampling strategies in this context
-
-Theoretical plots
-
-- One combined plot to see the effect (on prob correct or number of samples) of each parameter - should do this with one of the three types of sampling (probably regular)
-- Theoretical plot comparing the three different methods - we could choose one set of parameter values and show how the various methods perform with different sampling amounts (could use same plot structure as the case study)
-- SM plot that shows how the three different methods are affected by changes in the underlying parameter values
-
--->
+# Simulation approach
 
@@
 We began with a theoretical exploration of the sampling requirements to detect a changepoint.
-We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable ($X_t$) that represents either population size, biodiversity, or some metric of community composition at time $t$.
+We modeled a simple first order autoregressive (AR-1) process (the discrete-time version of the Ornstein–Uhlenbeck process) with a response variable ($X_t$) that represents either population size, biodiversity, or some other unidimensional metric of community composition at time $t$.
 The model includes temporal autocorrelation ($\phi$), the mean of the process ($\mu_X$), and a white noise term ($w_t$).
@@
 We explored how each of these model parameters affected our ability to detect a change point.
-We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every interval.
+We simulate an entire time series to serve as the "true" data for comparison [@White2020].
 We specifically study how the amount of samples included and the type of sampling affects the detection probability.
 For simulations, statistical power is the fraction of simulations that were able to detect a changepoint.
-We define a detection as an estimate within five time points (over 100 time points) of the true changepoint.
+We define an accurate changepont detection if an estimate is within five time points (over 100 time points) of the true changepoint.
 We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.
@@
 par(mfrow=c(2,2))
-# Sigma plot
-sigma15 <- param_sensitivity('sigma',15)
-plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
-
-sigma20 <- param_sensitivity('sigma',20)
-points(sigma20,ylab='Probability correct',xlab='sigma',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
-
-# phi plot
-phi15 <- param_sensitivity('phi',15)
-plot(phi15,ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
-
-phi20 <- param_sensitivity('phi',20)
-points(phi20,ylab='Probability correct',xlab='phi',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
 
@@
 
-# 1 - Calculate sigma
-sd_vr <- sd(vrall$matrix$x)
+vr_time_series = vrall$matrix$x
+vr_time_series = (vr_time_series -min(vr_time_series ))/(max(vr_time_series )-min(vr_time_series ))
+
+# 1 - Calculate coefficient of variation
+cv_vr <- sd(vr_time_series)/mean(vr_time_series)
 
 # 2 - Calculate autocorrelation of real time serie
-ar_vr <- acf(vrall$matrix$x, lag.max = 5, plot = FALSE)
+ar_vr <- acf(vr_time_series, lag.max = 5, plot = FALSE)$acf[2]
 
 # 3 - Calculate shift size
-shift_vr <- vrall$matrix$x
-# Normalize
-shift_vr <- (shift_vr-min(shift_vr))/(max(shift_vr)-min(shift_vr))
 # Difference between the two values before/after shift
-shift_vr <- abs(shift_vr[vrall$changepoint]-shift_vr[vrall$changepoint-1])
+shift_vr <- abs(vr_time_series[vrall$changepoint]-vr_time_series[vrall$changepoint-1])/vr_time_series[vrall$changepoint-1]
 
@@
 
-## Theoretical results
+# Simulation results
 

### 2020-03-18 R/supp_material.Rmd
made changes to intro and simulation sections

@@
 \begin{center}
-\textbf{\Large Supplementary Material for: Title here}
+\textbf{\Large Supplementary Material for: Sampling requirements and approaches to detect ecosystem shifts}
 \vspace{7 mm}
@@
 
-Data and code for all the figures can be found at (https://github.com/rosalieb/temporal-sampling.git). All the analyses were run in R.
+Data and code for all the figures can be found at (https://github.com/rosalieb/temporal-sampling.git). All the analyses were run in R (citation).
 
 \clearpage
+
+# Additional text
+
+We wrote three functions: 
+\begin{itemize}
+  \item sample_random(): sample first and last sample, in addition of n-2 random samples in between (n = maximum number of sample, chosen by the user),
+  \item sample_regular(): sample first and last sample, in addition of n-2 evenly spaced samples in between (n = maximum number of samples, chosen by the user),
+  \item sample_iterative(): sample first and last sample as well as 3 evenly spaced samples in between (3 is the default but can be modified), in order to initiate the changepoint detection.
+Upon detection of the changepoint on this 5 samples time series, a new sample is added between the changepoint and the previous sample, to narrow down the real changepoint.
+Detection of changepoint and addition of sample is repeated, until it was narrowed down to two consecutive samples, or n (maximum number of samples, chosen by the user) was reached, whichever comes first.
+\end{itemize}
 
@@
 
-__Main point__: Changepoint was detected even though we used a transformed vector (from community to first component of DCA)
-
 ## References

### 2020-03-19 R/manuscript_draft.Rmd
reorganizing and prepping for the discussion - I just couldn't focus on writing today

@@
 \ref{fig:simulation_min_time}).
+For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr$acf[2],3)`, the variance is `r round(sd_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
 
@@
 
-Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
-For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
-However, in some cases, analyzing the whole samples collected is not an option, because there is not enough processing time or money available.
-We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
+<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
+There is often a difference between the number of samples analyzed and the total number of samples collected, due to time or funding limitations.
+When the processing time is disproportionally higher than the collection time especially, a subsampling can be done posterior to sampling.
+A decision must then be taken of which subsamples to analyze.
+We tested the performance at detecting changepoint of three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}).
+
+We found that the iterative methods was systematically better at detecting changes than the two other methods, the random subsampling being the least efficient (Figs.
+\ref{simulation_number_sample} and \ref{fig:case_study_number_sample}).
+Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.
+(Fig.
+\ref{fig:simulation_min_time}).
+\rosalie{I'm curious to see, once you'll run Fig.
+2 for the  3 methods, whether autocorrelation, shift size and variance really impact iterative method.
+It seems pretty robust!}
+
+<!---- money implications ---->
 
@@
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
-In the context of a real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (\ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
+In the context of a real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
+\ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
 Same results for the major change, with only 12% of the samples analyzed.
@@
 
-Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
-
-Note that the number of sample needed varies depending on the characteristics of the time series.
-For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example is `r round(ar_vr$acf[2],3)`, the variance is `r round(sd_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
+<!---- how this fits in with other work: examples of situation it would be applied (potential) ---->
 
@@
 
-<!--Limitations paragraph-->
+Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
+
+<!---- how this doesn't fit with other work (limitations) ---->
 
@@
 
+<!---- conclusion ---->
+Eventually, it is the researchers choice to choose their approach.
+Our objective was to quantify the implications of an approach vs.
+another.
+In the eventuality that the main objective is to detect a changepoint, we find that an approach is largely superior to the two others.
+
+Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
+For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
+However, in some cases, analyzing the whole samples collected is not an option, because there is not enough processing time or money available.
+We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
+
 Concluding paragraph:

### 2020-03-25 R/manuscript_draft.Rmd
updated code to build theoretical plots

@@
 Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
-* Both authors contributed contributed equally and order was chosen alphabetically / by latitudinal gradient of birth place
+* Both authors contributed contributed equally and order was chosen alphabetically
 
@@
 
-Submission as: Article at \emph{Ecological Applications}
+Submission as: Article at \emph{Journal of Applied Ecology}
 
@@
 
-\textsc{Rosalie Bruel$^{1}$\footnote{The authors contributed equally to this work.} and Easton R. White$^{2}$}
+\textsc{Rosalie Bruel$^{*1}$\footnote{*The authors contributed equally to this work.} and Easton R. White$^{*2}$}
 \vspace{3 mm}
@@
 # Sigma results
-sigma_vec = seq(0.1,30,4)
+sigma_vec = seq(0.01,1,0.05)
 min_time  = rep(0,times=length(sigma_vec))
@@
 # param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
-#                        "value" = c(ar_vr$acf[2], sd_vr, shift_vr))
+#                        "value" = c(ar_vr, cv_vr, shift_vr))
 
@@
   geom_line() +
-  geom_vline(xintercept=ar_vr$acf[2], lty=2) + 
+  geom_vline(xintercept=ar_vr, lty=2) + 
   facet_wrap(~Parameter, scales="free_x", 
@@
   geom_line() +
-  geom_vline(xintercept = sd_vr, lty=2) +
+  geom_vline(xintercept = cv_vr, lty=2) +
   facet_wrap(~Parameter, scales="free_x", 
@@
 \ref{fig:simulation_min_time}).
-For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr$acf[2],3)`, the variance is `r round(sd_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
+For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
 

### 2020-03-25 R/build_time_series.R

@@ -2,16 +2,19 @@
 
 # Build ar-1 time series with various model inputs
 
-build_time_series <- function(mu=0,sigma=3,phi=0.5,shift_size=0.75,shift_time=50,max_time=100){
+build_time_series <- function(mu=0,sigma=0.053,phi=0.404,shift_size=0.81,shift_time=50,max_time=100){
   
   ar1 <- vector(mode = 'numeric',max_time)
-  meanXt = 100
+  meanXt = 0.899
   ar1[1] <- meanXt
   #shift_time = sample(5:(max_time-5),size = 1)
   for (t in 2:max_time){
-    #if(t == shift_time){ar1[t-1]=(1-shift_size)*ar1[1]}
-    if(t == shift_time){meanXt = (1-shift_size)*ar1[1]}
-    ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)
+    #if(t == shift_time){meanXt = (1-shift_size)*ar1[1]}
+    #ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)
+   if(t == shift_time){meanXt = meanXt - shift_size;ar1[t]=meanXt}
+    else{ ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)}
   }
+  
+  ar1 = (ar1-min(ar1))/(max(ar1)-min(ar1))
   return(ar1) 
 }

### 2020-03-25 R/manuscript_draft.Rmd
updated discussion + conclusion section

@@
 # Sigma results
-sigma_vec = seq(0.01,1,0.05)
+sigma_vec = seq(0.01,1,0.45)
 min_time  = rep(0,times=length(sigma_vec))
@@
 ```
-
-\rosalie{I wonder whether we should show the results with random sampling for the figure showing the minimum number sampled required for different sigma, phi, and shift size: we show later that it is the worst case scenario.
-But maybe that's a discussion we already had and I am missing on something.
-<br> I calculated the autocorrelation at lag-1 for the case study time series and added it to the plot here.}
 
@@
 As one application of our approach, we examined paleosequence data.
-To obtain this data sediment cores are taken from aquatic environments to reconstruct past ecosystems.
+To obtain this data sediment cores were taken from aquatic environments to reconstruct past ecosystems.
 Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @Perga2015).
@@
 A decision must then be taken of which subsamples to analyze.
-We tested the performance at detecting changepoint of three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}).
+To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig.
+\ref{fig:conceptual}).
 
@@
 
-Multiple subsampling strategies can be chosen(Fig.
-\ref{fig:conceptual}), but only iterative sampling allowed to approach the real changepoint in a limited number of samples (Fig.
+Multiple subsampling strategies can be chosen (Fig.
+\ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig.
 \ref{fig:simulation_min_time}c).
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
+
 In the context of a real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive (citations).
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019,@White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach in this paper was on data where more samples can be added long after the dynamics occurred.
+Our specific approach was with data where more samples can be added, or processed, long after the dynamics occurred.
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
+However, both different types of data and different questions can be addressed with the same approach.
 Suppose instead that the goal was to detect a change in observed individuals with video-based approaches.
-It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address our question of interest.
+It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address our question of interest (citations).
 Using an interval sampling approach, one could take a fixed number of samples to start.
@@
 If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample.
-Sampling would continue until the trend estimate began to settle down.
+Sampling would continue until the trend estimate reached an equilibrium.
 Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
-These approaches are also applicable to other questions related to the optimal sample size and approach in determining long-term trends, biodiversity estimates, and more (citations).
+These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples (citations).
 
@@
 The specific number of samples and optimal sampling approach will largely be system specific.
-A similar system may be used to get some sense for the best sampling scheme, but it will depend on the VARIOUS FACTORS HERE.
-Similarly, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-However, more samples could be added to address these additional questions.
-
-Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
+A similar system may be used to get some sense for the best sampling scheme, but it will depend on the system dynamics (e.g.
+interannual variation, trend effect size).
+The specific sampling scheme also depends on the question of interest.
+For instance, changepoint detection may require fewer samples than detecting a change in biodiversity indices (citations).
+Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
+However, more samples could be processed later to address these additional questions.
+Lastly, we know systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
 The lower number of input data prior to ordination did not impact the finding of the main transition.
@@
 
+<!--not sure where to include this 
 Many ecological processes are non linear (e.g., D'Amario et al 2019, Scientific reports).
-Detecting a changepoint on one aspect of the ecosystem does not guarantee that the sampling resolution is appropriate to detect the driver for the changepoint.
+Detecting a changepoint on one aspect of the ecosystem does not guarantee that the sampling resolution is appropriate to detect the driver for the changepoint.--->
 
 <!---- conclusion ---->
+
+Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
+For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
+We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
+This is valuable given the high costs of many monitoring programs (citations).
+Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
+The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach (@Lovett2007).
+Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
+
+<!--
 Eventually, it is the researchers choice to choose their approach.
@@
 In the eventuality that the main objective is to detect a changepoint, we find that an approach is largely superior to the two others.
-
-Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
-For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
-However, in some cases, analyzing the whole samples collected is not an option, because there is not enough processing time or money available.
-We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
-
-Concluding paragraph:
-
-About recommendation to reduce the frequency of data collected as part of monitoring program:  
-While long-term monitoring program are often expensive, these cost are nothing compared to the value of the resources monitored and the policies affecting these resources (@Lovett2007).
-Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may indeed not be necessary to manage ecosystems in the future.
+-->
 

### 2020-03-25 R/calc_min_time2.R

@@ -1,0 +1,61 @@
+
+calc_min_time = function(param_name,param_value){
+
+#sigma_level = 20
+power_level = 0.8
+sample_size = 15
+gap_size = 5
+prob_correct = 0
+#combined_correct = NULL
+
+switch(param_name,
+       sigma = {params <- tibble(sigma=rep(param_value,50))},
+       phi = {params <- tibble(phi=rep(param_value,100))},
+       shift_size = {params <- tibble(shift_size=rep(param_value,100))},
+       shift_time = {params <- tibble(shift_time=rep(param_value,100))}
+)
+
+params$shift_time = sample(30:70,nrow(params),replace = T)  
+
+
+time_series <- params %>%
+  pmap(build_time_series) 
+
+while (prob_correct<power_level & sample_size<50){
+  sample_size = sample_size + gap_size
+  
+  # Sample regular for each time series
+  output <- time_series %>%
+    map_dbl(function(df) sample_regular(df,sample_size)$changepoint) %>%
+    mutate(params,detection=.)
+  
+  names(output) = c('param_name','shift_time','detection')
+  output$detection[is.na(output$detection)] <- 0
+  output$detection <- ifelse(abs(output$detection-output$shift_time)<5,1,0) # If detected changepoint is within 5 time points, record a successful detection
+  
+  prob_correct <- output %>%
+    group_by(param_name) %>%
+    summarize(prob_correct = sum(detection==1)/length(param_name/unique(param_name))) %>%
+    select(prob_correct)
+    #%>%
+  #plot(ylab='Probability correct',xlab=paste(param_name),pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+  
+ # combined_correct <- c(combined_correct,prob_correct$prob_correct)
+
+if (as.numeric(prob_correct) > power_level & gap_size > 1){
+    sample_size = sample_size - gap_size
+    gap_size = 1
+    prob_correct=0
+  }
+
+
+  
+} # End of while loop  
+
+if(sample_size>15 & sample_size<51){
+  return(as.numeric(sample_size))
+}else{
+  return(NA)
+}
+
+} # End of function

### 2020-03-25 R/manuscript_draft.Rmd
manuscript edits

@@
 We demonstrate that standard approaches to sample processing are inefficient.
-For our case study, using an optimal sampling approach would have resulted in savings of XX person-hours---thousands of dollars in labor costs.
+For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs.
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
-We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.
+We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling, and where the objective is to detect ecosystem change.
 
@@
 
-Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
+Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a].
 Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
@@
 The specific number of years required depended on the species biology and the detection method used [@White2019d].
-Other work has focused on the frequency of monitoring [@Wauchope2019a], the allocation of resources spatially versus temporally [@Rhodes2011a,@Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
+Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitave change in population.
+Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
-Thus, monitoring programs also need to be designed to be cost-effective given limited budgets [@Caughlan2001a,@Grantham2008b,@Bennett2016a].
+The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].
 
@@
 
+Each subsampling was done from the initial community dataset (Fig.
+\ref{fig:case_study_iterative_sampling}a).
+An ordination was run on the number of samples selected, and the changepoint analysis was run on that new vector.
+In case of the iterative method, a new sample was added and the ordination was run again (Fig.
+\ref{fig:case_study_iterative_sampling}d).
+
 ## Case study results
@@
 
+We make the point that if a subsample of the samples must be used, the temptation to be unbiased by using a regular subsampling will lead to losing a lot of accuracy in the process.
+If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable
+
 <!--

### 2020-03-26 R/manuscript_draft.Rmd
work on the discussion

@@
 header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
-  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
+  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount} \usepackage{soul}
   \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \renewcommand{\thefootnote}{\fnsymbol{footnote}} \addtocounter{footnote}{-1} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Total number of samples was set between 5 and 30, out of the 74 initial time series.
-\\label{fig:case_study_number_sample}'}
-# Read data
-myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
-myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
-myoutput$diff_real <- abs(myoutput$diff_real)
-
-# Change to TRUE if we decide it is worth it to show several initial number of n2. It looks more confusing -- but it would bring more information
-several_n2_init = F
-
-# Plot
-p1 <- if(several_n2_init) {
-  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method,
-             labeller = label_glue('{.l}. {method} sampling ')) + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        strip.background = element_rect(colour="transparent", fill="transparent"),
-        strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.72),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-} else {
-  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method,
-             labeller = label_glue('{.l}. {method} sampling ')) + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        strip.background = element_rect(colour="transparent", fill="transparent"),
-        strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint") 
-}
-
-p1
-
-```
-
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 8, fig.height= 7.5, fig.cap='Iterative sampling and impact on ordination.
@@
 
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in commuper total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample_change_biomass}'}
+# Read data
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Change to TRUE if we decide it is worth it to show several initial number of n2. It looks more confusing -- but it would bring more information
+several_n2_init = F
+
+# Plot
+p1 <- if(several_n2_init) {
+  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.72),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+} else {
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") 
+}
+
+p1
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample}'}
+#investigating species average biomass
+vr_count <- rowSums(vr[,-1])
+myoutput_count <- NULL
+for (i in 5:30) {
+  vrsub <- sample_regular(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_random(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 6)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 7)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+}
+
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese_biomass.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Plot
+p1 <-
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint")
+
+p1
+
+```
+
+\rosalie{About figure \ref{fig:case_study_number_sample_change_biomass}, I tried testing the real changepoint on another potential metric, i.e., biomass.
+Potential tools for discussion?
+I don't know!
+Maybe worth talking about that during our next Skype.}
+
 # General Discussion
@@
 
-We found that the iterative methods was systematically better at detecting changes than the two other methods, the random subsampling being the least efficient (Figs.
+We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
 \ref{simulation_number_sample} and \ref{fig:case_study_number_sample}).
@@
 \ref{fig:simulation_min_time}).
-\rosalie{I'm curious to see, once you'll run Fig.
-2 for the  3 methods, whether autocorrelation, shift size and variance really impact iterative method.
-It seems pretty robust!}
 
@@
 
-In the context of a real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
+Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
 \ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
@@
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach was with data where more samples can be added, or processed, long after the dynamics occurred.
+Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred.
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
@@
 
+ Getting an average biomass when the increase in biomass is continuous can also benefit from our approach (Fig. \ref{fig:case_study_number_sample_change_biomass}).
+
 <!---- how this doesn't fit with other work (limitations) ---->
 
-Although our approach is flexible to be applied to various data sources and questions, this is also one of its limitations.
-The specific number of samples and optimal sampling approach will largely be system specific.
+Our approach is able to address other data sources and questions, but not any.
+When processing time and resources are less than the collection time, investigators will be likely to process every sample.
+If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Fig.
+\ref{simulation_number_sample}).
+More generally, some questions require a continuous time serie, or at least a regular sampling interval.
+For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts (Frossard et al 2015, Doncaster et al 2016).
+However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down (Spears et al, 2017).
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings (Doncaster et al 2016), but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
+
+\rosalie{I don't understand what you meant here. I tried reformulating some of it, but we can talk about it more if needed}:  
+\st{The specific number of samples and optimal sampling approach will largely be system specific.
 A similar system may be used to get some sense for the best sampling scheme, but it will depend on the system dynamics (e.g.
@@
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-However, more samples could be processed later to address these additional questions.
-Lastly, we know systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016).
-The lower number of input data prior to ordination did not impact the finding of the main transition.
-However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.
+However, more samples could be processed later to address these additional questions.}
 
@@
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
-
 We make the point that if a subsample of the samples must be used, the temptation to be unbiased by using a regular subsampling will lead to losing a lot of accuracy in the process.
-If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable
+If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable.
 

### 2020-03-26 R/manuscript_draft.Rmd
small edit to discussion

@@
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
+It matches very well the paleoecological approach (Wingard et al, 2017).
 However, both different types of data and different questions can be addressed with the same approach.
@@
 
-Other dataset that could use our method, e.g., paleo science in general? see review by Wingard et al 2017 https://www.frontiersin.org/articles/10.3389/fevo.2017.00011/full#B140
-
- Getting an average biomass when the increase in biomass is continuous can also benefit from our approach (Fig. \ref{fig:case_study_number_sample_change_biomass}).
+The approach is versatile: it also finds in 9 samples (11 to confirm) the change in species biomass (Fig.
+\ref{fig:case_study_number_sample_change_biomass}).
+The initial 5 samples analyzed are the same: a well tailored question from the start allows to reach the changepoint much more efficiently than random and regular subsampling methods.
+(\rosalie{do we want to include that?
+not sure})
 

### 2020-03-29 R/manuscript_draft.Rmd
updated theoretical plot

--- 2020-03-25 R/manuscript_draft.Rmd
@@
 We demonstrate that standard approaches to sample processing are inefficient.
-For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs.
+For our case study, using an optimal sampling approach would have resulted in savings of XX person-hours---thousands of dollars in labor costs.
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
-We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling, and where the objective is to detect ecosystem change.
+We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.
 
@@
 
-Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a].
+Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
 Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
@@
 The specific number of years required depended on the species biology and the detection method used [@White2019d].
-Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitave change in population.
-Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
+Other work has focused on the frequency of monitoring [@Wauchope2019a], the allocation of resources spatially versus temporally [@Rhodes2011a,@Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
-The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].
+Thus, monitoring programs also need to be designed to be cost-effective given limited budgets [@Caughlan2001a,@Grantham2008b,@Bennett2016a].
 
@@
 # Sigma results
-sigma_vec = seq(0.01,1,0.45)
+sigma_vec = seq(0.01,0.5,0.025)
 min_time  = rep(0,times=length(sigma_vec))
@@
 
-phi_vec = seq(-0.8,0.8,0.2)
+phi_vec = seq(-0.8,0.8,0.4)
 min_time  = rep(0,times=length(phi_vec))
@@
 
-shift_size_vec = rev(seq(0.1,0.9,0.025))
+shift_size_vec = rev(seq(0.1,0.9,0.05))
 min_time  = rep(0,times=length(shift_size_vec))
@@
   geom_line() +
-  geom_vline(xintercept=ar_vr, lty=2) + 
+  geom_vline(xintercept=0.05, lty=2) + 
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('a. {Parameter}')) +
-  ylim(20,50) +
+  ylim(10,50) +
   labs(x=c("sigma"), y= "Minimum number of samples required") + 
@@
   geom_line() +
-  geom_vline(xintercept = cv_vr, lty=2) +
+  geom_vline(xintercept = 0.404, lty=2) +
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('b. {Parameter}')) +
-  ylim(20,50) +
+  ylim(10,50) +
   labs(x="phi", y= "") + 
@@
   geom_line() + 
-  geom_vline(xintercept = shift_vr, lty=2) +
+  geom_vline(xintercept = 0.81, lty=2) +
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('c. {Parameter}')) +
-  ylim(20,50) +
+  ylim(10,50) +
   labs(x="shift size", y= "") + 
@@
 For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.
-
-Each subsampling was done from the initial community dataset (Fig.
-\ref{fig:case_study_iterative_sampling}a).
-An ordination was run on the number of samples selected, and the changepoint analysis was run on that new vector.
-In case of the iterative method, a new sample was added and the ordination was run again (Fig.
-\ref{fig:case_study_iterative_sampling}d).
 
@@
 
-We make the point that if a subsample of the samples must be used, the temptation to be unbiased by using a regular subsampling will lead to losing a lot of accuracy in the process.
-If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable
-
 <!--

### 2020-03-30 R/manuscript_draft.Rmd
updated plots in both manuscript and supp mat

@@
 ---
-bibliography: Whitebib.bib
+bibliography: ChangepointSampling.bib
 csl: ecology-letters.csl
@@
 Thus, for many fields there is a disparity between the amount of data that can be aquired and stored, and the ultimate number of samples that can be processed.
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.<!-- processing methods that could allow to extract the most information from that data.
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
 \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
@@
 Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a,@McDonald-Madden2010a,@Lindenmayer2020a].
 -->
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions (Cohen, 2003).
+For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
 Similarly, environmental samples (e.g.
 water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a].
-Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later (citations).
+Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a,@Mallet2014].
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
@@
 We tailor our analysis to the detection of a changepoint in a time series.
-Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes (citations).
+Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes [@James2014].
 We focus on paleoecological core samples as one example of this type of data.
@@
 We first investigate these tools using a simulation-based approach.
-We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy (citation).
+We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy [@Bruel2018].
 
@@
 In the context of sediment cores, this would mean analyzing community composition at random locations along the core.
-Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018).
+Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018].
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
@@
 
-We detect changepoints with the function _e.divisive_ in the R package _ecp_ (James et al 2019).
-There are several methods available to detect changepoint (reviewed in James and Matteson, 2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
+We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019].
+There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
 We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of changepoints.
 Herein, we focus on detecting the most important changepoint (i.e.
-the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Supplementary Materials 1).
+the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4).
 In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one.
@@
 
-```{r theoretical_results_old,message=F,warning=F,echo=F,eval=F}
-
-source('R/sampling_regular.R')
-source('R/sampling_random.R')
-source('R/build_time_series.R')
-source('R/updated_param_sensitivity.R')
-
-require(tidyverse)
-
-par(mfrow=c(2,2))
-
-```
-
 ```{r simulation_min_time,message=F,warning=F,echo=F,eval=F}
@@
 # Sigma results
-sigma_vec = seq(0.01,0.5,0.025)
+sigma_vec = seq(0.01,0.4,0.025)
 min_time  = rep(0,times=length(sigma_vec))
@@
 for (param_num in 1:nrow(sigma_output)){
-  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
+  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'regular')
   print(sigma_output)
@@
 
-phi_vec = seq(-0.8,0.8,0.4)
+phi_vec = seq(-0.8,0.8,0.05)
 min_time  = rep(0,times=length(phi_vec))
@@
 for (param_num in 1:nrow(phi_output)){
-  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
+  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num],'regular')
   print(phi_output)
@@
 for (param_num in 1:nrow(shift_size_output)){
-  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
+  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num],'regular')
   print(shift_size_output)
@@
 ```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
-Random sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+Regular sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
 Vertical line indicate respective parameter calculated from the case study time series.
-\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
+The same analyses for random and iterative sampling is in Figs.
+S2,S3.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}
 
@@
 Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
-\ref{fig:simulation_min_time}).
+\ref{fig:simulation_min_time_plot}).
 For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
@@
 
-```{r simulation_approaches_graph, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_number_sample}'}
+```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_approaches_plot}'}
 
@@
 To obtain this data sediment cores were taken from aquatic environments to reconstruct past ecosystems.
-Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @Perga2015).
+Samples can be taken before and after a perturbation to compare community composition, or continuously to identify the timing of change and underlying mechanisms of the change [@Perga2015].
 
@@
 Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999,@Bruel2018].
-Air temperature is now driving changes in plankton communities (@Bruel2018).
+Air temperature is now driving changes in plankton communities [@Bruel2018].
 
@@
 # Read data
+
+require(stickylabeller)
+
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+```{r echo=FALSE, eval=F,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
@@
 We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
-\ref{simulation_number_sample} and \ref{fig:case_study_number_sample}).
+\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3).
 Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.
 (Fig.
-\ref{fig:simulation_min_time}).
+\ref{fig:simulation_min_time_plot}).
 
@@
 \ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig.
-\ref{fig:simulation_min_time}c).
+\ref{fig:simulation_approaches_plot}c).
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
@@
 Same results for the major change, with only 12% of the samples analyzed.
-The method also work to detect further shifts (Supplementary Materials 1).
+The method also work to detect further shifts (Fig.
+S4).
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019,@White2020].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
@@
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
-It matches very well the paleoecological approach (Wingard et al, 2017).
+It matches very well the paleoecological approach [@Wingard2017].
 However, both different types of data and different questions can be addressed with the same approach.
 Suppose instead that the goal was to detect a change in observed individuals with video-based approaches.
-It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address our question of interest (citations).
+It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
 Using an interval sampling approach, one could take a fixed number of samples to start.
@@
 Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
-These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples (citations).
+These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
 
@@
 <!---- how this doesn't fit with other work (limitations) ---->
+<!-- need line that references issue that iterative sampling can get stuck on incorrect points-->
 
-Our approach is able to address other data sources and questions, but not any.
+Our approach is able to address other data sources and questions, but not all.
 When processing time and resources are less than the collection time, investigators will be likely to process every sample.
-If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Fig.
-\ref{simulation_number_sample}).
-More generally, some questions require a continuous time serie, or at least a regular sampling interval.
-For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts (Frossard et al 2015, Doncaster et al 2016).
-However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down (Spears et al, 2017).
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings (Doncaster et al 2016), but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
+If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Figs.
+\ref{fig:simulation_approaches_plot},S2,S3).
+More generally, some questions require a continuous time series, or at least a regular sampling interval.
+For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016].
+However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017].
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
 
@@
 We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
-This is valuable given the high costs of many monitoring programs (citations).
+This is valuable given the high costs of many monitoring programs [@Caughlan2001a,Bennett2014a].
 Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
-The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach (@Lovett2007).
+The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007,@Bennett2018b].
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.

### 2020-03-30 R/supp_material.Rmd
updated plots in both manuscript and supp mat

@@
   \usepackage{amsmath}
-  \renewcommand{\theequation}{S\arabic{equation}}
+  \renewcommand{\theequation}{S\arabic{equation}} \floatplacement{figure}{H}
 output:
@@
 
-We wrote three functions: 
-\begin{itemize}
-  \item sample_random(): sample first and last sample, in addition of n-2 random samples in between (n = maximum number of sample, chosen by the user),
-  \item sample_regular(): sample first and last sample, in addition of n-2 evenly spaced samples in between (n = maximum number of samples, chosen by the user),
-  \item sample_iterative(): sample first and last sample as well as 3 evenly spaced samples in between (3 is the default but can be modified), in order to initiate the changepoint detection.
+For each sampling approach, we wrote three custom functions:
+
+- sample_random(): sample first and last sample, in addition of n-2 random samples in between (n = maximum number of sample, chosen by the user),  
+- sample_regular(): sample first and last sample, in addition of n-2 evenly spaced samples in between (n = maximum number of samples, chosen by the user), and  
+- sample_iterative(): sample first and last sample as well as 3 evenly spaced samples in between (3 is the default but can be modified), in order to initiate the changepoint detection.
 Upon detection of the changepoint on this 5 samples time series, a new sample is added between the changepoint and the previous sample, to narrow down the real changepoint.
 Detection of changepoint and addition of sample is repeated, until it was narrowed down to two consecutive samples, or n (maximum number of samples, chosen by the user) was reached, whichever comes first.
-\end{itemize}
 
-# Additional figures
+# Parameter sensitivity
 
-## Simulation approach
-
-```{r theoretical_results,message=F,warning=F,echo=F,eval=F}
+```{r parameter_sensitivity,message=F,warning=F,echo=F,eval=F,cache=F}
+# Build 1x3 plot with sigma vs phi with color equal to distance to changepoint, power, or min time. The three panel plots would represent different shift size values
 
@@
 source('R/build_time_series.R')
-source('R/updated_param_sensitivity.R')
-
+source('R/calc_min_time2.R')
 require(tidyverse)
 
-par(mfrow=c(2,2))
-# Sigma plot
-sigma15 <- param_sensitivity('sigma',15)
-plot(sigma15,ylab='Probability correct',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+number_sims <- 10
+# Find minimum time for different parameter combos
 
-sigma20 <- param_sensitivity('sigma',20)
-points(sigma20,ylab='Probability correct',xlab='sigma',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
+sigma_vec = seq(0.01,0.3,0.02)
+phi_vec = seq(-0.8,0.8,0.1)
+shift_size_vec = c(0.1,0.4,0.8)
 
-# phi plot
-phi15 <- param_sensitivity('phi',15)
-plot(phi15,ylab='Probability correct',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,1),col=1)
+output_theory_approach <- data.frame(expand.grid(sigma_vec,phi_vec,shift_size_vec))
+names(output_theory_approach) <- c('sigma_vec', 'phi_vec', 'shift_size_vec')
+output_theory_approach$power = 0
 
-phi20 <- param_sensitivity('phi',20)
-points(phi20,ylab='Probability correct',xlab='phi',pch=15,cex.lab=1.3,las=1,ylim=c(0,1),col=3)
+#params <- tibble(shift_time = sample(30:70,max(simulation_number),replace = T))
+
+shift_time = sample(30:70,number_sims,replace=T) # Keep shift times constant for each param combo
+
+# Run through params
+for (i in 1:nrow(output_theory_approach)){
+
+# Build XX number of time series
+params <- list(sigma=rep(output_theory_approach$sigma_vec[i],number_sims),phi=rep(output_theory_approach$phi_vec[i],number_sims),shift_size=rep(output_theory_approach$shift_size_vec[i],number_sims),shift_time = shift_time)
+
+time_series <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) sample_regular(df,20)$changepoint)
+
+power <- sum(abs(time_series-shift_time)<5)/number_sims
+
+#print(i)
+output_theory_approach$power[i] <- power
+}
+
+write.csv(x = output_theory_approach,file = 'Output/parameter_sensitivity.csv',quote = FALSE )
+
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
+  facet_wrap(~shift_size_vec) +
+   labs(fill='Statistical power') +
+   xlab("Lag-1 autocorrelation (phi)") +
+   ylab("Standard deviation (sigma)") +
+   theme_classic(base_size = 14, base_family = "")
 
@@
 
-```{r theoretical_results_min_time,message=F,warning=F,echo=F,eval=F}
+```{r parameter_sensitivity_plot,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Regular sampling statistical power (fraction of 100 simulations which detecting a changepoint within five time points of the true changepoint) for different levels of standard deviation, lag-1 autocorrelation, and shift size.\\label{fig:parameter_sensitivity_plot}'}
+output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity.csv')
 
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
+  facet_wrap(~shift_size_vec) + 
+   theme_classic(base_size = 14, base_family = "")+
+  theme(plot.subtitle = element_text(hjust = 0.5)) +
+   labs(fill='Power',subtitle = 'Shift size') +
+   xlab("Lag-1 autocorrelation (phi)") +
+   ylab("Standard deviation (sigma)")
+```
+
+# Minimum time for other sampling approaches
+
+## Random sampling
+
+```{r simulation_min_time_random,message=F,warning=F,echo=F,eval=F,cache=T}
+
+source('R/sampling_random.R')
 source('R/sampling_regular.R')
-source('R/sampling_random.R')
+source('R/sampling_iterative.R')
 source('R/build_time_series.R')
@@
 # Sigma results
-sigma_vec = seq(0.1,30,4)
+sigma_vec = seq(0.01,0.4,0.025)
 min_time  = rep(0,times=length(sigma_vec))
@@
 for (param_num in 1:nrow(sigma_output)){
-  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
-  print(sigma_output)
+  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'random')
+  #print(sigma_output)
 }
 
-write.csv(x = sigma_output,file = 'Output/theoretical_results_sigma.csv',quote=FALSE)
+write.csv(x = sigma_output,file = 'Output/theoretical_results_random_sigma.csv',quote=FALSE)
 
@@
 
-phi_vec = seq(-0.8,0.8,0.2)
+phi_vec = seq(-0.8,0.8,0.05)
 min_time  = rep(0,times=length(phi_vec))
@@
 for (param_num in 1:nrow(phi_output)){
-  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
-  print(phi_output)
+  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num],'random')
+  #print(phi_output)
 }
 
-write.csv(x = phi_output,file = 'Output/theoretical_results_phi.csv',quote=FALSE)
+write.csv(x = phi_output,file = 'Output/theoretical_results_random_phi.csv',quote=FALSE)
 
@@
 for (param_num in 1:nrow(shift_size_output)){
-  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
-  print(shift_size_output)
+  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num],'random')
+  #print(shift_size_output)
 }
 
-write.csv(x = shift_size_output,file = 'Output/theoretical_results_shift_size.csv',quote=FALSE)
+write.csv(x = shift_size_output,file = 'Output/theoretical_results_random_shift_size.csv',quote=FALSE)
 
@@
 
-```{r theoretical_results_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
-The default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+```{r simulation_min_time_plot_random, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
+Random sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
-\\label{fig:theoretical_min_time}',message=F,warning=F,echo=F,eval=T}
-
+Each vertical line indicates the respective parameter calculated from the case study time series.
+\\label{fig:simulation_min_time_random}',message=F,warning=F,echo=F,eval=T}
+require(ggplot2)
+require(dplyr)
+require(stickylabeller)
 par(mfrow=c(1,3))
 
-sigma <- read.csv('Output/theoretical_results_sigma.csv',header=TRUE)
-plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+sigma <- read.csv('Output/theoretical_results_random_sigma.csv',header=TRUE)
 
-# Phi results
+phi <- read.csv('Output/theoretical_results_random_phi.csv',header=TRUE)
 
-phi <- read.csv('Output/theoretical_results_phi.csv',header=TRUE)
-plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+shift_size <- read.csv('Output/theoretical_results_random_shift_size.csv',header=TRUE)
 
-# Shift size results
+# Melt the 3 dataset to be able to use facet_wrap()
+colnames(sigma) <- colnames(phi) <- colnames(shift_size) <- c("X", "Param", "min_time")
+simulation_all <- 
+  rbind(cbind("Parameter" = rep("Standard deviation", nrow(sigma)), sigma),
+        cbind("Parameter" = rep("Autocorrelation at lag-1", nrow(phi)), phi),
+        cbind("Parameter" = rep("Normalized amplitude of shift", nrow(shift_size)), shift_size))
+# param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
+#                        "value" = c(ar_vr, cv_vr, shift_vr))
 
-shift_size <- read.csv('Output/theoretical_results_shift_size.csv',header=TRUE)
-plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(20,50),col=1,type='o')
+p1 <- 
+  ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept=0.05, lty=2) + 
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('a. {Parameter}')) +
+  ylim(10,50) +
+  labs(x=c("sigma"), y= "Minimum number of samples required") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept = 0.404, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('b. {Parameter}')) +
+  ylim(10,50) +
+  labs(x="phi", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
+  geom_point() +
+  geom_line() + 
+  geom_vline(xintercept = 0.81, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('c. {Parameter}')) +
+  ylim(10,50) +
+  labs(x="shift size", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+
+# Remove the y-axis for the 2nd plot - p2 then merge 2 plots
+cowplot::plot_grid(p1 + theme(plot.margin = unit(c(0.15, 0.1, 0.15, 0.15), "cm")), 
+                   p2 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   p3 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   nrow = 1,
+                   rel_widths = c(1.2, 1,1),
+                   align = 'h', axis = 'tb')
 
@@
 
-```{r theoretical_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
+## Iterative sampling
 
-source('R/sampling_regular.R')
-source('R/sampling_random.R')
+```{r simulation_min_time_iterative,message=F,warning=F,echo=F,eval=F,cache=T}
+
+source('R/sampling_iterative.R')
 source('R/build_time_series.R')
@@
 
-# Need to run each method on a set of simulated time series. We would then have 100 runs for each method and each sampling amount
+# Sigma results
+sigma_vec = seq(0.01,0.4,0.025)
+min_time  = rep(0,times=length(sigma_vec))
+sigma_output <- data.frame(sigma_vec,min_time)
 
-approach <- c('random','regular','iterative')
-number_samples <- 1:30
-simulation_number <- 1:100
-output_theory_approach <- data.frame(expand.grid(approach,number_samples,simulation_number))
-names(output_theory_approach) <- c('approach', 'number_samples', 'simulation_number')
-output_theory_approach$distance_to_real = 0
+for (param_num in 1:nrow(sigma_output)){
+  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'iterative')
+  #print(sigma_output)
+}
 
-# Build function that takes in approach, simulated time series, and number of samples. The function should then spit out the distance to the real changepoint
+write.csv(x = sigma_output,file = 'Output/theoretical_results_iterative_sigma.csv',quote=FALSE)
 
-# Set of time series (need other params)
-params <- tibble(shift_time = sample(30:70,nrow(params),replace = T))
+# Phi results
 
-time_series <- params %>%
-  pmap(build_time_series)
+phi_vec = seq(-0.8,0.8,0.05)
+min_time  = rep(0,times=length(phi_vec))
+phi_output <- data.frame(phi_vec,min_time)
 
-# Run through params
+for (param_num in 1:nrow(phi_output)){
+  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num],'iterative')
+  #print(phi_output)
+}
 
-for (i in 1:nrow(output_theory_approach)){
+write.csv(x = phi_output,file = 'Output/theoretical_results_iterative_phi.csv',quote=FALSE)
 
-  output_theory_approach$distance_to_real[i] <- calc_distance_to_real(output_theory_approach$approach,output_theory_approach$number_samples,time_series[output_theory_approach$simulation_number])
+# Shift size
+
+shift_size_vec = rev(seq(0.1,0.9,0.05))
+min_time  = rep(0,times=length(shift_size_vec))
+shift_size_output <- data.frame(shift_size_vec,min_time)
+
+for (param_num in 1:nrow(shift_size_output)){
+  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num],'iterative')
+  #print(shift_size_output)
 }
+
+write.csv(x = shift_size_output,file = 'Output/theoretical_results_iterative_shift_size.csv',quote=FALSE)
 
@@
 
-## Case study
+```{r simulation_min_time_plot_iterative, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
+Iterative sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study.
+The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
+Each vertical line indicates the respective parameter calculated from the case study time series.
+\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
+
+par(mfrow=c(1,3))
+
+sigma <- read.csv('Output/theoretical_results_iterative_sigma.csv',header=TRUE)
+
+phi <- read.csv('Output/theoretical_results_iterative_phi.csv',header=TRUE)
+
+shift_size <- read.csv('Output/theoretical_results_iterative_shift_size.csv',header=TRUE)
+
+# Melt the 3 dataset to be able to use facet_wrap()
+colnames(sigma) <- colnames(phi) <- colnames(shift_size) <- c("X", "Param", "min_time")
+simulation_all <- 
+  rbind(cbind("Parameter" = rep("Standard deviation", nrow(sigma)), sigma),
+        cbind("Parameter" = rep("Autocorrelation at lag-1", nrow(phi)), phi),
+        cbind("Parameter" = rep("Normalized amplitude of shift", nrow(shift_size)), shift_size))
+# param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
+#                        "value" = c(ar_vr, cv_vr, shift_vr))
+
+p1 <- 
+  ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept=0.05, lty=2) + 
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('a. {Parameter}')) +
+  ylim(10,50) +
+  labs(x=c("sigma"), y= "Minimum number of samples required") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
+  geom_point() +
+  geom_line() +
+  geom_vline(xintercept = 0.404, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('b. {Parameter}')) +
+  ylim(10,50) +
+  labs(x="phi", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
+  geom_point() +
+  geom_line() + 
+  geom_vline(xintercept = 0.81, lty=2) +
+  facet_wrap(~Parameter, scales="free_x", 
+             labeller = label_glue('c. {Parameter}')) +
+  ylim(10,50) +
+  labs(x="shift size", y= "") + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0))
+
+# Remove the y-axis for the 2nd plot - p2 then merge 2 plots
+cowplot::plot_grid(p1 + theme(plot.margin = unit(c(0.15, 0.1, 0.15, 0.15), "cm")), 
+                   p2 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   p3 + 
+                     theme(plot.margin = unit(c(0, 0.1, 0, 0), "cm"),
+                           axis.text.y = element_blank(),
+                           axis.line.y = element_blank(),
+                           axis.title.y= element_blank(),
+                           axis.ticks.y= element_blank()),
+                   nrow = 1,
+                   rel_widths = c(1.2, 1,1),
+                   align = 'h', axis = 'tb')
+
+```
+
+# Case study
 
@@
 
+\clearpage
+
 ## References

### 2020-03-30 R/calc_min_time2.R

@@ -1,17 +1,17 @@
 
-calc_min_time = function(param_name,param_value){
+calc_min_time = function(param_name,param_value,sampling_approach){
 
 #sigma_level = 20
 power_level = 0.8
-sample_size = 15
+sample_size = 10
 gap_size = 5
 prob_correct = 0
 #combined_correct = NULL
 
 switch(param_name,
        sigma = {params <- tibble(sigma=rep(param_value,50))},
-       phi = {params <- tibble(phi=rep(param_value,100))},
-       shift_size = {params <- tibble(shift_size=rep(param_value,100))},
+       phi = {params <- tibble(phi=rep(param_value,50))},
+       shift_size = {params <- tibble(shift_size=rep(param_value,50))},
        shift_time = {params <- tibble(shift_time=rep(param_value,100))}
 )
 
@@ -21,13 +21,24 @@ params$shift_time = sample(30:70,nrow(params),replace = T)
 time_series <- params %>%
   pmap(build_time_series) 
 
-while (prob_correct<power_level & sample_size<50){
+while (prob_correct<power_level & sample_size<85){
   sample_size = sample_size + gap_size
+
+if (sampling_approach == 'regular'){
+  output <- time_series %>% map_dbl(function(df) sample_regular(df,sample_size)$changepoint) %>% mutate(params,detection=.)
+}else if (sampling_approach == 'random'){
+  output <- time_series %>% map_dbl(function(df) sample_random(df,sample_size)$changepoint) %>% mutate(params,detection=.)
+}else if (sampling_approach == 'iterative'){
+  output <- time_series %>% map_dbl(function(df) sample_iterative(df,sample_size)$changepoint) %>% mutate(params,detection=.)
+}else{
+    
+  }
+
   
   # Sample regular for each time series
-  output <- time_series %>%
-    map_dbl(function(df) sample_regular(df,sample_size)$changepoint) %>%
-    mutate(params,detection=.)
+#  output <- time_series %>%
+#    map_dbl(function(df) sample_regular(df,sample_size)$changepoint) %>%
+#    mutate(params,detection=.)
   
   names(output) = c('param_name','shift_time','detection')
   output$detection[is.na(output$detection)] <- 0
@@ -52,7 +63,7 @@ if (as.numeric(prob_correct) > power_level & gap_size > 1){
   
 } # End of while loop  
 
-if(sample_size>15 & sample_size<51){
+if(sample_size>10 & sample_size<75){
   return(as.numeric(sample_size))
 }else{
   return(NA)

### 2020-03-30 R/manuscript_draft.Rmd
made small edits to text in discussion section

@@
 
+<!--
 Outline:
@@
 - future work  (?)
+-->
 
 <!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
-There is often a difference between the number of samples analyzed and the total number of samples collected, due to time or funding limitations.
-When the processing time is disproportionally higher than the collection time especially, a subsampling can be done posterior to sampling.
+Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed.
+When the processing time is disproportionally higher than the collection time especially, a subsampling can be done prior to processing.
 A decision must then be taken of which subsamples to analyze.
@@
 \ref{fig:conceptual}).
-
 We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
@@
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
-
 Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
 \ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
-Same results for the major change, with only 12% of the samples analyzed.
 The method also work to detect further shifts (Fig.
 S4).
+
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
@@
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
-It matches very well the paleoecological approach [@Wingard2017].
+\easton{I'm not sure what this means.
+What do you mean by it ``matches"?}It matches very well the paleoecological approach [@Wingard2017].
 However, both different types of data and different questions can be addressed with the same approach.
@@
 
-Our approach is able to address other data sources and questions, but not all.
-When processing time and resources are less than the collection time, investigators will be likely to process every sample.
-If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Figs.
+Our approach is applicable to a wide range of systems and questions, but it does have limitations.
+When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample.
+When sample analysis costs are high, then regular and interative sampling are better than random sampling (Figs.
 \ref{fig:simulation_approaches_plot},S2,S3).
-More generally, some questions require a continuous time series, or at least a regular sampling interval.
+However, some questions require a continuous time series, or at least a regular sampling interval.
 For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016].
@@
 Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
-
-\rosalie{I don't understand what you meant here. I tried reformulating some of it, but we can talk about it more if needed}:  
-\st{The specific number of samples and optimal sampling approach will largely be system specific.
-A similar system may be used to get some sense for the best sampling scheme, but it will depend on the system dynamics (e.g.
-interannual variation, trend effect size).
-The specific sampling scheme also depends on the question of interest.
-For instance, changepoint detection may require fewer samples than detecting a change in biodiversity indices (citations).
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-However, more samples could be processed later to address these additional questions.}
+However, more samples could be processed later to address these additional questions.
 
@@
 We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
-This is valuable given the high costs of many monitoring programs [@Caughlan2001a,Bennett2014a].
+This is valuable given the high costs of many monitoring programs [@Caughlan2001a,@Bennett2014a].
 Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
@@
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
-We make the point that if a subsample of the samples must be used, the temptation to be unbiased by using a regular subsampling will lead to losing a lot of accuracy in the process.
-If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable.
+If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling.
+This can improve the accuracy of the results and reduce costs overall.
+If the objective is to detect a changepoint, an iterative approach is optimal.
 

### 2020-03-31 R/manuscript_draft.Rmd
added figure caption and corrected small spelling mistakes

@@
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a].
-However, environmental monitoring is still often expensive and time-consuming, especially when further procesing is needed following sample collection (citations).
-Thus, for many fields there is a disparity between the amount of data that can be aquired and stored, and the ultimate number of samples that can be processed.
+However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection (citations).
+Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed.
 Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
@@
 The specific number of years required depended on the species biology and the detection method used [@White2019d].
-Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitave change in population.
+Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population.
 Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
@@
 There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
-We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of changepoints.
+We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints.
 Herein, we focus on detecting the most important changepoint (i.e.
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in commuper total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
@@
 Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed.
-When the processing time is disproportionally higher than the collection time especially, a subsampling can be done prior to processing.
+When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing.
 A decision must then be taken of which subsamples to analyze.
@@
 When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample.
-When sample analysis costs are high, then regular and interative sampling are better than random sampling (Figs.
+When sample analysis costs are high, then regular and iterative sampling are better than random sampling (Figs.
 \ref{fig:simulation_approaches_plot},S2,S3).
@@
 For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016].
-However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017].
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
+However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in addressing the driver.
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.

### 2020-03-31 R/supp_material.Rmd
added figure caption and corrected small spelling mistakes

@@
 
-For each sampling approach, we wrote three custom functions:
+We first detect the ``true" changepoint of a full length time series with the function _e.divisive_ in the R package _ecp_ [@James2019].
+We focus on the changepoint with the largest magnitude, altough this package allows lower order changepoint detection as well (Fig.
+\ref{fig:case_study_other_changepoints}).
+We then subsample the full time series [@White2020] with different numbers of subsamples and different sampling approaches.
+For each sampling approach, we wrote a custom function:
 
@@
 Detection of changepoint and addition of sample is repeated, until it was narrowed down to two consecutive samples, or n (maximum number of samples, chosen by the user) was reached, whichever comes first.
+
+The subsampled time series are then compared to the full time series to assess the effectiveness of each subsampling approach.
 
@@
 source('R/sampling_random.R')
-source('R/sampling_regular.R')
-source('R/sampling_iterative.R')
 source('R/build_time_series.R')
@@
              labeller = label_glue('a. {Parameter}')) +
-  ylim(10,50) +
+  ylim(10,52) +
   labs(x=c("sigma"), y= "Minimum number of samples required") + 
@@
              labeller = label_glue('b. {Parameter}')) +
-  ylim(10,50) +
+  ylim(10,52) +
   labs(x="phi", y= "") + 
@@
              labeller = label_glue('c. {Parameter}')) +
-  ylim(10,50) +
+  ylim(10,52) +
   labs(x="shift size", y= "") + 
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8}
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8,fig.cap='Distance to true (a) first changepoint (largest magnitude), (b) second changepoint, and (c) third changepoint for total number of samples analyzed, following random sampling, regular sampling, and iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_other_changepoints}'}
 library(ggplot2)

### 2020-03-31 R/manuscript_draft.Rmd
need to commit before push

--- 2020-03-30 R/manuscript_draft.Rmd
@@
 ---
-bibliography: ChangepointSampling.bib
+bibliography: Whitebib.bib
 csl: ecology-letters.csl
@@
 Thus, for many fields there is a disparity between the amount of data that can be aquired and stored, and the ultimate number of samples that can be processed.
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.<!-- processing methods that could allow to extract the most information from that data.
 \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
@@
 Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a,@McDonald-Madden2010a,@Lindenmayer2020a].
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@McDonald-Madden2010a,@White2019d,@Lindenmayer2020a].
 -->
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
+For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions (Cohen, 2003).
 Similarly, environmental samples (e.g.
 water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a].
-Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a,@Mallet2014].
+Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later (citations).
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
@@
 We tailor our analysis to the detection of a changepoint in a time series.
-Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes [@James2014].
+Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes (citations).
 We focus on paleoecological core samples as one example of this type of data.
@@
 We first investigate these tools using a simulation-based approach.
-We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy [@Bruel2018].
+We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy (citation).
 
@@
 In the context of sediment cores, this would mean analyzing community composition at random locations along the core.
-Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018].
+Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) (Nad’o and Kaňuch, 2018).
 We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
@@
 
-We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019].
-There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
+We detect changepoints with the function _e.divisive_ in the R package _ecp_ (James et al 2019).
+There are several methods available to detect changepoint (reviewed in James and Matteson, 2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis.
 We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of changepoints.
 Herein, we focus on detecting the most important changepoint (i.e.
-the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4).
+the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Supplementary Materials 1).
 In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one.
@@
 
+```{r theoretical_results_old,message=F,warning=F,echo=F,eval=F}
+
+source('R/sampling_regular.R')
+source('R/sampling_random.R')
+source('R/build_time_series.R')
+source('R/updated_param_sensitivity.R')
+
+require(tidyverse)
+
+par(mfrow=c(2,2))
+
+```
+
 ```{r simulation_min_time,message=F,warning=F,echo=F,eval=F}
@@
 # Sigma results
-sigma_vec = seq(0.01,0.4,0.025)
+sigma_vec = seq(0.01,1,0.45)
 min_time  = rep(0,times=length(sigma_vec))
@@
 for (param_num in 1:nrow(sigma_output)){
-  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'regular')
+  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
   print(sigma_output)
@@
 
-phi_vec = seq(-0.8,0.8,0.05)
+phi_vec = seq(-0.8,0.8,0.2)
 min_time  = rep(0,times=length(phi_vec))
@@
 for (param_num in 1:nrow(phi_output)){
-  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num],'regular')
+  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
   print(phi_output)
@@
 for (param_num in 1:nrow(shift_size_output)){
-  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num],'regular')
+  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
   print(shift_size_output)
@@
 ```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
-Regular sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+Random sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
 Vertical line indicate respective parameter calculated from the case study time series.
-The same analyses for random and iterative sampling is in Figs.
-S2,S3.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}
+\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
 
@@
   geom_line() +
-  geom_vline(xintercept=0.05, lty=2) + 
+  geom_vline(xintercept=ar_vr, lty=2) + 
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('a. {Parameter}')) +
-  ylim(10,50) +
+  ylim(20,50) +
   labs(x=c("sigma"), y= "Minimum number of samples required") + 
@@
   geom_line() +
-  geom_vline(xintercept = 0.404, lty=2) +
+  geom_vline(xintercept = cv_vr, lty=2) +
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('b. {Parameter}')) +
-  ylim(10,50) +
+  ylim(20,50) +
   labs(x="phi", y= "") + 
@@
   geom_line() + 
-  geom_vline(xintercept = 0.81, lty=2) +
+  geom_vline(xintercept = shift_vr, lty=2) +
   facet_wrap(~Parameter, scales="free_x", 
              labeller = label_glue('c. {Parameter}')) +
-  ylim(10,50) +
+  ylim(20,50) +
   labs(x="shift size", y= "") + 
@@
 Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
-\ref{fig:simulation_min_time_plot}).
+\ref{fig:simulation_min_time}).
 For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
@@
 
-```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_approaches_plot}'}
+```{r simulation_approaches_graph, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_number_sample}'}
 
@@
 To obtain this data sediment cores were taken from aquatic environments to reconstruct past ecosystems.
-Samples can be taken before and after a perturbation to compare community composition, or continuously to identify the timing of change and underlying mechanisms of the change [@Perga2015].
+Samples can be taken before and after perturbation to compare community composition (e.g., ), or at continuously to identify the timing of change, and contribution of environmental variable in driving these changes (e.g., @Perga2015).
 
@@
 Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999,@Bruel2018].
-Air temperature is now driving changes in plankton communities [@Bruel2018].
+Air temperature is now driving changes in plankton communities (@Bruel2018).
 
@@
 # Read data
-
-require(stickylabeller)
-
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
@@
 
-```{r echo=FALSE, eval=F,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
@@
 myoutput_count <- NULL
+j=3
 for (i in 5:30) {
@@
 We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
-\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3).
+\ref{simulation_number_sample} and \ref{fig:case_study_number_sample}).
 Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.
 (Fig.
-\ref{fig:simulation_min_time_plot}).
+\ref{fig:simulation_min_time}).
 
@@
 \ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig.
-\ref{fig:simulation_approaches_plot}c).
+\ref{fig:simulation_min_time}c).
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
@@
 Same results for the major change, with only 12% of the samples analyzed.
-The method also work to detect further shifts (Fig.
-S4).
+The method also work to detect further shifts (Supplementary Materials 1).
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019,@White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
@@
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
-It matches very well the paleoecological approach [@Wingard2017].
+It matches very well the paleoecological approach (Wingard et al, 2017).
 However, both different types of data and different questions can be addressed with the same approach.
 Suppose instead that the goal was to detect a change in observed individuals with video-based approaches.
-It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
+It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address our question of interest (citations).
 Using an interval sampling approach, one could take a fixed number of samples to start.
@@
 Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
-These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
+These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples (citations).
 
@@
 <!---- how this doesn't fit with other work (limitations) ---->
-<!-- need line that references issue that iterative sampling can get stuck on incorrect points-->
 
-Our approach is able to address other data sources and questions, but not all.
+Our approach is able to address other data sources and questions, but not any.
 When processing time and resources are less than the collection time, investigators will be likely to process every sample.
-If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Figs.
-\ref{fig:simulation_approaches_plot},S2,S3).
-More generally, some questions require a continuous time series, or at least a regular sampling interval.
-For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016].
-However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017].
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
+If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Fig.
+\ref{simulation_number_sample}).
+More generally, some questions require a continuous time serie, or at least a regular sampling interval.
+For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts (Frossard et al 2015, Doncaster et al 2016).
+However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down (Spears et al, 2017).
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings (Doncaster et al 2016), but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.
 
@@
 We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
-This is valuable given the high costs of many monitoring programs [@Caughlan2001a,Bennett2014a].
+This is valuable given the high costs of many monitoring programs (citations).
 Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
-The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007,@Bennett2018b].
+The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach (@Lovett2007).
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.

### 2020-03-31 R/supp_material.Rmd
need to commit before push
first edits of the day to supp material

@@
 
-# Additional text
+# Additional methods
 
 We first detect the ``true" changepoint of a full length time series with the function _e.divisive_ in the R package _ecp_ [@James2019].
-We focus on the changepoint with the largest magnitude, altough this package allows lower order changepoint detection as well (Fig.
+We focus on the changepoint with the largest magnitude, altough this package allows detection of further changepoints as well (Fig.
 \ref{fig:case_study_other_changepoints}).
@@
 The subsampled time series are then compared to the full time series to assess the effectiveness of each subsampling approach.
+
+Each function works with two types of input: vector and matrix.
+If the input is a matrix, the user must change the default argument *is_vector* to *FALSE*.
+The matrix is transformed to independent vectors using Detrented Correspondance Analysis [@hill_detrended_1980], and a single axis on which changepoint analysis is run is chosen with the argument *DCA_axis* (default to 1, for first component).
 
@@
 
+## Case study: biomass
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample}'}
+#investigating species average biomass
+vr_count <- rowSums(vr[,-1])
+myoutput_count <- NULL
+j=3
+for (i in 5:30) {
+  vrsub <- sample_regular(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_random(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 6)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 7)
+  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
+}
+
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese_biomass.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Plot
+p1 <-
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint")
+
+p1
+
+```
+
+(Fig. \ref{fig:case_study_number_sample_change_biomass})
+
 \clearpage

### 2020-03-31 R/manuscript_draft.Rmd
fixed merge conflict again

--- 2020-03-31 R/manuscript_draft.Rmd
@@
 
-```{r echo=FALSE, eval=F,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Total number of samples was set between 5 and 30, out of the 74 initial time series.
-\\label{fig:case_study_number_sample}'}
-#investigating species average biomass
-vr_count <- rowSums(vr[,-1])
-myoutput_count <- NULL
-for (i in 5:30) {
-  vrsub <- sample_regular(vr_count, i, input_vector = T, messages = F)
-  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
-  vrsub <- sample_random(vr_count, i, input_vector = T, messages = F)
-  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
-  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F)
-  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
-  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 6)
-  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
-  vrsub <- sample_iterative(vr_count, i, input_vector = T, messages = F, n2 = 7)
-  myoutput_count <- c(myoutput_count, vrsub$method, vrsub$final_n, vrsub$changepoint_all$order.found[j], length(vrsub$changepoint_all$estimates)-2)
-}
-
-myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese_biomass.txt"))
-myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
-myoutput$diff_real <- abs(myoutput$diff_real)
-
-# Plot
-p1 <-
-  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method,
-             labeller = label_glue('{.l}. {method} sampling ')) + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        strip.background = element_rect(colour="transparent", fill="transparent"),
-        strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint")
-
-p1
-
-```
-
-\rosalie{About figure \ref{fig:case_study_number_sample_change_biomass}, I tried testing the real changepoint on another potential metric, i.e., biomass.
-Potential tools for discussion?
-I don't know!
-Maybe worth talking about that during our next Skype.}
-
 # General Discussion

### 2020-03-31 R/supp_material.Rmd
work on supp materials

--- 2020-03-31 R/supp_material.Rmd
@@
 
-# Parameter sensitivity
+Each function works with two types of input: vector and matrix.
+If the input is a matrix, the user must change the default argument *is_vector* to *FALSE*.
+The matrix is then transformed to independent vectors using Detrented Correspondance Analysis [@hill_detrended_1980], and a single axis on which changepoint analysis is run is chosen with the argument *DCA_axis* (default to 1, for first component).
+Component scores are returned by the funtion.
+User can edit the function to use another ordination method (e.g., principal component analysis).
+
+# Method's sensitivity for varying parameters
+
+We investivated how different parameters change the power of the analysis, and found that autocorrelation at lag-1 (phi) had little impact on the power of the analysis amplitude of the shift was the most important determinant of the power of the analysis.
+For null standard deviation, the analysis performed well independently of the shift size or autocorrelation.
+When standard deviation increased, the power of the analysis was higher for null autocorrelation (Fig.
+\ref{fig:parameter_sensitivity_plot}).
 
@@
 
-# Case study
+# Detecting further changepoints
+
+The initial study found 3 changepoints on the main axis, in 1926/1928, 1946/1948 and 1983/1988 [@Bruel2018].
+We focused our analysis on the main changepoint, but the code also allows to target further changepoints, using the argument *c*.
+*c* is a numeric value indicating the maximum number of changepoints to find (Fig.
+\ref{fig:case_study_other_changepoints}).
+
+The iterative method requires overall less subsamples to find the real changepoints (Fig.
+\ref{fig:case_study_other_changepoints}, 3rd column).
+Note that if the initial number of samples (controled with the argument *n2* in the code) is at the default 5, the code struggles to rank appropriately the changepoints, resulting in a changepoint being found, but not the overall "true" second  (e.g., Fig.
+\ref{fig:case_study_other_changepoints}b, 3rd column), or "true" third.
+In a time-series, we recommend users to examinethe p-values returned by the *e.divisive* function, embeded in the funtion *sample_iterative* output.
+If the first few changepoints have close, low, p-values, it may be worth to not directly pursue a subsample close to the first changepoint, but also add another subsample further away.
+An alternative is to initiate the subsampling with an higher number of samples, e.g., 7 subsamples.
 
@@
 
+## Case study: biomass
+
+```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 3, fig.cap='Number of individuals per gram of sediment, infered from Cladocera remains for the 74 subsamples of the Lake Varese sediment core.
+\\label{fig:case_study_biomass_per_year}'}
+# Investigating change in species average biomass
+# Read code
+source('R/sampling_regular.R')
+# Read data
+getpath4data <- function() {
+  if(Sys.getenv("USER")=="Rosalie") return("/Volumes/-/Script R/Data-off-GitHub/temporal-sampling/")
+  if(Sys.getenv("USER")=="eastonwhite") return("~/Desktop/Research/soil-temporal-sampling/")
+  if(Sys.getenv("USER")!="Rosalie"|Sys.getenv("USER")!="put here your USER") stop("You need to get the data and define their location.")
+}
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+vr_count <- data.frame(Year=vr[,1], Biomass=rowSums(vr[,-1]))
+cpvar_count <- sample_regular(vr_count[,2], nrow(vr_count), input_vector = T, messages = F)
+
+ggplot(vr_count, aes(Year, Biomass)) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent")) +
+           geom_rect(mapping=aes(xmin=vr_count$Year[cpvar_count$changepoint-1], xmax=vr_count$Year[cpvar_count$changepoint], ymin=0, ymax=max(vr_count)), alpha=0.05, fill="grey") + 
+  geom_line() +
+  labs(y="Number of individuals / g")
+
+```
+
+For further ilustration, we reproduced the approach on a single vector.
+We re-used the community dataset from Lake Varese, but this time, looked at the number of individuals per gram of sediment in each subsamples.
+The full time series show a changepoint in `r paste(vr_count$Year[cpvar_count$changepoint], vr_count$Year[cpvar_count$changepoint-1], sep="/")`.
+We used the three changepoint method to estimate changepoint on the vector, and found that the iterative method, with an initial number of 5 regularly-spaced subsamples, performed best (Fig.
+\ref{fig:case_study_number_sample_biomass}).
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in biomass per total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample_biomass}'}
+
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese_biomass.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Plot
+p1 <-
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint")
+
+p1
+
+```
+
 \clearpage

### 2020-04-01 R/supp_material.Rmd
References should be a title of order 1 not 2 -- fixed
change title subsection
other editorial edits

@@
 
-# Method's sensitivity for varying parameters
+# Method sensitivity for varying parameters
 
@@
 
-## Case study: biomass
+# Case study: changepoint on a vector of number of individuals
 
-```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 3, fig.cap='Number of individuals per gram of sediment, infered from Cladocera remains for the 74 subsamples of the Lake Varese sediment core.
-\\label{fig:case_study_biomass_per_year}'}
+```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
 # Investigating change in species average biomass
@@
 
+```
+
+For further ilustration, we reproduced the approach on a single vector.
+We re-used the community dataset from Lake Varese, but this time, looked at the number of individuals per gram of sediment in each subsamples.
+The full time series show a changepoint in `r paste(vr_count$Year[cpvar_count$changepoint], vr_count$Year[cpvar_count$changepoint-1], sep="/")`.
+We used the three changepoint method to estimate changepoint on the vector, and found that the iterative method, with an initial number of 5 regularly-spaced subsamples, performed best (Fig.
+\ref{fig:case_study_number_sample_biomass}).
+
+```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 3, fig.cap='Number of individuals per gram of sediment, infered from Cladocera remains for the 74 subsamples of the Lake Varese sediment core.
+\\label{fig:case_study_biomass_per_year}'}
+# plot
 ggplot(vr_count, aes(Year, Biomass)) + 
@@
 ```
-
-For further ilustration, we reproduced the approach on a single vector.
-We re-used the community dataset from Lake Varese, but this time, looked at the number of individuals per gram of sediment in each subsamples.
-The full time series show a changepoint in `r paste(vr_count$Year[cpvar_count$changepoint], vr_count$Year[cpvar_count$changepoint-1], sep="/")`.
-We used the three changepoint method to estimate changepoint on the vector, and found that the iterative method, with an initial number of 5 regularly-spaced subsamples, performed best (Fig.
-\ref{fig:case_study_number_sample_biomass}).
 
@@
 
-## References
+# References

### 2020-04-01 R/manuscript_draft.Rmd
edits to manuscript

@@
 Total number of samples was set between 5 and 30, out of the 74 initial time series.
-\\label{fig:case_study_number_sample_change_biomass}'}
+\\label{fig:case_study_number_sample}'}
 # Read data
@@
 \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3).
-Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.
-(Fig.
+Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion (Fig.
 \ref{fig:simulation_min_time_plot}).
@@
 
+The approach is versatile: we converted the community data in a single vector of number of individuals per gram of sediment from the same sediment core (Fig.
+S5).
+We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
+S6).
+The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
+\ref{fig:case_study_number_sample} (c)).
+However, depending on the metric the user is interested in, the subsampling will target different subsamples to add.
+In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.
+
 Our approach goes beyond just paleoecological analyses.
@@
 \easton{I'm not sure what this means.
-What do you mean by it ``matches"?}It matches very well the paleoecological approach [@Wingard2017].
+What do you mean by it ``matches"?} \rosalie{is it clearer that way?
+I changed "match" to "corresponds" and added more details.} It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017].
 However, both different types of data and different questions can be addressed with the same approach.
@@
 These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
-
-The approach is versatile: it also finds in 9 samples (11 to confirm) the change in species biomass (Fig.
-\ref{fig:case_study_number_sample_change_biomass}).
-The initial 5 samples analyzed are the same: a well tailored question from the start allows to reach the changepoint much more efficiently than random and regular subsampling methods.
-(\rosalie{do we want to include that?
-not sure})
 

### 2020-04-01 R/manuscript_draft.Rmd
made small changes to text throughout and added citations

@@
 ---
-bibliography: ChangepointSampling.bib
+bibliography: [ChangepointSampling.bib, packages.bib]
 csl: ecology-letters.csl
@@
 header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
-  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount} \usepackage{soul}
+  \usepackage[numbers]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount} \usepackage{soul}
   \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \renewcommand{\thefootnote}{\fnsymbol{footnote}} \addtocounter{footnote}{-1} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
@@
     number_sections: no
+link-citations: yes
 ---
@@
 knitr::opts_knit$set(root.dir = dirname(getwd()))
+# Optional code below to include package citations
+# knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
+# Use [@R-dplyr; @R-rmarkdown] in main text
 ```
@@
 
+\linenumbers
+
 \textbf{Abstract}
 
-Environmental monitoring is a key component of managing ecosystems.
+Environmental monitoring is a key component of understanding and managing ecosystems.
 Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective.
@@
 We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power.
-We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling, and where the objective is to detect ecosystem change.
+We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.
 
@@
 
-Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a].
+Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019d; @Lindenmayer2020a].
 Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a].
-However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection (citations).
+However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012].
 Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed.
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
 \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
@@
 Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a,@McDonald-Madden2010a,@Lindenmayer2020a].
+Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a;@McDonald-Madden2010a;@Lindenmayer2020a].
 -->
@@
 Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population.
-Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
+Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
-The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].
+Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].
 
@@
 water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a].
-Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a,@Mallet2014].
-In each of these cases, decisions have to made about how much data to extract from the previously collected samples.
+Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014].
+In each of these cases, decisions have to made about how much data to extract from the previously collected samples [@Zhang2012].
 Should the paleoecological core be analyzed at every centimeter?
@@
 Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018].
-We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint.
-Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig.
+We hypothesize that random sampling will perform the worst in estimating the changepoint.
+Regular sampling is commonly used (e.g.
+@Bruel2018) and requires that a set number of samples be taken at regular intervals (Fig.
 \ref{fig:conceptual}d).
@@
 \ref{fig:conceptual}e-g).
-In each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period.
+For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period.
 We describe each approach in more detail in the supplementary material and provide code.
@@
 the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4).
-In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one.
-The distance to real changepoint served as performance diagnostic.
+In order to test the performance, we detected the "true" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "true" one.
+The distance to true changepoint served as the performance diagnostic.
 
@@
 We simulate an entire time series to serve as the "true" data for comparison [@White2020].
-We specifically study how the amount of samples included and the type of sampling affects the detection probability.
+We specifically study how the number of samples and the type of sampling affects the detection probability.
 For simulations, statistical power is the fraction of simulations that were able to detect a changepoint.
 We define an accurate changepont detection if an estimate is within five time points (over 100 time points) of the true changepoint.
-We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.
+The minimum number of samples required is the number needed for 0.8 statistical power.
 
@@
 source('R/build_time_series.R')
-source('R/updated_param_sensitivity.R')
+#source('R/updated_param_sensitivity.R')
 source('R/calc_min_time2.R')
@@
 
-# 2 - Calculate autocorrelation of real time serie
+# 2 - Calculate autocorrelation of true time serie
 ar_vr <- acf(vr_time_series, lag.max = 5, plot = FALSE)$acf[2]
@@
 
-In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$).
+In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$).
 We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size.
-Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
+Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
 \ref{fig:simulation_min_time_plot}).
@@
 
-```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_approaches_plot}'}
+```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time_plot}.
+The error bars represent the middle 95% of the simlations.\\label{fig:simulation_approaches_plot}'}
 
@@
         strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint") #+
+  labs(x="Total number of samples analyzed", y="Distance to true changepoint") #+
  # theme(legend.position = c(0.93, 0.72),
@@
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
-It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref).
-Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999,@Bruel2018].
+It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed.
+Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018].
 Air temperature is now driving changes in plankton communities [@Bruel2018].
@@
 Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
-\ref{fig:case_study_number_sample} (c)).
+\ref{fig:case_study_number_sample}c).
 We show how iterative sampling changes the ordination, and the resulting changepoint, as more samples are added (Fig.
 \ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig.
-\ref{fig:case_study_number_sample} (c), square markers), the distance to the real changepoint was initially very high.
+\ref{fig:case_study_number_sample} (c), square markers), the distance to the true changepoint was initially very high.
 Without any prior, the initial number sampled fell far away from the true changepoint.-->
@@
         strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint") 
+  labs(x="Total number of samples analyzed", y="Distance to true changepoint") 
 }
@@
 When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing.
-A decision must then be taken of which subsamples to analyze.
-To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig.
+A decision must then be made as to which subsamples to analyze.
+To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at random points, random intervals, and an iterative sampling approach (Fig.
 \ref{fig:conceptual}).
-We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
+We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
 \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3).
-Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion (Fig.
+Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig.
 \ref{fig:simulation_min_time_plot}).
@@
 Multiple subsampling strategies can be chosen (Fig.
-\ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig.
+\ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig.
 \ref{fig:simulation_approaches_plot}c).
 Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
-Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig.
-\ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples.
-The method also work to detect further shifts (Fig.
+Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig.
+\ref{fig:case_study_number_sample}).
+The method also worked well to detect other changepoints of lower magnitude (Fig.
 S4).
-
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
 Each sample took an average of 3 hours to process.
-We found that using an interval approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
+We found that using an interative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
 This correspond to several thousands of US dollars depending on labor costs.
@@
 
-The approach is versatile: we converted the community data in a single vector of number of individuals per gram of sediment from the same sediment core (Fig.
+As another application of our approach, we examined the same lake core data, but examined total abundance as opposed to community composition (Fig.
 S5).
@@
 The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
-\ref{fig:case_study_number_sample} (c)).
-However, depending on the metric the user is interested in, the subsampling will target different subsamples to add.
+\ref{fig:case_study_number_sample}c).
+However, depending on the metric of interest, the subsampling will target different subsamples to add.
 In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d;@White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred.
+Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
 Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
-\easton{I'm not sure what this means.
-What do you mean by it ``matches"?} \rosalie{is it clearer that way?
-I changed "match" to "corresponds" and added more details.} It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017].
+It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017].
 However, both different types of data and different questions can be addressed with the same approach.
-Suppose instead that the goal was to detect a change in observed individuals with video-based approaches.
+Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
 It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
@@
 \ref{fig:simulation_approaches_plot},S2,S3).
-However, some questions require a continuous time series, or at least a regular sampling interval.
-For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016].
+In addition, some questions require a continuous time series, or at least a regular sampling interval.
+For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015;@Doncaster2016].
 However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in addressing the driver.
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implication in addressing the driver.
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
@@
 Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
-For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014).
-We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint.
-This is valuable given the high costs of many monitoring programs [@Caughlan2001a,@Bennett2014a].
+We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series.
 Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
-The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007,@Bennett2018b].
+In addition, optimizing sampling efforts is valuable given the high costs of many monitoring programs [@Caughlan2001a;@Bennett2014a].
+Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b].
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
@@
 This can improve the accuracy of the results and reduce costs overall.
-If the objective is to detect a changepoint, an iterative approach is optimal.
 

### 2020-04-01 R/supp_material.Rmd
again, tiny errors that slipped through
made small changes to text throughout and added citations

@@
 
-Data and code for all the figures can be found at (https://github.com/rosalieb/temporal-sampling.git). All the analyses were run in R (citation).
+Data and code for all the figures can be found at (https://github.com/rosalieb/temporal-sampling). All the analyses were run in R [@RCoreTeam2019].
 
@@
 
-We first detect the ``true" changepoint of a full length time series with the function _e.divisive_ in the R package _ecp_ [@James2019].
-We focus on the changepoint with the largest magnitude, altough this package allows detection of further changepoints as well (Fig.
+We first detect the "true" changepoint of a full length time series with the function _e.divisive_ in the R package _ecp_ [@James2019].
+We focus on the changepoint with the largest magnitude, although this package allows detection of further changepoints as well (Fig.
 \ref{fig:case_study_other_changepoints}).
@@
 If the input is a matrix, the user must change the default argument *is_vector* to *FALSE*.
-The matrix is then transformed to independent vectors using Detrented Correspondance Analysis [@hill_detrended_1980], and a single axis on which changepoint analysis is run is chosen with the argument *DCA_axis* (default to 1, for first component).
-Component scores are returned by the funtion.
-User can edit the function to use another ordination method (e.g., principal component analysis).
+The matrix is then transformed to independent vectors using Detrended Correspondence Analysis [@Hill1980], and a single axis on which changepoint analysis is run is chosen with the argument *DCA_axis* (default to 1, for first component).
+Component scores are returned by the function.
+A user can edit the function to use another ordination method (e.g., principal component analysis).
 
-# Method sensitivity for varying parameters
+# Parameter sensitivity
 
-We investivated how different parameters change the power of the analysis, and found that autocorrelation at lag-1 (phi) had little impact on the power of the analysis amplitude of the shift was the most important determinant of the power of the analysis.
+We investigated how different parameters change the power of the analysis, and found that autocorrelation at lag-1 (phi) had little impact on the power of the analysis amplitude of the shift was the most important determinant of the power of the analysis.
 For null standard deviation, the analysis performed well independently of the shift size or autocorrelation.
@@
 
-for (param_num in 1:nrow(sigma_output)){
+for (param_num in 11:nrow(sigma_output)){
   sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'iterative')
-  #print(sigma_output)
+  print(sigma_output)
 }
@@
 
-The iterative method requires overall less subsamples to find the real changepoints (Fig.
+The iterative method requires less subsamples overall to find the real changepoints (Fig.
 \ref{fig:case_study_other_changepoints}, 3rd column).
-Note that if the initial number of samples (controled with the argument *n2* in the code) is at the default 5, the code struggles to rank appropriately the changepoints, resulting in a changepoint being found, but not the overall "true" second  (e.g., Fig.
+Note that if the initial number of samples (controlled with the argument *n2* in the code) is at the default 5, the code struggles to rank appropriately the changepoints, resulting in a changepoint being found, but not the overall "true" second  (e.g., Fig.
 \ref{fig:case_study_other_changepoints}b, 3rd column), or "true" third.
-In a time-series, we recommend users to examinethe p-values returned by the *e.divisive* function, embeded in the funtion *sample_iterative* output.
-If the first few changepoints have close, low, p-values, it may be worth to not directly pursue a subsample close to the first changepoint, but also add another subsample further away.
+For a time-series, we recommend users examine the p-values returned by the *e.divisive* function, embedded in the function *sample_iterative* output.
+If the first few changepoints have close, low p-values, it may be useful to add another subsample further away from the presumed changepoint.
+This avoids the iterative algorithm mistakenly converging on the wrong changepoint.
 An alternative is to initiate the subsampling with an higher number of samples, e.g., 7 subsamples.
@@
 
-# Case study: changepoint on a vector of number of individuals
+# Case study: Changepoint detection of abundance time series
 
@@
 
-For further ilustration, we reproduced the approach on a single vector.
-We re-used the community dataset from Lake Varese, but this time, looked at the number of individuals per gram of sediment in each subsamples.
+As another example of our approach, we we re-used the community dataset from Lake Varese, but this time, looked at the total number of individuals per gram of sediment in each subsamples.
+Thus, this is a univariate time series as opposed to the community-level data used in the main text.
 The full time series show a changepoint in `r paste(vr_count$Year[cpvar_count$changepoint], vr_count$Year[cpvar_count$changepoint-1], sep="/")`.
@@
 ```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width = 4, fig.height= 3, fig.cap='Number of individuals per gram of sediment, infered from Cladocera remains for the 74 subsamples of the Lake Varese sediment core.
+Vertical grey band indicates the "true" changepoint.
 \\label{fig:case_study_biomass_per_year}'}

### 2020-04-01 R/manuscript_draft.Rmd
adding arkilanian etal 2020 paper

@@
 In addition, some questions require a continuous time series, or at least a regular sampling interval.
-For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015;@Doncaster2016].
-However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
+For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016], although recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@arkilanian_effect_nodate].
+Extending on the question of being able to detect early warning signals, this objective must be weighed in with the ultimate goal: detecting ecosystem shift.
+Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
 Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implication in addressing the driver.
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-However, more samples could be processed later to address these additional questions.
+Finally, more subsamples could be processed later to address additional questions.
 

### 2020-04-06 R/manuscript_draft.Rmd
edit to the text

@@
 # Optional code below to include package citations
-# knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
+ knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
 # Use [@R-dplyr; @R-rmarkdown] in main text
@@
 We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change.
-We test these ideas with a case study of Cladocera community assemblage, which was reconstructed from a sediment core.
-We demonstrate that standard approaches to sample processing are inefficient.
+We test these ideas with a case study of Cladocera community assemblage reconstructed from a sediment core.
+We demonstrate that standard approaches to sample processing are less time efficient than an iterative approach.
 For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs.
@@
 Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population.
-Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a].
+Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the benefit of increasing sampling breadth relying on citizen science [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
+For example, sediment cores can be retrieved from aquatic ecosystem with little sediment disturbances such as lakes or lagoon, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
 Similarly, environmental samples (e.g.
@@
 We hypothesize that random sampling will perform the worst in estimating the changepoint.
-Regular sampling is commonly used (e.g.
-@Bruel2018) and requires that a set number of samples be taken at regular intervals (Fig.
+Regular sampling is commonly used (e.g., pigments in @Milan2015) and requires that a set number of samples be taken at regular intervals (Fig.
 \ref{fig:conceptual}d).
@@
 
-As one application of our approach, we examined paleosequence data.
-To obtain this data sediment cores were taken from aquatic environments to reconstruct past ecosystems.
-Samples can be taken before and after a perturbation to compare community composition, or continuously to identify the timing of change and underlying mechanisms of the change [@Perga2015].
+We examined the performance of our approach on detecting changepoint from a paleosequence.
+Paleolimnology allows to reconstruct past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
+A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).
 
-We tested different sampling methods on a real community time-series, from Lake Varese (IT).
+We tested different sampling methods on a real community time-series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
@@
 In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
-We used the same ordination method as Bruel et al (2018) to reduce the dimensionality of the assemblage-level data.
-As a result, we detected several changepoints on the first component (in 1926/1928, 1946/1948 and 1983/1988).
+Our objective was to evaluate whether the same changepoints could be identified using less samples.
+In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
+Changepoints were then detected on the first component (46% of the total variability) in 1926/1928, 1946/1948 and 1983/1988.
 We defined these as the "true" changepoints given they came from an analysis of the complete data.
-The 1983-1988 changepoint was the largest in magnitude.
+The 1983/1988 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
+We identified second and third changepoints in Supplementary Materials.
 
 In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative).
+These subsamples were from the initial community dataset (Fig.
+\ref{fig:case_study_iterative_sampling}a).
+We reduced the dimensionality of the assemblage-level data to ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
+In the case of the iterative method, a new sample was added and the ordination was run again (Fig.
+\ref{fig:case_study_iterative_sampling}d).
 For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.
-
-Each subsampling was done from the initial community dataset (Fig.
-\ref{fig:case_study_iterative_sampling}a).
-An ordination was run on the number of samples selected, and the changepoint analysis was run on that new vector.
-In case of the iterative method, a new sample was added and the ordination was run again (Fig.
-\ref{fig:case_study_iterative_sampling}d).
 
@@
 \ref{fig:case_study_number_sample}c).
-We show how iterative sampling changes the ordination, and the resulting changepoint, as more samples are added (Fig.
+We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig.
 \ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig.
@@
 We tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
-Iterative sampling still performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials \textcolor{red}{2}).
+Iterative sampling still performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials S5).
 <!--This could probably be moved to the discussion: The initial number of samples should be chosen depending on the objective of the study.
@@
 A decision must then be made as to which subsamples to analyze.
-To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at random points, random intervals, and an iterative sampling approach (Fig.
+To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling random points, regular intervals, and an iterative sampling approach (Fig.
 \ref{fig:conceptual}).
 We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
-\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3).
+\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S2, S3).
 Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig.
@@
 
-As another application of our approach, we examined the same lake core data, but examined total abundance as opposed to community composition (Fig.
+As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
 S5).
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d;@White2020].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d; @White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
 Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
-Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint.
 It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017].
-However, both different types of data and different questions can be addressed with the same approach.
+However, both different types of data and different questions than the one we focused on can be addressed with the same approach.
 Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
@@
 Our approach is applicable to a wide range of systems and questions, but it does have limitations.
-When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample.
-When sample analysis costs are high, then regular and iterative sampling are better than random sampling (Figs.
-\ref{fig:simulation_approaches_plot},S2,S3).
-In addition, some questions require a continuous time series, or at least a regular sampling interval.
-For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016], although recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@arkilanian_effect_nodate].
+When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be prefered.
+We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs.
+\ref{fig:simulation_approaches_plot}, S2, S3).
+Another example where our method present less interest is when adressing questions that require a continuous time series, or at least a regular sampling interval.
+For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
+
+However recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@arkilanian_effect_nodate].
 Extending on the question of being able to detect early warning signals, this objective must be weighed in with the ultimate goal: detecting ecosystem shift.

### 2020-04-07 R/manuscript_draft.Rmd
Changed order of figures, added figure, and edited text

@@
 
-Authors: Rosalie Bruel$^{1*}$ and Easton R. White$^{2*}$
+Authors: Rosalie Bruel$^{1*}$ and Easton R. White$^{* \dagger 2,3}$
 
@@
 
-Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
-* Both authors contributed contributed equally and order was chosen alphabetically
+Addresses: 
+$^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, $^{2}$Department of Biology, University of Vermont, VT 05405, USA, $^{3}$Gund Institute for Environment, University of Vermont, VT 05405, USA, * Both authors contributed contributed equally and order was chosen alphabetically, $\dagger$Corresponding author
 
 \vspace{5 mm}
-Keywords: time-series, changepoint \vspace{10 mm}
+Keywords: time series, changepoint, monitoring, optimal sampling, Cladocera  \vspace{10 mm}
 
@@
 
-Number of references: \total{citenum}
+Number of references: <!--\total{citenum}-->
 
-Number of figures and panels: X
+Number of figures: 5
 
@@
 
-\textsc{Rosalie Bruel$^{*1}$\footnote{*The authors contributed equally to this work.} and Easton R. White$^{*2}$}
+\textsc{Rosalie Bruel$^{*1}$\footnote{*The authors contributed equally to this work. $\dagger$Corresponding author (eastonrwhite@gmail.com)} and Easton R. White$^{* \dagger 2,3}$}
 \vspace{3 mm}
 
-\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
+\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT, USA \\ $^{2}$Department of Biology, University of Vermont, VT, USA \\
+$^{3}$Gund Institute for Environment, University of Vermont, VT, USA}
 \end{center}
 
-\linenumbers
+<!--\linenumbers-->
+\singlespacing
 
@@
 Standard sub-sampling approaches often involve equally-spaced sampling.
-We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change.
+We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change.
 We test these ideas with a case study of Cladocera community assemblage reconstructed from a sediment core.
-We demonstrate that standard approaches to sample processing are less time efficient than an iterative approach.
+We demonstrate that standard approaches to sample processing are less efficient than an iterative approach.
 For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs.
@@
 
-Keywords: {time series}
+\textbf{Keywords:} time series, changepoint, monitoring, optimal sampling, Cladocera
 
@@
 
-<!--
-# Things to do:
-
-- clean up introduction
-- writing methods/results pair of section
-- build out the supplementary material
-- run theoretical results for more simulations
-- concluding paragraph
-- add citations
-- Possible journals: Journal of Environmental Management, Environmental Monitoring and Assessment, Ecological Applications
-
--->
-
 # Introduction
 
-Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019d; @Lindenmayer2020a].
+Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019e; @Lindenmayer2020a].
 Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007].
@@
 Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed.
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data.
-\rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths.
-With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at.
-If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}
-
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently.
-Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a;@McDonald-Madden2010a;@Lindenmayer2020a].
--->
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].
 
@@
 The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
-For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
-The specific number of years required depended on the species biology and the detection method used [@White2019d].
+For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
+The specific number of years required depended on the species biology and the detection method used [@White2019e].
 Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population.
@@
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
-For example, sediment cores can be retrieved from aquatic ecosystem with little sediment disturbances such as lakes or lagoon, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
+For example, sediment cores can be retrieved from aquatic ecosystems with little sediment disturbances, such as lakes or lagoon, allowing reconstruction of past ecological communities or conditions [@Cohen2003].
 Similarly, environmental samples (e.g.
 water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a].
-Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014].
+Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014a].
 In each of these cases, decisions have to made about how much data to extract from the previously collected samples [@Zhang2012].
@@
 
-Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added.
-We tailor our analysis to the detection of a changepoint in a time series.
+Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed.
+We tailor our analysis to the detection of a changepoint in a time series, but our approach is applicable to other questions as well.
 Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes [@James2014].
@@
 For simulations, statistical power is the fraction of simulations that were able to detect a changepoint.
-We define an accurate changepont detection if an estimate is within five time points (over 100 time points) of the true changepoint.
+We define an accurate changepont detection if an estimate is within five time points (given a time series of 100 time points) of the true changepoint.
 The minimum number of samples required is the number needed for 0.8 statistical power.
+
+```{r parameter_sensitivity,message=F,warning=F,echo=F,eval=F,cache=F}
+# Build 1x3 plot with sigma vs phi with color equal to distance to changepoint, power, or min time. The three panel plots would represent different shift size values
+
+source('R/sampling_regular.R')
+source('R/sampling_random.R')
+source('R/build_time_series.R')
+source('R/calc_min_time2.R')
+require(tidyverse)
+
+number_sims <- 100
+# Find minimum time for different parameter combos
+
+sigma_vec = seq(0.01,0.3,0.02)
+phi_vec = seq(-0.8,0.8,0.1)
+shift_size_vec = c(0.1,0.4,0.8)
+
+output_theory_approach <- data.frame(expand.grid(sigma_vec,phi_vec,shift_size_vec))
+names(output_theory_approach) <- c('sigma_vec', 'phi_vec', 'shift_size_vec')
+output_theory_approach$power = 0
+
+#params <- tibble(shift_time = sample(30:70,max(simulation_number),replace = T))
+
+shift_time = sample(30:70,number_sims,replace=T) # Keep shift times constant for each param combo
+
+# Run through params
+for (i in 1:nrow(output_theory_approach)){
+
+# Build XX number of time series
+params <- list(sigma=rep(output_theory_approach$sigma_vec[i],number_sims),phi=rep(output_theory_approach$phi_vec[i],number_sims),shift_size=rep(output_theory_approach$shift_size_vec[i],number_sims),shift_time = shift_time)
+
+time_series <- params %>%
+  pmap(build_time_series) %>%
+  map_dbl(function(df) sample_regular(df,20)$changepoint)
+
+power <- sum(abs(time_series-shift_time)<5)/number_sims
+
+#print(i)
+output_theory_approach$power[i] <- power
+}
+
+write.csv(x = output_theory_approach,file = 'Output/parameter_sensitivity.csv',quote = FALSE )
+
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
+  facet_wrap(~shift_size_vec) +
+   labs(fill='Statistical power') +
+   xlab("Lag-1 autocorrelation (phi)") +
+   ylab("Standard deviation (sigma)") +
+   theme_classic(base_size = 14, base_family = "")
+
+```
+
+```{r parameter_sensitivity_plot,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Regular sampling statistical power (fraction of 100 simulations which detected a changepoint within five time points of the true changepoint) for different levels of standard deviation ($\\sigma$), lag-1 autocorrelation ($\\phi$), and shift size ($\\delta$).
+For each parameter combination, 20 samples were used.
+An increase in samples would increase the statistical power across this graph.\\label{fig:parameter_sensitivity_plot}'}
+output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity.csv')
+
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
+  facet_wrap(~shift_size_vec) + 
+   theme_classic(base_size = 14, base_family = "")+
+  theme(plot.subtitle = element_text(hjust = 0.5)) +
+   labs(fill='Power',subtitle = 'Shift size') +
+   xlab("Lag-1 autocorrelation (phi)") +
+   ylab("Standard deviation (sigma)")
+```
 
@@
 
-```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size.
-Regular sampling was used with the default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$.
+```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size ($\\delta$).
+Regular sampling was used with the default parameters: $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study.
 The exact timing of the true changepoint varied for each simulation between time steps 30 and 70.
@@
 
-In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$).
-We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size.
+In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$) (Fig.
+\ref{fig:parameter_sensitivity_plot}).
+We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size (Fig.
+\ref{fig:parameter_sensitivity_plot}).
+There were interaction effects between the variables.
+For example, autocorrelation was only important if population variability was high (Fig.
+\ref{fig:parameter_sensitivity_plot}).
 Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
 \ref{fig:simulation_min_time_plot}).
-For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
+As predicted, iterative sampling performed best, followed by regular and random sampling (Fig.
+\ref{fig:simulation_min_time_plot}).
+The distance to the true changepoint, and consequently the minimum number of samples required, was lower for iterative sampling.
+(Fig.
+\ref{fig:simulation_min_time_plot})
 
@@
 
-We examined the performance of our approach on detecting changepoint from a paleosequence.
-Paleolimnology allows to reconstruct past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
+We examined the performance of our approach to detect changepoints in a paleosequence.
+Paleolimnology provides a reconstruction of past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
 A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).
 
-We tested different sampling methods on a real community time-series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
+We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
@@
 In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
-Changepoints were then detected on the first component (46% of the total variability) in 1926/1928, 1946/1948 and 1983/1988.
+Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983.
 We defined these as the "true" changepoints given they came from an analysis of the complete data.
-The 1983/1988 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
-We identified second and third changepoints in Supplementary Materials.
+The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
+We also identified second and third changepoints (Fig.
+S3).
 
@@
 \ref{fig:case_study_iterative_sampling}a).
-We reduced the dimensionality of the assemblage-level data to ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
+We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
 In the case of the iterative method, a new sample was added and the ordination was run again (Fig.
@@
 
-We found that random sampling performed the worst, as changepoint analysis was left to chance.
-Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
-Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
+We found that random sampling performed the worst, as changepoint analysis was left to chance (Fig.
+\ref{fig:case_study_number_sample}).
+Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
+Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
 \ref{fig:case_study_number_sample}c).
 We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig.
-\ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig.
-\ref{fig:case_study_number_sample} (c), square markers), the distance to the true changepoint was initially very high.
-Without any prior, the initial number sampled fell far away from the true changepoint.-->
+\ref{fig:case_study_iterative_sampling})
 
-We tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
-Iterative sampling still performed best, especially if an higher number of initial samples was chosen (7 samples out of 74, Supplementary Materials S5).
+We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
+Iterative sampling still performed best, especially if an higher number of initial samples was chosen (Fig.
+S3).
+
+As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
+S5).
+We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
+S6).
+The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
+\ref{fig:case_study_number_sample}c).
+<!--However, depending on the metric of interest, the subsampling will target different subsamples to add.
+In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->
+
 <!--This could probably be moved to the discussion: The initial number of samples should be chosen depending on the objective of the study.
@@
 If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.-->
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
+Total number of samples was set between 5 and 30, out of the 74 initial time series.
+\\label{fig:case_study_number_sample}'}
+# Read data
+
+require(stickylabeller)
+
+myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
+myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
+myoutput$diff_real <- abs(myoutput$diff_real)
+
+# Change to TRUE if we decide it is worth it to show several initial number of n2. It looks more confusing -- but it would bring more information
+several_n2_init = F
+
+# Plot
+p1 <- if(several_n2_init) {
+  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
+  theme(legend.position = c(0.93, 0.72),
+        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
+        legend.title=element_text(size=7), 
+        legend.text=element_text(size=8)) +
+  guides(pch=guide_legend(title="Initial number\nof sample"))
+} else {
+  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
+  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
+  geom_point(alpha = .3) +
+  facet_wrap(~method,
+             labeller = label_glue('{.l}. {method} sampling ')) + 
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        strip.background = element_rect(colour="transparent", fill="transparent"),
+        strip.text.x = element_text(angle = 0, hjust = 0)) +
+  labs(x="Total number of samples analyzed", y="Distance to true changepoint") 
+}
+
+p1
+
+```
 
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
-Total number of samples was set between 5 and 30, out of the 74 initial time series.
-\\label{fig:case_study_number_sample}'}
-# Read data
-
-require(stickylabeller)
-
-myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
-myoutput$method <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))
-myoutput$diff_real <- abs(myoutput$diff_real)
-
-# Change to TRUE if we decide it is worth it to show several initial number of n2. It looks more confusing -- but it would bring more information
-several_n2_init = F
-
-# Plot
-p1 <- if(several_n2_init) {
-  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method,
-             labeller = label_glue('{.l}. {method} sampling ')) + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        strip.background = element_rect(colour="transparent", fill="transparent"),
-        strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
-  theme(legend.position = c(0.93, 0.72),
-        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
-        legend.title=element_text(size=7), 
-        legend.text=element_text(size=8)) +
-  guides(pch=guide_legend(title="Initial number\nof sample"))
-} else {
-  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
-  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
-  geom_point(alpha = .3) +
-  facet_wrap(~method,
-             labeller = label_glue('{.l}. {method} sampling ')) + 
-  theme_bw() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        strip.background = element_rect(colour="transparent", fill="transparent"),
-        strip.text.x = element_text(angle = 0, hjust = 0)) +
-  labs(x="Total number of samples analyzed", y="Distance to true changepoint") 
-}
-
-p1
-
-```
-
-# General Discussion
-
-<!--
-Outline:
-
-- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2))
-- money implications (before or now)  
-- how this fits in with other work: examples of situation it would be applied (Rosalie)  
-- how this doesn't fit with other work (limitations) (Rosalie)
-- future work  (?)
--->
+# Discussion
 
@@
 Each sample took an average of 3 hours to process.
-We found that using an interative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
+We found that using an iterative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
 This correspond to several thousands of US dollars depending on labor costs.
@@
 
-As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
-S5).
-We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
-S6).
-The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
-\ref{fig:case_study_number_sample}c).
-However, depending on the metric of interest, the subsampling will target different subsamples to add.
-In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.
-
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d; @White2020].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020].
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
 Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
-It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017].
+It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
 However, both different types of data and different questions than the one we focused on can be addressed with the same approach.
@@
 
-<!---- how this doesn't fit with other work (limitations) ---->
-<!-- need line that references issue that iterative sampling can get stuck on incorrect points-->
-
 Our approach is applicable to a wide range of systems and questions, but it does have limitations.
-When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be prefered.
+When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred.
 We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs.
 \ref{fig:simulation_approaches_plot}, S2, S3).
-Another example where our method present less interest is when adressing questions that require a continuous time series, or at least a regular sampling interval.
+Another example where our method present less interest is when addressing questions that require a continuous time series, or at least a regular sampling interval.
 For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
 
-However recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@arkilanian_effect_nodate].
+However recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@Arkilanian2020].
 Extending on the question of being able to detect early warning signals, this objective must be weighed in with the ultimate goal: detecting ecosystem shift.
@@
 
-<!--not sure where to include this 
-Many ecological processes are non linear (e.g., D'Amario et al 2019, Scientific reports).
-Detecting a changepoint on one aspect of the ecosystem does not guarantee that the sampling resolution is appropriate to detect the driver for the changepoint.--->
-
 <!---- conclusion ---->
 
-Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d].
+# Conclusions
+
+Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019e].
 We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series.
@@
 Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b].
-Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
+Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
 If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling.
@@
 
-<!--
-Eventually, it is the researchers choice to choose their approach.
-Our objective was to quantify the implications of an approach vs.
-another.
-In the eventuality that the main objective is to detect a changepoint, we find that an approach is largely superior to the two others.
--->
-
-\clearpage
-
 # References

### 2020-04-07 R/supp_material.Rmd
change title
correct title
Changed order of figures, added figure, and edited text

@@
 \begin{center}
-\textbf{\Large Supplementary Material for: Sampling requirements and approaches to detect ecosystem shifts}
-\vspace{7 mm}
-
-\textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
+\textbf{\Large Supplementary material for: Sampling requirements and approaches to detect ecosystem shifts}
 \vspace{5 mm}
 
-\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
+\textsc{Rosalie Bruel$^{*}$\footnote{*The authors contributed equally to this work. $\dagger$Corresponding author (eastonrwhite@gmail.com)} and Easton R. White$^{* \dagger 2,3}$}
+\vspace{3 mm}
+
+\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT, USA \\ $^{2}$Department of Biology, University of Vermont, VT, USA \\
+$^{3}$Gund Institute for Environment, University of Vermont, VT, USA}
 \end{center}
@@
 
+<!--
 # Parameter sensitivity
@@
 \ref{fig:parameter_sensitivity_plot}).
-
-```{r parameter_sensitivity,message=F,warning=F,echo=F,eval=F,cache=F}
-# Build 1x3 plot with sigma vs phi with color equal to distance to changepoint, power, or min time. The three panel plots would represent different shift size values
-
-source('R/sampling_regular.R')
-source('R/sampling_random.R')
-source('R/build_time_series.R')
-source('R/calc_min_time2.R')
-require(tidyverse)
-
-number_sims <- 10
-# Find minimum time for different parameter combos
-
-sigma_vec = seq(0.01,0.3,0.02)
-phi_vec = seq(-0.8,0.8,0.1)
-shift_size_vec = c(0.1,0.4,0.8)
-
-output_theory_approach <- data.frame(expand.grid(sigma_vec,phi_vec,shift_size_vec))
-names(output_theory_approach) <- c('sigma_vec', 'phi_vec', 'shift_size_vec')
-output_theory_approach$power = 0
-
-#params <- tibble(shift_time = sample(30:70,max(simulation_number),replace = T))
-
-shift_time = sample(30:70,number_sims,replace=T) # Keep shift times constant for each param combo
-
-# Run through params
-for (i in 1:nrow(output_theory_approach)){
-
-# Build XX number of time series
-params <- list(sigma=rep(output_theory_approach$sigma_vec[i],number_sims),phi=rep(output_theory_approach$phi_vec[i],number_sims),shift_size=rep(output_theory_approach$shift_size_vec[i],number_sims),shift_time = shift_time)
-
-time_series <- params %>%
-  pmap(build_time_series) %>%
-  map_dbl(function(df) sample_regular(df,20)$changepoint)
-
-power <- sum(abs(time_series-shift_time)<5)/number_sims
-
-#print(i)
-output_theory_approach$power[i] <- power
-}
-
-write.csv(x = output_theory_approach,file = 'Output/parameter_sensitivity.csv',quote = FALSE )
-
-require(ggplot2)
-require(viridis)
-ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
-   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
-  facet_wrap(~shift_size_vec) +
-   labs(fill='Statistical power') +
-   xlab("Lag-1 autocorrelation (phi)") +
-   ylab("Standard deviation (sigma)") +
-   theme_classic(base_size = 14, base_family = "")
-
-```
-
-```{r parameter_sensitivity_plot,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Regular sampling statistical power (fraction of 100 simulations which detecting a changepoint within five time points of the true changepoint) for different levels of standard deviation, lag-1 autocorrelation, and shift size.\\label{fig:parameter_sensitivity_plot}'}
-output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity.csv')
-
-require(ggplot2)
-require(viridis)
-ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
-   scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
-  facet_wrap(~shift_size_vec) + 
-   theme_classic(base_size = 14, base_family = "")+
-  theme(plot.subtitle = element_text(hjust = 0.5)) +
-   labs(fill='Power',subtitle = 'Shift size') +
-   xlab("Lag-1 autocorrelation (phi)") +
-   ylab("Standard deviation (sigma)")
-```
+-->
 
@@
 
-for (param_num in 11:nrow(sigma_output)){
+for (param_num in 1:nrow(sigma_output)){
   sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num],'iterative')

### 2020-04-07 R/manuscript_draft.Rmd
Updates to main manuscript and supp mat before publishing preprint

@@
 Herein, we focus on detecting the most important changepoint (i.e.
-the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4).
+the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig.
+S3).
 In order to test the performance, we detected the "true" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "true" one.
@@
 The same analyses for random and iterative sampling is in Figs.
-S2,S3.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}
+S1,S2.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}
 
@@
 As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
+S4).
+We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
 S5).
-We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
-S6).
 The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
@@
 We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
-\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S2, S3).
+\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S1, S2).
 Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig.
@@
 The method also worked well to detect other changepoints of lower magnitude (Fig.
-S4).
+S3).
 @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
@@
 The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
-It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
-However, both different types of data and different questions than the one we focused on can be addressed with the same approach.
+Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
+It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
+However, both different types of data and different questions than those used here can be addressed with the same approach.
 Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
@@
 We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs.
-\ref{fig:simulation_approaches_plot}, S2, S3).
-Another example where our method present less interest is when addressing questions that require a continuous time series, or at least a regular sampling interval.
+\ref{fig:simulation_approaches_plot}, S1, S2).
+Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval.
 For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
 
-However recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@Arkilanian2020].
-Extending on the question of being able to detect early warning signals, this objective must be weighed in with the ultimate goal: detecting ecosystem shift.
+However, recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows forecasating population collapses even with at lower resolution and time-series length [@Arkilanian2020].
 Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implication in addressing the driver.
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implications in addressing the underlying driver.
 Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-Finally, more subsamples could be processed later to address additional questions.
 

### 2020-04-07 R/supp_material.Rmd
Updates to main manuscript and supp mat before publishing preprint

@@
 require(tidyverse)
+require(stickylabeller)
 
@@
 Each vertical line indicates the respective parameter calculated from the case study time series.
-\\label{fig:simulation_min_time_random}',message=F,warning=F,echo=F,eval=T}
+\\label{fig:simulation_min_time_random}',message=F,warning=F,echo=F,eval=F}
 require(ggplot2)
@@
 
+```{r parameter_sensitivity_plot_random,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Random sampling statistical power (fraction of 100 simulations which detected a changepoint within five time points of the true changepoint) for different levels of standard deviation ($\\sigma$), lag-1 autocorrelation ($\\phi$), and shift size ($\\delta$).
+For each parameter combination, 20 samples were used.
+An increase in samples would increase the statistical power across this graph.\\label{fig:parameter_sensitivity_plot_random}'}
+output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity_random.csv')
+
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9, limits=c(0,1)) +
+  facet_wrap(~shift_size_vec) + 
+  theme_classic(base_size = 14, base_family = "")+
+  theme(plot.subtitle = element_text(hjust = 0.5)) +
+  labs(fill='Power',subtitle = 'Shift size') +
+  xlab("Lag-1 autocorrelation (phi)") +
+  ylab("Standard deviation (sigma)")
+```
+
 ## Iterative sampling
@@
 Each vertical line indicates the respective parameter calculated from the case study time series.
-\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}
+\\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=F}
 
@@
 
+```
+
+```{r parameter_sensitivity_plot_iterative,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Iterative sampling statistical power (fraction of 100 simulations which detected a changepoint within five time points of the true changepoint) for different levels of standard deviation ($\\sigma$), lag-1 autocorrelation ($\\phi$), and shift size ($\\delta$).
+For each parameter combination, 20 samples were used.
+An increase in samples would increase the statistical power across this graph.\\label{fig:parameter_sensitivity_plot_iterative}'}
+output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity_iterative.csv')
+
+require(ggplot2)
+require(viridis)
+ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
+  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
+  facet_wrap(~shift_size_vec) + 
+  theme_classic(base_size = 14, base_family = "")+
+  theme(plot.subtitle = element_text(hjust = 0.5)) +
+  labs(fill='Power',subtitle = 'Shift size') +
+  xlab("Lag-1 autocorrelation (phi)") +
+  ylab("Standard deviation (sigma)")
 ```
@@
 
+require(stickylabeller)
+
 myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese_biomass.txt"))

### 2020-06-18 R/manuscript_draft.Rmd
edit to the manuscript post review. Starting a paragraph on what has already been done, and what is our specific contribution.

@@
 As long as processing samples is expensive, these trade-offs will remain.
+
+Two dimensions are to be considered when sampling: spatial and temporal dimension.
+To estimate a same parameter, population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@kermorvant_optimizing_2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@jackson_sampling_2008].
+_Add at least another sentence here to make the transition to our work, but right now I can't think of a good formulation_
 

### 2020-06-22 R/manuscript_draft.Rmd
changes to cover page

--- 2020-04-07 R/manuscript_draft.Rmd
@@
 
-Number of words: XXXX \vspace{5 mm}
+Number of words: 3400 \vspace{5 mm}
 
-Number of references: <!--\total{citenum}-->
+Number of references: 35<!--\total{citenum}-->
 
-Number of figures: 5
+Number of figures: 6
 
@@
 
-Submission as: Article at \emph{Journal of Applied Ecology}
+Submission as: Article at \emph{Ecological Applications}
 
@@
 
-Preprint available at:
+Preprint available at: https://www.biorxiv.org/content/10.1101/2020.04.07.030643v1
 
@@
 \textbf{\Large Sampling requirements and approaches to detect ecosystem shifts}
-\vspace{5 mm}
+\vspace{2 mm}
 
 \textsc{Rosalie Bruel$^{*1}$\footnote{*The authors contributed equally to this work. $\dagger$Corresponding author (eastonrwhite@gmail.com)} and Easton R. White$^{* \dagger 2,3}$}
-\vspace{3 mm}
+\vspace{1 mm}
 
@@
 
-<!--\linenumbers-->
-\singlespacing
+\linenumbers
+\doublespacing
 
@@
 
-\vspace{3 mm}
+\vspace{0.5 mm}
 
-\textbf{Keywords:} time series, changepoint, monitoring, optimal sampling, Cladocera
+\textbf{Keywords:} time series, changepoint, monitoring, optimal sampling, Cladocera, paleoecology
 
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual diagram illustrating the process of taking (a) simulations of a time series and (b) selecting a single simulation to analyze with three different sampling approaches: (c) random, (d) regular, and (e) iterative.
-The iterative sampling approach requires (f) adding samples around a detected changepoint until (g) a certain level of accuracy is achieved.
-\\label{fig:conceptual}'}
-source(paste0(getwd(),"/R/Fig_conceptual.R"))
-```
-
 The random sampling approach involves taking a set number of random points throughout the time series (Fig.
@@
 The minimum number of samples required is the number needed for 0.8 statistical power.
+
+# Simulation results
+
+In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$) (Fig.
+\ref{fig:parameter_sensitivity_plot}).
+We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size (Fig.
+\ref{fig:parameter_sensitivity_plot}).
+There were interaction effects between the variables.
+For example, autocorrelation was only important if population variability was high (Fig.
+\ref{fig:parameter_sensitivity_plot}).
+Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
+\ref{fig:simulation_min_time_plot}).
+As predicted, iterative sampling performed best, followed by regular and random sampling (Fig.
+\ref{fig:simulation_min_time_plot}).
+The distance to the true changepoint, and consequently the minimum number of samples required, was lower for iterative sampling.
+(Fig.
+\ref{fig:simulation_min_time_plot})
+
+# Case study
+
+## Case study background
+
+We examined the performance of our approach to detect changepoints in a paleosequence.
+Paleolimnology provides a reconstruction of past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
+A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).
+
+We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
+Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
+It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed.
+Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018].
+Air temperature is now driving changes in plankton communities [@Bruel2018].
+
+In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
+Our objective was to evaluate whether the same changepoints could be identified using less samples.
+In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
+Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983.
+We defined these as the "true" changepoints given they came from an analysis of the complete data.
+The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
+We also identified second and third changepoints (Fig.
+S3).
+
+In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative).
+These subsamples were from the initial community dataset (Fig.
+\ref{fig:case_study_iterative_sampling}a).
+We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
+In the case of the iterative method, a new sample was added and the ordination was run again (Fig.
+\ref{fig:case_study_iterative_sampling}d).
+For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.
+
+## Case study results
+
+We found that random sampling performed the worst, as changepoint analysis was left to chance (Fig.
+\ref{fig:case_study_number_sample}).
+Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
+Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
+\ref{fig:case_study_number_sample}c).
+We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig.
+\ref{fig:case_study_iterative_sampling})
+
+We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
+Iterative sampling still performed best, especially if an higher number of initial samples was chosen (Fig.
+S3).
+
+As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
+S4).
+We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
+S5).
+The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
+\ref{fig:case_study_number_sample}c).
+<!--However, depending on the metric of interest, the subsampling will target different subsamples to add.
+In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->
+
+<!--This could probably be moved to the discussion: The initial number of samples should be chosen depending on the objective of the study.
+If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoretical results?).
+If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.-->
+
+# Discussion
+
+<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
+Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed.
+When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing.
+A decision must then be made as to which subsamples to analyze.
+To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling random points, regular intervals, and an iterative sampling approach (Fig.
+\ref{fig:conceptual}).
+We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
+\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S1, S2).
+Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig.
+\ref{fig:simulation_min_time_plot}).
+
+<!---- money implications ---->
+
+Multiple subsampling strategies can be chosen (Fig.
+\ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig.
+\ref{fig:simulation_approaches_plot}c).
+Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
+Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig.
+\ref{fig:case_study_number_sample}).
+The method also worked well to detect other changepoints of lower magnitude (Fig.
+S3).
+@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
+Each sample took an average of 3 hours to process.
+We found that using an iterative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
+This correspond to several thousands of US dollars depending on labor costs.
+
+<!---- how this fits in with other work: examples of situation it would be applied (potential) ---->
+
+Our approach goes beyond just paleoecological analyses.
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020].
+The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
+Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
+It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
+However, both different types of data and different questions than those used here can be addressed with the same approach.
+Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
+It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
+Using an interval sampling approach, one could take a fixed number of samples to start.
+The trend over time from simple linear regression could be taken.
+Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates.
+If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample.
+Sampling would continue until the trend estimate reached an equilibrium.
+Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
+These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
+
+Our approach is applicable to a wide range of systems and questions, but it does have limitations.
+When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred.
+We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs.
+\ref{fig:simulation_approaches_plot}, S1, S2).
+Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval.
+For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
+
+However, recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows forecasting population collapses even with at lower resolution and time-series length [@Arkilanian2020].
+Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
+Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implications in addressing the underlying driver.
+Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
+
+<!---- conclusion ---->
+
+# Conclusions
+
+Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019e].
+We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series.
+Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
+In addition, optimizing sampling efforts is valuable given the high costs of many monitoring programs [@Caughlan2001a;@Bennett2014a].
+Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b].
+Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
+If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling.
+This can improve the accuracy of the results and reduce costs overall.
+
+# Code availability
+
+Data and code for all the figures can be found at https://github.com/rosalieb/temporal-sampling.
+
+# References
+
+<div id="refs"></div>
+
+<!--# Figures-->
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual diagram illustrating the process of taking (a) simulations of a time series and (b) selecting a single simulation to analyze with three different sampling approaches: (c) random, (d) regular, and (e) iterative.
+The iterative sampling approach requires (f) adding samples around a detected changepoint until (g) a certain level of accuracy is achieved.
+\\label{fig:conceptual}'}
+source(paste0(getwd(),"/R/Fig_conceptual.R"))
+```
+
+\clearpage
 
@@
 
-# Simulation results
-
-In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$) (Fig.
-\ref{fig:parameter_sensitivity_plot}).
-We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size (Fig.
-\ref{fig:parameter_sensitivity_plot}).
-There were interaction effects between the variables.
-For example, autocorrelation was only important if population variability was high (Fig.
-\ref{fig:parameter_sensitivity_plot}).
-Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig.
-\ref{fig:simulation_min_time_plot}).
-As predicted, iterative sampling performed best, followed by regular and random sampling (Fig.
-\ref{fig:simulation_min_time_plot}).
-The distance to the true changepoint, and consequently the minimum number of samples required, was lower for iterative sampling.
-(Fig.
-\ref{fig:simulation_min_time_plot})
-
 ```{r simulation_results_sampling_approaches,message=F,warning=F,echo=F,eval=F}
@@
 
+\clearpage
+
 ```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling.
@@
 
-# Case study
-
-## Case study background
-
-We examined the performance of our approach to detect changepoints in a paleosequence.
-Paleolimnology provides a reconstruction of past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
-A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).
-
-We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
-Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
-It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed.
-Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018].
-Air temperature is now driving changes in plankton communities [@Bruel2018].
-
-In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
-Our objective was to evaluate whether the same changepoints could be identified using less samples.
-In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
-Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983.
-We defined these as the "true" changepoints given they came from an analysis of the complete data.
-The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
-We also identified second and third changepoints (Fig.
-S3).
-
-In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative).
-These subsamples were from the initial community dataset (Fig.
-\ref{fig:case_study_iterative_sampling}a).
-We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
-In the case of the iterative method, a new sample was added and the ordination was run again (Fig.
-\ref{fig:case_study_iterative_sampling}d).
-For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.
-
-## Case study results
-
-We found that random sampling performed the worst, as changepoint analysis was left to chance (Fig.
-\ref{fig:case_study_number_sample}).
-Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
-Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
-\ref{fig:case_study_number_sample}c).
-We show how iterative sampling slightly changes the scores on the first component but not the overall ordination, as more samples are added (Fig.
-\ref{fig:case_study_iterative_sampling})
-
-We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
-Iterative sampling still performed best, especially if an higher number of initial samples was chosen (Fig.
-S3).
-
-As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
-S4).
-We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig.
-S5).
-The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig.
-\ref{fig:case_study_number_sample}c).
-<!--However, depending on the metric of interest, the subsampling will target different subsamples to add.
-In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->
-
-<!--This could probably be moved to the discussion: The initial number of samples should be chosen depending on the objective of the study.
-If investigators want to find the biggest transition only, then a low number might be enough (include here results from theoretical results?).
-If investigators suspect or want to test for several changepoints, we recommend analyzing initially about 10% of the length of the time series to have a good initial ranking.-->
+\clearpage
 
@@
 ```
+
+\clearpage
 
@@
 ```
-
-# Discussion
-
-<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
-Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed.
-When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing.
-A decision must then be made as to which subsamples to analyze.
-To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling random points, regular intervals, and an iterative sampling approach (Fig.
-\ref{fig:conceptual}).
-We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs.
-\ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S1, S2).
-Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig.
-\ref{fig:simulation_min_time_plot}).
-
-<!---- money implications ---->
-
-Multiple subsampling strategies can be chosen (Fig.
-\ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig.
-\ref{fig:simulation_approaches_plot}c).
-Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.
-Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig.
-\ref{fig:case_study_number_sample}).
-The method also worked well to detect other changepoints of lower magnitude (Fig.
-S3).
-@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core.
-Each sample took an average of 3 hours to process.
-We found that using an iterative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work.
-This correspond to several thousands of US dollars depending on labor costs.
-
-<!---- how this fits in with other work: examples of situation it would be applied (potential) ---->
-
-Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020].
-The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
-It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
-However, both different types of data and different questions than those used here can be addressed with the same approach.
-Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
-It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
-Using an interval sampling approach, one could take a fixed number of samples to start.
-The trend over time from simple linear regression could be taken.
-Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates.
-If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample.
-Sampling would continue until the trend estimate reached an equilibrium.
-Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
-These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
-
-Our approach is applicable to a wide range of systems and questions, but it does have limitations.
-When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred.
-We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs.
-\ref{fig:simulation_approaches_plot}, S1, S2).
-Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval.
-For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
-
-However, recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows forecasating population collapses even with at lower resolution and time-series length [@Arkilanian2020].
-Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017].
-Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implications in addressing the underlying driver.
-Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.
-
-<!---- conclusion ---->
-
-# Conclusions
-
-Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019e].
-We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series.
-Monitoring programs have to be able to address our questions of interest with sufficient statistical power.
-In addition, optimizing sampling efforts is valuable given the high costs of many monitoring programs [@Caughlan2001a;@Bennett2014a].
-Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b].
-Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
-If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling.
-This can improve the accuracy of the results and reduce costs overall.
-
-# References

### 2020-06-23 R/manuscript_draft.Rmd
updated intro paragraph

@@
 
-pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot)
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp)
 ```
@@
 
+Because ecological systems are dynamic in both space and time, it is essential that sampling designs account for spatio-temporal dynamics (Williams et al.
+2018).
+To estimate a same parameter, population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@kermorvant_optimizing_2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@jackson_sampling_2008].
+In either case, sampling time and lcoations can be chosen in an iterative process to be cost-effective and reduce the uncertainty in the process (Hooten et al.
+2009).
+With the ability to choose precisely when to sample, we can move beyond random, interval, or opportunistic sampling designs (citations).
+This is particuraly relevant in situations where a subset of samples already collected need to be analyzed.
+
 Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred.
@@
 
-Two dimensions are to be considered when sampling: spatial and temporal dimension.
-To estimate a same parameter, population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@kermorvant_optimizing_2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@jackson_sampling_2008].
-_Add at least another sentence here to make the transition to our work, but right now I can't think of a good formulation_
-
-Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed.
+Building on ideas from dynamic survey design (citations), we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed.
 We tailor our analysis to the detection of a changepoint in a time series, but our approach is applicable to other questions as well.

### 2020-06-29 R/manuscript_draft.Rmd
made small updates to text

--- 2020-06-22 R/manuscript_draft.Rmd
@@
 
-Number of references: 35<!--\total{citenum}-->
+Number of references: 38<!--\total{citenum}-->
 
@@
 The specific number of years required depended on the species biology and the detection method used [@White2019e].
-Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population.
-Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the benefit of increasing sampling breadth relying on citizen science [@Weiser2020a].
+Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].
 Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].
+
+Because ecological systems are dynamic in both space and time, it is essential that sampling designs account for spatio-temporal dynamics [@Williams2018].
+To estimate a particular parameter, e.g.
+population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@Kermorvant2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@Jackson2008a].
+In either case, sampling time and lcoations can be chosen in an iterative process to be cost-effective and reduce the uncertainty in the process [@Hooten2009].
+With the ability to choose precisely when to sample, we can move beyond random, interval, or opportunistic sampling designs.
+This is particuraly relevant in situations where a subset of samples already collected need to be analyzed.
 
@@
 
-Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed.
+Building on ideas from dynamic survey design and optimal monitoring research, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed.
 We tailor our analysis to the detection of a changepoint in a time series, but our approach is applicable to other questions as well.
@@
 Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.
-If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling.
+If only a subsample of the samples can be analyzed, it may be better to choose samples strategically as opposed to random or regular sampling.
 This can improve the accuracy of the results and reduce costs overall.
 
-# Code availability
+# Author contributions
+
+ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication
+
+# Data availability
 

### 2020-06-29 R/manuscript_draft.Rmd
small spelling changes

@@
 population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@Kermorvant2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@Jackson2008a].
-In either case, sampling time and lcoations can be chosen in an iterative process to be cost-effective and reduce the uncertainty in the process [@Hooten2009].
+In either case, sampling time and locations can be chosen in an iterative process to be cost-effective and reduce the uncertainty in the process [@Hooten2009].
 With the ability to choose precisely when to sample, we can move beyond random, interval, or opportunistic sampling designs.
-This is particuraly relevant in situations where a subset of samples already collected need to be analyzed.
+This is particularly relevant in situations where a subset of samples already collected need to be analyzed.
 

### 2020-06-29 R/manuscript_draft.Rmd
modified citations

@@
 
-Number of references: 38<!--\total{citenum}-->
+Number of references: 40<!--\total{citenum}-->
 
@@
 
-Submission as: Article at \emph{Journal of Applied Ecology}
+Submission as: Article at \emph{Ecological Indicators}
 
@@
 We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change.
-We test these ideas with a case study of Cladocera community assemblage reconstructed from a sediment core.
+We test these ideas with a case study of Cladocera community assemblage indicators reconstructed from a sediment core.
 We demonstrate that standard approaches to sample processing are less efficient than an iterative approach.
@@
 Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed.
-Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].
+Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lengyel2018; @Lindenmayer2020a].
 
@@
 The specific number of years required depended on the species biology and the detection method used [@White2019e].
-Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a].
+Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a].
 Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a].

### 2020-06-30 R/manuscript_draft.Rmd
Manuscript edits

@@
 # Use [@R-dplyr; @R-rmarkdown] in main text
+
+# Load packages
+if (!require("pacman",character.only = TRUE))
+  {
+    install.packages("pacman",dep=TRUE)
+    if(!require("pacman",character.only = TRUE)) stop("Package not found")
+  }
+
+# Keeping below source for github package. Ask Easton whether pacman works for github packages or not.
+#devtools::install_github("rensa/stickylabeller")
+
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp)
 ```
@@
 
-```{r load_packages,echo=F,warning=F,message=F}
-if (!require("pacman",character.only = TRUE))
-  {
-    install.packages("pacman",dep=TRUE)
-    if(!require("pacman",character.only = TRUE)) stop("Package not found")
-  }
-
-# Keeping below source for github package. Ask Easton whether pacman works for github packages or not.
-#devtools::install_github("rensa/stickylabeller")
-
-pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp)
-```
-
 # Introduction
@@
 
-Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change.
-The specifics of the monitoring program will determine the power with which a question of interest can be addressed.
+Monitoring programs characteristics must be tightly linked to the questions of interest.
 For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
@@
 \ref{fig:conceptual}).
-We then subsampled each time series [@White2020] to test the effect of three different sampling approaches along with varying the sample size (Fig.
+We then subsampled each time series to test the effect of three different sampling approaches along with varying the sample size (Fig.
 \ref{fig:conceptual}c-f).
@@
 We examined the performance of our approach to detect changepoints in a paleosequence.
-Paleolimnology provides a reconstruction of past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
+Paleolimnology allows to reconstruct past environments over long periods of time, under the premise that sedimentation was not perturbed (low mixing and disturbances).
 A sediment core is typically subsampled to narrow down periods of time to be compared, either at regular intervals (e.g., @Milan2015), or continuously (e.g., @Perga2015).
 
-We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in Cladocera assemblage.
+We tested different sampling methods on a real community time series from Lake Varese (IT), with the objective to detect the main changepoint in zooplankton Cladocera assemblage.
 Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl).
@@
 
-In a previous study, Cladoceran assemblage was reconstructed at a 2-3 years resolution, from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
+In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
 Our objective was to evaluate whether the same changepoints could be identified using less samples.
@@
 \ref{fig:case_study_number_sample}).
-Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance).
+Regular sampling provided good estimates from 8 samples, but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., also left to chance).
 Iterative sampling performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig.
@@
 We also tested how the three methods performed at detecting other changepoints of lower magnitude (as three changepoints were detected in the initial study, @Bruel2018).
-Iterative sampling still performed best, especially if an higher number of initial samples was chosen (Fig.
+Iterative sampling still performed best, especially if an higher number of initial samples (7) was chosen (Fig.
 S3).
@@
 \ref{fig:case_study_number_sample}c).
+The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset.
 <!--However, depending on the metric of interest, the subsampling will target different subsamples to add.
@@
 
-ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication
+ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication
 

### 2020-09-23 R/build_time_series.R
edit build_time_series to add the option of a linear shift

@@ -1,20 +1,29 @@
 
+# Build ar-1 time series with various model inputs, and a linear transition
 
-# Build ar-1 time series with various model inputs
-
-build_time_series <- function(mu=0,sigma=0.053,phi=0.404,shift_size=0.81,shift_time=50,max_time=100){
+build_time_series <- function(mu=0,sigma=0.053,phi=0.404,shift_size=0.81,shift_time=50,max_time=100, shift_duration = 1){
   
   ar1 <- vector(mode = 'numeric',max_time)
   meanXt = 0.899
   ar1[1] <- meanXt
+  if(shift_duration !=1) {
+    shift_size = shift_size / shift_duration
+    shift_time <- rep(shift_time, shift_duration)
+    for (i in 2:shift_duration) {
+      shift_time[i] <- shift_time[i-1]+1
+    }
+  }
+  
   #shift_time = sample(5:(max_time-5),size = 1)
   for (t in 2:max_time){
     #if(t == shift_time){meanXt = (1-shift_size)*ar1[1]}
     #ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)
-   if(t == shift_time){meanXt = meanXt - shift_size;ar1[t]=meanXt}
+    if(t %in% shift_time){meanXt = meanXt - shift_size;ar1[t]=meanXt}
     else{ ar1[t] <- meanXt + phi*(ar1[t-1] - meanXt) + rnorm(1,mean=mu,sd = sigma)}
   }
   
   ar1 = (ar1-min(ar1))/(max(ar1)-min(ar1))
   return(ar1) 
 }
+
+# End of the function

### 2020-09-23 R/work_for_answer_to_reviewers.R

@@ -1,0 +1,105 @@
+x = 1:100
+
+y = -0.000001*x + 1 + rnorm(length(x),0,0.001)
+
+plot(x,y)
+
+vrall <- sample_regular(y, n=length(x), input_vector = T, xcol = 1, DCA_axis = 1, messages = F)
+
+vrall$changepoint
+
+
+
+vrsub <- sample_iterative(y, n=10, n2=5, input_vector = T, xcol = 1, DCA_axis = 1, messages = F)
+
+vrsub$changepoint
+
+
+#pdf(paste0(getwd(),"/Output/Worst_case_scenario_1.pdf"), fonts = "Helvetica", height = 5, width = 8 )
+plot(x,y, ylab = "y = -0.000001*x + 1 + rnorm(length(x),0,0.001)")
+abline(v=vrsub$changepoint)
+points(vrsub$matrix$x ~ vrsub$matrix$index, col="tomato")
+points(vrsub$matrix$x ~ vrsub$matrix$index, pch=16, col="tomato")
+#dev.off()
+
+summary(lm(vrsub$matrix$x ~ vrsub$matrix$index))
+
+vrsub$changepoint_all$order.found
+
+vrsub$changepoint_all$p.values
+
+
+# Time serie
+y <- build_time_series_complex(shift_duration = 10)
+vrall <- sample_regular(y, n=length(x), input_vector = T, xcol = 1, DCA_axis = 1, messages = F)
+
+vrall$changepoint
+
+
+
+vrsub <- sample_iterative(y, n=10, n2=5, input_vector = T, xcol = 1, DCA_axis = 1, messages = F)
+
+vrsub$changepoint
+
+
+#pdf(paste0(getwd(),"/Output/linear_shift.pdf"), fonts = "Helvetica", height = 5, width = 8 )
+plot(x,y)
+abline(v=vrsub$changepoint)
+points(vrsub$matrix$x ~ vrsub$matrix$index, col="tomato")
+points(vrsub$matrix$x ~ vrsub$matrix$index, pch=16, col="tomato")
+#dev.off()
+
+summary(lm(vrsub$matrix$x ~ vrsub$matrix$index))
+
+
+
+source('R/sampling_regular.R')
+source('R/sampling_random.R')
+source('R/build_time_series.R')
+source('R/calc_min_time2.R')
+require(tidyverse)
+
+number_sims <- 100
[…]

### 2020-10-02 R/sampling_iterative.R
add the possibility in the function to edit R (permutation number) and sig.lvl, for the function e.divisive, wrapped into each of the sampling function

@@ -11,13 +11,16 @@
 #' @param input_vector Default input vector = T. If turned to F, allow a data frame
 #' @param xcol if input_vector = F, index of columns not to use for ordination.
 #' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
+#' @param R number of permutation in the e.divisive function from the package ecp
+#' @param sig.lvl significance level in the e.divisive function from the package ecp
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
-  .sample_iterative(x, n, n2, c, messages, input_vector, xcol, DCA_axis)
-.sample_iterative <- function(x, n, n2, c, messages, input_vector, xcol, DCA_axis) {
+sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, xcol = NA, DCA_axis = 1,
+                             R = 199, sig.lvl = 0.05)
+  .sample_iterative(x, n, n2, c, messages, input_vector, xcol, DCA_axis, R, sig.lvl)
+.sample_iterative <- function(x, n, n2, c, messages, input_vector, xcol, DCA_axis, R, sig.lvl) {
   require(ecp)
   require(zoo)
   require(vegan)
@@ -51,7 +54,7 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
   if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
-  temp_chgpt <- e.divisive(x2, min.size = 2, sig.lvl = .7)$order.found[3:(2+c)]
+  temp_chgpt <- e.divisive(x2, min.size = 2, R = R, sig.lvl = .7)$order.found[3:(2+c)]
   
   if(all(is.na(temp_chgpt)) && length(which_samples) < n ) {
     which_samples <- seq(1,2,lastpoint,length.out = n2*2-1)
@@ -73,7 +76,7 @@ sample_iterative <- function(x, n, n2 = 5, c=1, messages = T, input_vector = T, 
     if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
     x2 <- matrix(x2[,2])
     
[…]

### 2020-10-02 R/sampling_random.R

@@ -9,13 +9,16 @@
 #' @param input_vector Default input vector = T. If turned to F, allow a data frame
 #' @param xcol if input_vector = F, index of columns not to use for ordination.
 #' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
+#' @param R number of permutation in the e.divisive function from the package ecp
+#' @param sig.lvl significance level in the e.divisive function from the package ecp
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
-  .sample_random(x, n, messages, input_vector, xcol, DCA_axis)
-.sample_random <- function(x, n, messages, input_vector, xcol, DCA_axis) {
+sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1,
+                          R = 199, sig.lvl = 0.05)
+  .sample_random(x, n, messages, input_vector, xcol, DCA_axis, R, sig.lvl)
+.sample_random <- function(x, n, messages, input_vector, xcol, DCA_axis, R, sig.lvl) {
   require(ecp)
   require(zoo)
   require(vegan)
@@ -46,11 +49,19 @@ sample_random <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_a
   if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
-  changepoints <- e.divisive(x2, min.size = 2)
+  changepoints <- e.divisive(x2, min.size = 2, R = R, sig.lvl = sig.lvl)
   if(length(changepoints$p.values)>1) changepoints$p.values <- changepoints$p.values[-length(changepoints$p.values)]
+  if(length(changepoints$p.values)>1) changepoints$estimates <- changepoints$estimates[-c(1, length(changepoints$estimates))]
   
   # Print the main changepoint
   if(messages) print(changepoints$order.found[3])
+  
+  mmessage <- paste0("We used the function e.divisive() in the package ecp to calculate the changepoint(s). ",
+                     "The original function let the user choose R (the number of permutation, default = 199) and sig.lev (the significance level, default = 0.05). ",
+                     "Herein, the two parameters are: R = ", R, ", and sig.lvl = ", sig.lvl,". ",
+                     "p-values in the output are calculated during the permutation process by the e.divisive function. ",
[…]

### 2020-10-02 R/sampling_regular.R

@@ -9,13 +9,16 @@
 #' @param input_vector Default input vector = T. If turned to F, allow a data frame
 #' @param xcol if input_vector = F, index of columns not to use for ordination.
 #' @param DCA_axis which DCA axis to use as vector on which to detect changepoint
+#' @param R number of permutation in the e.divisive function from the package ecp
+#' @param sig.lvl significance level in the e.divisive function from the package ecp
 #' @keywords sampling
 #' @examples 
 #'
 
-sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1)
-  .sample_regular(x, n, messages, input_vector, xcol, DCA_axis)
-.sample_regular <- function(x, n, messages, input_vector, xcol, DCA_axis) {
+sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_axis = 1,
+                           R = 199, sig.lvl = 0.05)
+  .sample_regular(x, n, messages, input_vector, xcol, DCA_axis, R, sig.lvl)
+.sample_regular <- function(x, n, messages, input_vector, xcol, DCA_axis, R, sig.lvl) {
   require(ecp)
   require(zoo)
   require(vegan)
@@ -48,11 +51,19 @@ sample_regular <- function(x, n, messages = T, input_vector = T, xcol = NA, DCA_
   if(any(is.na(x2[,2]))) x2[,2] <- na.locf(x2[,2])
   x2 <- matrix(x2[,2])
   
-  changepoints <- e.divisive(x2, min.size = 2)
+  changepoints <- e.divisive(x2, min.size = 2, R = R, sig.lvl = sig.lvl)
   if(length(changepoints$p.values)>1) changepoints$p.values <- changepoints$p.values[-length(changepoints$p.values)]
+  if(length(changepoints$p.values)>1) changepoints$estimates <- changepoints$estimates[-c(1, length(changepoints$estimates))]
   
   # Print the main changepoint
   if(messages) print(changepoints$order.found[3])
+  
+  mmessage <- paste0("We used the function e.divisive() in the package ecp to calculate the changepoint(s). ",
+                     "The original function let the user choose R (the number of permutation, default = 199) and sig.lev (the significance level, default = 0.05). ",
+                     "Herein, the two parameters are: R = ", R, ", and sig.lvl = ", sig.lvl,". ",
+                     "p-values in the output are calculated during the permutation process by the e.divisive function. ",
[…]

### 2020-10-05 R/supp_material.Rmd
edits to address reviewer comments

@@
 require(stickylabeller)
+require(ggplot2)
+require(viridis)
 
@@
 
-require(ggplot2)
-require(viridis)
 ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
@@
 
+# Case study: Testing other types of models
+
+Throughout the main text and supplementary material, we focus on detecting changepoints.
+However, other types of models and criteria may be more appropriate depending on the question.
+This is especially true for situations where changepoints are not expected apriori.
+To address this, we used our random and regular sampling approaches with two additional models: linear models and generalized additive models.
+For simple linear regression, we measured the percent difference in the estimated slope between the subsampled data and the full time series (Fig.
+\ref{fig:lineartrends}).
+This is similar to past work [@White2019e; @White2020] on detecting population trends.
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling.
+Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples.
+\\label{fig:lineartrends}'}
+
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+vr_count <- data.frame(Year=vr[,1], Biomass=rowSums(vr[,-1]))
+cpvar_count <- sample_regular(vr_count[,2], nrow(vr_count), input_vector = T, messages = F)
+
+#plot(vr_count)
+#abline(lm(vr_count$Biomass~vr_count$Year))
+true_slope <- as.numeric(coef(lm(vr_count$Biomass~vr_count$Year))[2])
+
+regular_linear <- function(x,y,num_samples){
+  true_slope <- as.numeric(coef(lm(y~x))[2])
+
+  subset_index <- c(seq(from = 1, to = length(x),length.out =num_samples),length(x))
+  subset_x <- x[subset_index]
+  subset_y <- y[subset_index]
+
+  subset_slope <- as.numeric(coef(lm(subset_y~subset_x))[2])
+
+ return(abs(true_slope - subset_slope)/true_slope)  
+}
+
+spacing_vec <- seq(3,70,by=3)
+distance <- vector(mode='numeric',length = length(spacing_vec))
+for (i in 1:length(spacing_vec)){
+  distance[i] <- regular_linear(x = vr_count$Year,y=vr_count$Biomass,num_samples =spacing_vec[i])
+
+}
+par(mfrow=c(1,2),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))
+plot(spacing_vec,distance,las=1,ylim=c(0,1.2),col='darkgrey',pch=16)
+abline(h=0,col='darkgrey',lwd=1.5)
+mtext('a. Random sampling - linear model',side = 3,line = 0,adj = 0)
+
+# Random sampling
+
+random_linear <- function(x,y,num_samples){
+  true_slope <- as.numeric(coef(lm(y~x))[2])
+
+  subset_index <- sort(c(1,sample(2:(length(x)-1),size = num_samples),length(x)))
+  subset_x <- x[subset_index]
+  subset_y <- y[subset_index]
+
+  subset_slope <- as.numeric(coef(lm(subset_y~subset_x))[2])
+
+  return(abs(true_slope - subset_slope)/true_slope)  
+}
+
+#spacing_vec <- 3:60
+distance <- vector(mode='numeric',length = length(spacing_vec))
+for (i in 1:length(spacing_vec)){
+  distance[i] <- random_linear(x = vr_count$Year,y=vr_count$Biomass,num_samples =spacing_vec[i])
+
+}
+plot(spacing_vec,distance,las=1,ylim=c(0,1.2),yaxt='n',col='darkgrey',pch=16)
+abline(h=0,col='darkgrey',lwd=1.5)
+mtext('b. Regular sampling - linear model',side = 3,line = 0,adj = 0)
+mtext('Percent difference',side = 2,line = 2.5,outer = T,font = 2)
+mtext('Total number of samples analyzed',side = 1,line = 2.5,outer = T,font = 2)
+
+```
+
+Similarly, for a generalized additive model, we examined the shape of the estimated curve for subsets of data compared to the full time series (Fig.
+\ref{fig:gamtrends}).
+In line with our other results, the regular sampling performed better than random sampling upon visual inspection.
+Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models.
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling.
+Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples.
+\\label{fig:gamtrends}'}
+
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+vr_count <- data.frame(Year=vr[,1], Biomass=rowSums(vr[,-1]))
+cpvar_count <- sample_regular(vr_count[,2], nrow(vr_count), input_vector = T, messages = F)
+
+smoothing_parameter=5
+require(mgcv)
+
+par(mfrow=c(2,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))
+
+subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 10),length(vr_count$Year)))
+vr_count_subset <- vr_count[subset_index,]
+gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam1_rand,ylim=c(-50000,50000),xaxt='n')
+mtext('a. Random (n=10)',side = 3,line = 0,adj = 0)
+
+subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 20),length(vr_count$Year)))
+vr_count_subset <- vr_count[subset_index,]
+gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+mtext('b. Random (n=20)',side = 3,line = 0,adj = 0)
+
+subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 30),length(vr_count$Year)))
+vr_count_subset <- vr_count[subset_index,]
+gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+mtext('c. Random (n=30)',side = 3,line = 0,adj = 0)
+
+gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
+plot(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+mtext('d. Random (n=74)',side = 3,line = 0,adj = 0)
+
+# Regular GAMs
+
+subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =10),length(vr_count$Year))
+vr_count_subset <- vr_count[subset_index,]
+gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam1_rand,ylim=c(-50000,50000))
+mtext('e. Regular (n=10)',side = 3,line = 0,adj = 0)
+
+subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =20),length(vr_count$Year))
+vr_count_subset <- vr_count[subset_index,]
+gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam2_rand,ylim=c(-50000,50000),yaxt='n')
+mtext('f. Regular (n=20)',side = 3,line = 0,adj = 0)
+
+subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =30),length(vr_count$Year))
+vr_count_subset <- vr_count[subset_index,]
+gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot(gam3_rand,ylim=c(-50000,50000),yaxt='n')
+mtext('g. Regular (n=30)',side = 3,line = 0,adj = 0)
+
+gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
+plot(gam4_rand,ylim=c(-50000,50000),yaxt='n')
+mtext('h. Regular (n=40)',side = 3,line = 0,adj = 0)
+
+mtext('Standardized abundance',side = 2,line = 2.5,outer = T,font = 2)
+mtext('Time (year)',side = 1,line = 2.5,outer = T,font = 2)
+
+```
+
 \clearpage

### 2020-10-05 R/manuscript_draft.Rmd
trying out a new col palette

@@
 For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting.
-Standard sub-sampling approaches often involve equally-spaced sampling.
-We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change.
+Standard sub-sampling approaches often involve equally-spaced sampling \textcolor{red}{on depth}.\linelabel{text:depth} We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change.
 We test these ideas with a case study of Cladocera community assemblage indicators reconstructed from a sediment core.
@@
 Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a].
-However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012].
+\linelabel{text:remote1} However, environmental monitoring \textcolor{red}{can still be logistically difficult, expensive, and time-consuming, especially} when further processing is needed following sample collection [@Zhang2012].
 Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed.
@@
 
-Monitoring programs characteristics must be tightly linked to the questions of interest.
-For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
+\linelabel{text:program} \textcolor{red}{Monitoring program characteristics must be tightly linked to the questions of interest.} For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time.
 The specific number of years required depended on the species biology and the detection method used [@White2019e].
+\linelabel{text:newcitation} \textcolor{red}{In addition, the sampling effort required differs depending on the question.
+For example, questions regarding phenology would require many sampling points within a season and across years} [@Filippa2015a, @White2020e].
 Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a].
@@
 Should the paleoecological core be analyzed at every centimeter?
-Should the video be assessed once per minute?
+Should the video be assessed once per minute if automated tools are not available?
 As long as processing samples is expensive, these trade-offs will remain.
@@
 \ref{fig:conceptual}d).
-Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
+Lastly, iterative sampling involves first taking a set number of samples (i.e., regular sampling) and then iteratively adding samples until a pre-determined level of precision is achieved (Fig.
 \ref{fig:conceptual}e-g).
-For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period.
-We describe each approach in more detail in the supplementary material and provide code.
+\linelabel{text:iterative2} \textcolor{red}{In the context of a changepoint, this means adding a new sample near the best estimate for the changepoint and iteratively updating the estimate (Fig.
+\ref{fig:conceptual}e-g).} For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period.
+\textcolor{red}{We describe each approach in more detail in the supplementary material and provide code.} \linelabel{text:iterative1}
 
@@
 \ref{fig:parameter_sensitivity_plot}).
+\linelabel{text:detection1} \textcolor{red}{In line with theory, if the shift size was small, and thus there was no changepoint, there was almost never enough power to detect a shift.
+Similar results have been shown for the detection of linear trends} [@White2019e].
 There were interaction effects between the variables.
@@
 In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018].
-Our objective was to evaluate whether the same changepoints could be identified using less samples.
-In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
+Our objective was to evaluate whether the same changepoints could be identified using fewer samples.\linelabel{text:fewer} In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980].
 Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983.
@@
 \ref{fig:case_study_iterative_sampling}a).
-We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector).
-In the case of the iterative method, a new sample was added and the ordination was run again (Fig.
+We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method in the original study, and detected the changepoint on the first component (univariate vector).
+\linelabel{text:iterative} \textcolor{red}{In the case of the iterative method, a new sample was added in the temporal region closest to the estimate for the changepoint until a more precise estimate was obtained} (Fig.
 \ref{fig:case_study_iterative_sampling}d).
@@
 
-As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
+\linelabel{text:detection2} \textcolor{red}{To show the generality of our approach}, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig.
 S4).
@@
 The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset.
+\textcolor{red}{In addition, we examined the same abundance time series, but with linear and generalized additive models (GAMs)}.
+In line with findings from} @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs.
+S6, S7).}
+
 <!--However, depending on the metric of interest, the subsampling will target different subsamples to add.
@@
 Our approach goes beyond just paleoecological analyses.
-Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020].
-The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
-Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
+Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020] or \textcolor{red}{logistically difficult, such as with remote field sites}.\linelabel{text:remote} The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost.
+\linelabel{text:newcitation1} \textcolor{red}{Similarly, in order to evaluate ecosystem phenology,} @Filippa2015a \textcolor{red}{used similar techniques to show the effect of not only different levels of sampling, but also the effect of using different indices altogether.} Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012].
 It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017].
 However, both different types of data and different questions than those used here can be addressed with the same approach.
-Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches.
+Suppose instead that the goal was to detect a change in relative abundance over time with video-based \textcolor{red}{approaches where automated techniques are not possible}.
 It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest.
@@
 Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics.
-These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].
+\linelabel{text:otherapproaches} \textcolor{red}{These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples} [@Bohmann2014a].
+\textcolor{red}{The same approaches could also be used in studying evidence of stressers found in tree rings or coral reef cores} [@Carilli2012a; @Pretzsch2013a], \textcolor{red}{pollen in aerial traps or sediment cores to address phenological questions} [@Haselhorst2013a], \textcolor{red}{and many others situations described in} @White2020.
+\linelabel{text:otherapproaches1}
 
@@
 \ref{fig:simulation_approaches_plot}, S1, S2).
-Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval.
+\linelabel{red}{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected.
+Instead, more flexible models (e.g., GAMs) might be more appropriate (Fig.
+S7).} Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval.
 For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].
@@
 mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
+#mypal <-  c("#4C4F54", "#D57A5D", "#AE5654", "#D18437", "#9EC6DD")
 

### 2020-10-07 R/manuscript_draft.Rmd
work for answers to reviewer

@@
 
-pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp)
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp, ggthemes)
 ```
@@
 Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983.
-We defined these as the "true" changepoints given they came from an analysis of the complete data.
-The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
+We defined these as the "true" changepoints given they came from an analysis of the complete sequence The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method.
 We also identified second and third changepoints (Fig.
@@
 \textcolor{red}{In addition, we examined the same abundance time series, but with linear and generalized additive models (GAMs)}.
-In line with findings from} @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs.
+In line with findings from @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs.
 S6, S7).}
@@
 \ref{fig:simulation_approaches_plot}, S1, S2).
-\linelabel{red}{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected.
+\linelabel{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected.
 Instead, more flexible models (e.g., GAMs) might be more appropriate (Fig.
@@
 
-ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication
+RB and ERW conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches, and applied the method to the case study data; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication.
 
@@
 
+mypal = colorblind_pal()(8)
+mypal = mypal[c(7,2,3,6,8)]
 mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
-#mypal <-  c("#4C4F54", "#D57A5D", "#AE5654", "#D18437", "#9EC6DD")
+mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal[4] <- "#90b581"
+mypal <- mypal[c(1,3,4,7,9)]
 

### 2020-10-07 R/supp_material.Rmd
work for answers to reviewer

@@
 However, other types of models and criteria may be more appropriate depending on the question.
-This is especially true for situations where changepoints are not expected apriori.
+This is especially true for situations where changepoints are not expected _a priori_.
 To address this, we used our random and regular sampling approaches with two additional models: linear models and generalized additive models.
@@
 
-```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling.
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a-d) random and (e-h) regular sampling.
 Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples.
@@
 Similarly, for a generalized additive model, we examined the shape of the estimated curve for subsets of data compared to the full time series (Fig.
-\ref{fig:gamtrends}).
+\ref{fig:gamtrends} \textcolor{red}{the caption is not the correct one, I edited it, but check you agree with it!}).
 In line with our other results, the regular sampling performed better than random sampling upon visual inspection.
-Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models.
+Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models \textcolor{red}{(I do not understand this sentence)}.
+Iterative sampling requires only 8 samples to find the changepoint, but the a larger initial number is also necessary to capture the whole dynamic.
 
-```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling.
-Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples.
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 4.4,fig.cap='The fitted smooth functions for Year for different subset of the zooplankton abundance dataset for Lake Varese (IT).
+Subsamples using (a-d) random and (e-h) regular sampling.
+(i-j) shows the influence of the initial number of sample  (n2) when using the iterative method.
+For low initial numbers, the dynamic of the abundance is not caught.
+Choosing a larger initial number (n2) allows to capture the dynamic and find the real changepoint with a limited total number of samples (n).
 \\label{fig:gamtrends}'}
@@
 
-par(mfrow=c(2,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))
+par(mfrow=c(3,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))
+
+# Random GAMs
 
@@
 gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam1_rand,ylim=c(-50000,50000),xaxt='n')
+plot.gam(gam1_rand,ylim=c(-50000,50000),xaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('a. Random (n=10)',side = 3,line = 0,adj = 0)
@@
 gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+plot.gam(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('b. Random (n=20)',side = 3,line = 0,adj = 0)
@@
 vr_count_subset <- vr_count[subset_index,]
-gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian)
+plot.gam(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('c. Random (n=30)',side = 3,line = 0,adj = 0)
@@
 gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
-plot(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
+plot.gam(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count$Year, NA, tcl=0.3, lwd=.5)
 mtext('d. Random (n=74)',side = 3,line = 0,adj = 0)
@@
 gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam1_rand,ylim=c(-50000,50000))
+plot.gam(gam1_rand,ylim=c(-50000,50000),xaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('e. Regular (n=10)',side = 3,line = 0,adj = 0)
@@
 gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam2_rand,ylim=c(-50000,50000),yaxt='n')
+plot.gam(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('f. Regular (n=20)',side = 3,line = 0,adj = 0)
@@
 gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
-plot(gam3_rand,ylim=c(-50000,50000),yaxt='n')
+plot.gam(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
 mtext('g. Regular (n=30)',side = 3,line = 0,adj = 0)
@@
 gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
-plot(gam4_rand,ylim=c(-50000,50000),yaxt='n')
-mtext('h. Regular (n=40)',side = 3,line = 0,adj = 0)
+plot.gam(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
+axis(1, at=vr_count$Year, NA, tcl=0.3, lwd=.5)
+mtext('h. Regular (n=74)',side = 3,line = 0,adj = 0)
+
+# Iterative GAMs
+
+subset_index <- sample_iterative(vr, xcol = 1, n = 10, input_vector = F, DCA_axis = 1)
+subset_index <- subset_index$matrix$index
+vr_count_subset <- vr_count[subset_index,]
+gam1_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot.gam(gam1_it,ylim=c(-50000,50000), shade = T)
+mtext('i. Iterative (n=8, n2=5)',side = 3,line = 0,adj = 0)
+
+subset_index <- sample_iterative(vr, xcol = 1, n = 20, n2 = 10, input_vector = F, DCA_axis = 1)
+subset_index <- subset_index$matrix$index
+vr_count_subset <- vr_count[subset_index,]
+gam2_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot.gam(gam2_it,ylim=c(-50000,50000),yaxt='n', shade = T)
+mtext('j. Iterative (n=13, n2=10)',side = 3,line = 0,adj = 0)
+
+subset_index <- sample_iterative(vr, xcol = 1, n = 50, n2 = 30, input_vector = F, DCA_axis = 1)
+subset_index <- subset_index$matrix$index
+vr_count_subset <- vr_count[subset_index,]
+gam3_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
+plot.gam(gam3_it,ylim=c(-50000,50000),yaxt='n', shade = T)
+mtext('h. Iterative (n=32, n2=30)',side = 3,line = 0,adj = 0)
+
+gam4_it <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
+plot.gam(gam4_it,ylim=c(-50000,50000), shade = T)
+mtext('k. Iterative (n2=74)',side = 3,line = 0,adj = 0)
 
@@
 
+# Word of caution
+
+The limits of our three sub-sampling functions are tied to the limits of the _e.divise()_ function.
+In a situation where the change is a linear change instead of a sudden changepoint, the function _e.divise()_ will find a changepoint in the middle of the linear decline (Fig.
+\ref{fig:caution1}).
+In such situation, iterative sampling still performs the best.
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 6, fig.height= 8,fig.cap='Examples of two time-series where shift lasted over (a) 5 and (b) 10 observations.
+Building off the example time-series in (b), we tested the three sampling method in (c-d).
+(c-d) are same as in Fig.
+1 in the main manuscript.
+\\label{fig:caution1}'}
+
+mypal <- wes_palette("Chevalier1")
+mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5, type = "continuous")
+mypal[4] <- "#90b581"
+mypal <- mypal[c(1,3,4,7,9)]
+
+ar1 <- build_time_series()
+mydata <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+
+cex_titles <- .8
+
+# Require an output from any of the 3 functions sampling_regular, sampling_random, sampling_iterative
+pol_break <- function(cpt, col=1, y1=-.2, y2=1.2) {
+  if(col==1) mycol=mypal[1]
+  if(col==2) mycol=mypal[2]
+  x1   <- cpt$changepoint
+  x2   <- cpt$matrix$index[which(cpt$matrix$index==cpt$changepoint)-1]
+  pol_x <- c(x1, x2, x2, x1)
+  pol_y <- c(y1, y1, y2, y2)
+  polygon(pol_x, pol_y, col= adjustcolor(mycol, alpha.f = .3), density = NULL, border = NA)
+}
+
+layout(matrix(c(1,2,0,
+         3,4,5,
+         0,0,6,
+         0,0,7), ncol=3, byrow = T))
+par(mar=c(1,2,6,2))
+
+# plot 1: shift_duration = 5 ####
+ar1 <- build_time_series(shift_duration = 5, shift_time = 62)
+ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+mydata <- ar1
+
+plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp <- sample_regular(mydata, n=length(mydata), messages = F)
+pol_break(cpt = cp)
+lines(mydata, col="black", lwd=1.5)
+text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
+mtext("a. Example time-series with a shift \nlasting over 5 observations", side = 3, line = 2.5, adj = 0, font=1, cex=cex_titles)
+
+# plot2: shift_duration = 10  ####
+ar1 <- build_time_series(shift_duration = 10, shift_time = 35)
+ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+mydata <- ar1
+
+plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp <- sample_regular(mydata, n=length(mydata), messages = F)
+pol_break(cpt = cp)
+lines(mydata, col="black", lwd=1.5)
+text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
+mtext("a. Example time-series with a shift \nlasting over 10 observations", side = 3, line = 2.5, adj = 0, font=1, cex=cex_titles)
+
+# Plot 3 -- random sampling ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp_random <- sample_random(mydata, n=15, messages = F)
+points(cp_random$matrix$index, cp_random$matrix$x, pch=4, lwd=2)
+pol_break(cpt = cp, col=1)
+pol_break(cpt = cp_random, col=2)
+lines(mydata, col="black", lwd=1.5)
+text(100,0.95,labels = paste0("n = ",cp_random$final_n), pos = 2)
+mtext("c. Random sampling", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+
+# Plot 4 -- regular sampling ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp_regular <- sample_regular(mydata, n=15, messages = F)
+points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
+pol_break(cpt = cp, col=1)
+pol_break(cpt = cp_regular, col=2)
+lines(mydata, col="black", lwd=1.5)
+text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
+mtext("d. Regular sampling", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+
+# Plot 5 -- iterative sampling (1/3) ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+cp_iterative <- sample_regular(mydata, n=5, messages = F)
+pol_break(cpt = cp, col=1)
+pol_break(cpt = cp_iterative, col=2)
+lines(mydata, col="black", lwd=1.5)
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
+text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
+mtext("e. Iterative sampling (initial)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+xleft_temp <- cp_iterative$matrix$index[which(cp_iterative$matrix$index==cp_iterative$changepoint)-1]-4 
+xright_temp <- cp_iterative$changepoint+4 
+par(xpd=T)
+rect(xleft = xleft_temp, xright = xright_temp, ybottom = -0.1, ytop = 1.1, lty = 2)
+text(xright_temp, 1, labels = "f", pos = 4)
+par(xpd=F)
+
+# Plot 6 -- iterative sampling (2/3) ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="", xlim=c(xleft_temp,xright_temp))
+cp_iterative1 <- sample_regular(mydata, n=5, messages = F)
+cp_iterative2 <- sample_iterative(mydata, n=6, messages = F)
+cp_iterative3 <- sample_iterative(mydata, n=7, messages = F)
+cp_iterative4 <- sample_iterative(mydata, n=8, messages = F)
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+mypal2 <- wes_palette("GrandBudapest2")
+mypal2 <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal2[4] <- "#90b581"
+mypal2 <- mypal[c(1,3,4,7,9)]
+
+points(cp_iterative1$matrix$index, cp_iterative$matrix$x, col=grey(0.5), pch=4, lwd=2)
+points(cp_iterative2$matrix$index[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], cp_iterative2$matrix$x[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], 
+       col=mypal2[1], pch=4, lwd=2)
+points(cp_iterative3$matrix$index[!cp_iterative3$matrix$index%in%cp_iterative2$matrix$index], cp_iterative3$matrix$x[!cp_iterative3$matrix$index%in%cp_iterative2$matrix$index], 
+       col=mypal2[2], pch=4, lwd=2)
+points(cp_iterative4$matrix$index[!cp_iterative4$matrix$index%in%cp_iterative3$matrix$index], cp_iterative4$matrix$x[!cp_iterative4$matrix$index%in%cp_iterative3$matrix$index], 
+       col=mypal2[3], pch=4, lwd=2)
+legend("topright", legend = c("t=0", "t=1", "t=2", "t=3"),
+       col = c(grey(0.5), mypal2), pch=4, lwd=2, bty="n", lty=NA)
+mtext("f. Zoom from e\n(slowly adding samples)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+
+# Plot 7 -- iterative sampling (3/3) ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+cp_iterative <- sample_iterative(mydata, n=15, messages = F)
+pol_break(cpt = cp, col=1)
+pol_break(cpt = cp_iterative, col=2)
+lines(mydata, col="black", lwd=1.5)
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
+text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
+mtext("g. Iterative sampling (final)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
+
+```
+
+Another issue emerge in time series with no changepoint.
+The first case is when the time-series follows a declining trend.
+In some situations, depending on the noise and the trend, a changepoint will still be detected (Fig.
+\ref{fig:caution2}).
+Again, this would be the case even with a continuous sampling, but the iterative method can falsely reinfore this (by converging to a changepoint, and hiding the general trend).
+In this scenario, increasing the initial number of samples analyzed (~30% of the initial samples) would allow to roughly model the behavior of the dataset (Fig.
+\ref{fig:gamtrends}).
+
+Overall, while it is important for potential users to be aware of the behavior of the code, community changes would more often lead to larger restructurations (alternative stable states, @Bruel2018).
+We caution in general against using too few samples for a dataset or a type of data we know too little about.
+
+```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 6,fig.cap='Examples of two time-series where shift lasted over (a) 5 and (b) 10 observations.
+Building , off the example time-series in (b), we tested the three sampling method in (c-d).
+(c-d) are same as in Fig.
+1 in the main manuscript.
+\\label{fig:caution2}'}
+
+par(mfrow=c(2,2))
+# Plot 1: no shift, declining trend ####
+ar1 <- -0.0001*x + 1 + rnorm(length(x),0,0.001)
+ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+mydata <- ar1
+
+plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp <- sample_regular(mydata, n=length(mydata), messages = F)
+if(!is.na(cp$changepoint)) pol_break(cpt = cp)
+lines(mydata, col="black", lwd=1.5)
+if(!is.na(cp$changepoint)) {
+  text(cp$changepoint, .8, labels = "change point \ndetected",pos = 4, col=mypal[1])
+} else {
+    text(3,.95, labels = "No changepoint", adj = 0, col=mypal[1])
+  }
+mtext("a. Example time-series with no shift, only a declining trend. \nNo changepoint should have been detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)
+
+# Plot 2: no shift, trend, subsample ####
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp_regular <- sample_regular(mydata, n=15, messages = F)
+points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
+#pol_break(cpt = cp, col=1)
+pol_break(cpt = cp_regular, col=2)
+text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
+mtext("b. Subsampling a time-series with no shift and a trend. \nNo changepoint should have beend detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)
+
+# Plot 2: no shift, no trend ####
+ar1 <- -0.000001*x + 1 + rnorm(length(x),0,0.001)
+ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
+mydata <- ar1
+
+plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp <- sample_regular(mydata, n=length(mydata), messages = F)
+if(!is.na(cp$changepoint)) pol_break(cpt = cp)
+lines(mydata, col="black", lwd=1.5)
+if(!is.na(cp$changepoint)) {
+  text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
+} else {
+    text(3,.95, labels = "No changepoint", adj = 0, col=mypal[1])
+  }
+mtext("c. Example time-series with no shift and no trend.\nNo changepoint should have beend detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)
+
+plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
+axis(1, lwd=2, at = c(-10,120))
+axis(2, lwd=2, at = c(-1,2))
+cp_regular <- sample_regular(mydata, n=15, messages = F)
+points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
+pol_break(cpt = cp_regular, col=2)
+text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
+mtext("d. Subsampling a time-series with no shift and no trend. \nNo changepoint detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)
+
+```
+
 \clearpage

### 2020-10-07 R/Fig_conceptual.R

@@ -1,4 +1,4 @@
-R_U_KNITTING=T
+R_U_KNITTING=F
 
 source(paste0(getwd(),"/R/build_time_series.R"))
 source(paste0(getwd(),"/R/sampling_regular.R"))
@@ -7,6 +7,9 @@ source(paste0(getwd(),"/R/sampling_iterative.R"))
 
 library(wesanderson)
 mypal <- wes_palette("Chevalier1")
+mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal[4] <- "#90b581"
+mypal <- mypal[c(1,3,4,7,9)]
 
 ar1 <- build_time_series()
 mydata <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
@@ -106,6 +109,10 @@ cp_iterative4 <- sample_iterative(mydata, n=8, messages = F)
 axis(1, lwd=2, at = c(-10,120))
 axis(2, lwd=2, at = c(-1,2))
 mypal2 <- wes_palette("GrandBudapest2")
+mypal2 <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal2[4] <- "#90b581"
+mypal2 <- mypal[c(1,3,4,7,9)]
+
 points(cp_iterative1$matrix$index, cp_iterative$matrix$x, col=grey(0.5), pch=4, lwd=2)
 points(cp_iterative2$matrix$index[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], cp_iterative2$matrix$x[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], 
        col=mypal2[1], pch=4, lwd=2)
@@ -130,3 +137,4 @@ text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
 mtext("g. Iterative sampling (final)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
 
 if(!R_U_KNITTING) dev.off()
+

### 2020-10-07 R/Fig_conceptual.R
update review

@@ -1,4 +1,4 @@
-R_U_KNITTING=F
+R_U_KNITTING=T
 
 source(paste0(getwd(),"/R/build_time_series.R"))
 source(paste0(getwd(),"/R/sampling_regular.R"))
@@ -7,7 +7,7 @@ source(paste0(getwd(),"/R/sampling_iterative.R"))
 
 library(wesanderson)
 mypal <- wes_palette("Chevalier1")
-mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal <- wes_palette("Zissou1", 10,type = "continuous")
 mypal[4] <- "#90b581"
 mypal <- mypal[c(1,3,4,7,9)]
 
@@ -109,7 +109,7 @@ cp_iterative4 <- sample_iterative(mydata, n=8, messages = F)
 axis(1, lwd=2, at = c(-10,120))
 axis(2, lwd=2, at = c(-1,2))
 mypal2 <- wes_palette("GrandBudapest2")
-mypal2 <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
+mypal2 <- wes_palette("Zissou1", 10,type = "continuous")
 mypal2[4] <- "#90b581"
 mypal2 <- mypal[c(1,3,4,7,9)]
 

### 2020-10-07 R/supp_material.Rmd
addition to supp material of p-value explanation

@@
 
-The limits of our three sub-sampling functions are tied to the limits of the _e.divise()_ function.
+The limits of our three sub-sampling functions are tied to the limits of the _e.divise()_ function [@James2014].
 In a situation where the change is a linear change instead of a sudden changepoint, the function _e.divise()_ will find a changepoint in the middle of the linear decline (Fig.
@@
 
+Finally, note that the p-values in the output of each functions are calculated within the function _e.divisive()_ (package _ecp_, @James2014).
+By default, the p-value of 0.005 means that no other permutation within the matrix lead to more energy needed to break the matrix.
+The p-value is in fact a frequency: only one configuration in 200 scenarios (initial matrix and 199 permutations) led to that much energy (1/200 = 0.005).
+We included in the three functions we coded (*sampling_random(), sampling_regular(), sampling_iterative()*) the arguments _R_ (default = 199) and _sig.lvl_ (default = 0.005), two arguments within the initial _e.divisive()_ function.
+By changing the default value, the user can choose a different number of permutation (if R = 19, then the minimum p-value will be 1/(1+19) = 0.05), and another significance level (if R = 199, the default, and the user wants to stop looking for changepoints once 10 configurations result in higher energy, then sig.lvl must be changed to 10/(1+199) = 0.05).
+Refer to the initial function for better understanding [@James2014].
+
 \clearpage

### 2020-10-16 R/manuscript_draft.Rmd
gif maker script

@@
              annotate("text", label = paste0("Number of samples = ",i), x = min(vrsub$matrix$age), y = 2.8, colour = "black",  hjust=0,vjust=0), 
-           filename = paste0("Output/GIF_adding_sample_case_study/Iterative_",i,".pdf"), width = 6, height = 4)
+           filename = paste0("Output/GIF_adding_sample_case_study/Iterative_",i,".png"), width = 5, height = 3.5)
 

### 2020-10-16 R/Gif_maker.R

@@ -1,0 +1,134 @@
+# Script for gif
+pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp, ggthemes)
+
+# Read data
+vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
+vr_out_matrix <- read.table(paste0(getwd(),"/Output/output_iterative_sampling_varese.txt"))
+
+index_temp <- vr_out_matrix$index[vr_out_matrix$n==max(vr_out_matrix$n)]
+vr_out_order <-
+  data.frame("index" = index_temp,
+             "age"   = vrall$matrix$age[index_temp],
+             "x"     = vrall$matrix$x[index_temp],
+             "count" = summary(as.factor(vr_out_matrix$index)))
+vr_out_order$count <- abs(vr_out_order$count-max(vr_out_order$count)-1)
+
+# Detect 'true' changepoint
+vrall <- sample_regular(vr, n=nrow(vr), input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
+vrall$matrix$age <- vr$AGE[vrall$matrix$index]
+
+# Plot 'true' changepoint
+preal <- ggplot(vrall$matrix, aes(age, x)) + 
+  geom_rect(stat = "identity", 
+            mapping=aes(
+              xmin=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)-1], 
+              xmax=vrall$matrix$age[which(vrall$matrix$index==vrall$changepoint)], 
+              ymin=-1, ymax=3), col=NA, fill="grey", alpha=0.01) +
+  annotate("text", label = paste0("Number of samples = 74"), x = min(vrall$matrix$age), y = 2.8, colour = "black",  hjust=0,vjust=0) +  
+  geom_line() + geom_point() + 
+  guides(col=FALSE) +
+  scale_color_manual(values=mypal) +
+  theme_bw() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  labs(x="Year", y="DCA first component",
+       color='samples added\nat iteration t=') +
+  labs(subtitle = "A. Full time-series and 'real' changepoint")
+
+
+
+# Random sampling
+vrrdm <- sample_random(vr, n=10, input_vector = F, xcol = 1, DCA_axis = 1, messages = F)
[…]
